### 2014-03-14 notes/graph-adapt/main.tex
added start of write-up

\documentclass{sig-alternate}

\begin{document}

\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}

\title{Adaptive Disk Storage for Interaction Graphs}

\numberofauthors{1} %  in this sample file, there are a *total*

\author{ }

\date{}

\maketitle
\begin{abstract}
TODO
\end{abstract}

\section{Introduction}
%
Typical interaction graphs have data associated with their vertices and edges.
Most algorithms access this data as they traverse the graph. However, not all
of the data is accessed by all the queries. Typically, there are correlations
among the attributes accessed by different queries, such as Q1 and Q5
accessing attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and
so on.
%

In an interaction graph database organized as blocks of temporal neighborlists
(as in my earlier paper), it is important to store the edge and vertex
properties locally. For  instance, if the edge properties are kept away in a
relational table, there will be almost no benefit to the locality
optimizations performed for block organization (discussed in my earlier
paper), as we would have to go back and forth between the disk blocks to
access the edge attributes. 
%

On the other hand, putting all the edge attributes into the disk blocks
containing the graph structure is expected to cause significant overhead when
only a few of these attributes are read. This is somewhat similar to the 
problem with row-oriented databases, where the entire row needs to be accessed
from the disk despite the query needing only a small fraction of it.
Unfortunately, there is no clear correspondence to a column-oriented database
layout for the graph databases.
%

Recall that interaction graphs are temporal. As such, the co-access
correlations for the attributes can be different for different temporal
regions. It might also  be unknown at the insertion time and may be discovered
later depending on the workload. This points to the need for adaptively
optimizing the layout (somewhat similar to the H2O paper).
%

I imagine a solution to this problem, which I name the 'rail layout'. The idea
is to start with large blocks that contain the entire data. As we learn about
the access properties for different time regions, we might start splitting
such blocks into smaller blocks that run parallel to each other, almost like
having two or more graph databases for certain time regions, each containing a
 different subset of the attributes, but with a link between them in case a
query needs to access both. Some attributes can be replicated as well. 
More details to follow about all this...

\section{Example}

\section{Adaptation}

\section{Cost Model}

\bibliographystyle{abbrv}
\bibliography{main}

\end{document}

### 2014-03-24 notes/graph-adapt/main.tex
added some notes

@@
 \documentclass{sig-alternate}
+
+\usepackage{alltt}
 
@@
 \section{Introduction}
+An \emph{interaction graph} is an append-only graph, where new edges and
+vertices are added as time progresses. 
+%
+
 %
@@
 relational table, there will be almost no benefit to the locality
-optimizations performed for block organization (discussed in my earlier
-paper), as we would have to go back and forth between the disk blocks to
+optimizations performed for block organization (discussed in \cite{gedik14}), as we would have to go back and forth between the disk blocks to
 access the edge attributes. 
@@
 
+\begin{alltt}\scriptsize
+- Work through a concrete example of a database and a 
+   couple of queries that would change the layout on disk.
+
+- An interaction graph is an append-only graph, where new 
+   edges and vertices are added as time progresses.
+
+- Typically, in graph dbs, if a vertex is visited during a 
+   traversal, then it is likely that its neighbors will also 
+    be visited
+
+- So, we store we place highly connected neighbors in the 
+    same disk block, to minimize disk I/O.
+
+- Interaction graphs are temporal. The edges between vertices 
+   change over time.
+
+- For interaction graphs, we need to group edges spatially 
+   and temporally.
+
+- Vertices and edges have data associated with them.
+
+- We want to store as much relevant data as possible in the 
+   same block.
+
+- We will store:
+   - edges and vertices that are connected in the same time interval
+   - edges and vertices that are highly connected (neighbors)
+   - edge attributes that are likely to be accessed
+   - vertex attributes that are likely to be accessed
+
+- Example: We use an interaction graph to model connections
+   between people on Twitter. Every time
+
+- $C^d(B) = conductance, lower is better.$
+- $C^h(B) = cohesiveness, higher is better.$
+
+\end{alltt}
+
+$L(B) = \sqrt[2]{C^h(B) \cdot ( 1 - C^d(B))}$
+
 \section{Adaptation}

### 2014-03-25 notes/graph-adapt/main.tex
added rdk notes

@@
 
+\emph{In the problem you're describing, one natural idea that springs to mind is
+  to reduce it back to a graph problem by the edges of the original interaction
+  graph as vertices of an "augmented interaction graph" whose vertex set
+  consists of both the vertices and edges of the original graph, i.e. all
+  entities that might have data associated to them. And two vertices of the
+  augmented interaction graph are joined by an edge if either (a) both of them
+  are vertices of the original interaction graph, and they are joined by an
+  edge, or (b) one is an edge and the other is a vertex which is an endpoint of
+  that edge. Pictorially, the augmented interaction graph is obtained by
+  inserting a new vertex alongside each edge of the original interaction graph
+  and turning the edge into a triangle with the new vertex as its third
+  corner. It seems that this would work well assuming that queries which
+  traverse an edge are likely to traverse its endpoints. -- rdk}
+
 \section{Adaptation}

### 2014-04-09 notes/graph-adapt/main.tex
initial formalization for cost modelling

@@
   to reduce it back to a graph problem by the edges of the original interaction
-  graph as vertices of an "augmented interaction graph" whose vertex set
+  graph as vertices of an ``augmented interaction graph" whose vertex set
   consists of both the vertices and edges of the original graph, i.e. all
@@
 
+Let $Q$ be the query workload, where each query $q\in Q$ accesses a set of
+attributes $q.A$ and traverses parts of the graph for the time range
+$q.T=[q.t_s,q.t_e]$. We denote the set of all attributes as $A$. A given block
+$B$, we denote its time range as $B.T$, which is the union of the time ranges
+of its temporal neighborlists. Let $s(a)$ denote the size of an attribute $a$.
+We also use the same notation for block size, that is $s(B)$ for a block $B$. 
+We use $c(B)$ to denote the number of entries in the temporal neighborlists
+within block $B$, that is the number of edges in the block.
+
+Our goal is to create an overlapping partitioning of a block, denoted by
+$\mathcal{P}(B)$. We call the partitions \emph{sub-blocks}. We have
+$\bigcup_{B'\in \mathcal{P}(B)} B'.A = A$. We aim to find the function
+$\mathcal{P}$ that minimizes the query I/O over $B$, while keeping the
+relative storage overhead beyond a limit, say $1+\alpha$ times the original.
+If we represent the query I/O as $L(\mathcal{P}(B))$ and the relative storage
+overhead as $H(\mathcal{P}(B))$, our goal is to find:
+\begin{equation}
+\mathcal{P} \leftarrow \mbox{argmax}_{\{\mathcal{P}: H(\mathcal{P}(B)) < \alpha\}} L(\mathcal{P}(B))
+\end{equation}
+
+The storage overhead can be formalized as:
+\begin{equation}
+H(\mathcal{P}(B)) = (1-\frac{\sum_{a\in A} s(a)}{s(B)})\cdot |\mathcal{P}(B)| 
+\end{equation}
+
+The Query I/O can be formalized as:\\
+TODO for Robert.\\
+We need to use the workload characteristics for the blocks time range as
+well as $\mathcal{P}(B)$ to formalize the query I/O.\\
+
 \bibliographystyle{abbrv}

### 2014-04-11 notes/graph-adapt/main.tex
initial formalization for cost modelling
removed rdk note

@@
 
-\emph{In the problem you're describing, one natural idea that springs to mind is
-  to reduce it back to a graph problem by the edges of the original interaction
-  graph as vertices of an ``augmented interaction graph" whose vertex set
-  consists of both the vertices and edges of the original graph, i.e. all
-  entities that might have data associated to them. And two vertices of the
-  augmented interaction graph are joined by an edge if either (a) both of them
-  are vertices of the original interaction graph, and they are joined by an
-  edge, or (b) one is an edge and the other is a vertex which is an endpoint of
-  that edge. Pictorially, the augmented interaction graph is obtained by
-  inserting a new vertex alongside each edge of the original interaction graph
-  and turning the edge into a triangle with the new vertex as its third
-  corner. It seems that this would work well assuming that queries which
-  traverse an edge are likely to traverse its endpoints. -- rdk}
-
 \section{Adaptation}

### 2014-04-15 notes/graph-adapt/main.tex
initial formalization for cost modelling
removed dead text

@@
    couple of queries that would change the layout on disk.
-
-- An interaction graph is an append-only graph, where new 
-   edges and vertices are added as time progresses.
-
-- Typically, in graph dbs, if a vertex is visited during a 
-   traversal, then it is likely that its neighbors will also 
-    be visited
-
-- So, we store we place highly connected neighbors in the 
-    same disk block, to minimize disk I/O.
-
-- Interaction graphs are temporal. The edges between vertices 
-   change over time.
-
-- For interaction graphs, we need to group edges spatially 
-   and temporally.
-
-- Vertices and edges have data associated with them.
-
-- We want to store as much relevant data as possible in the 
-   same block.
-
-- We will store:
-   - edges and vertices that are connected in the same time interval
-   - edges and vertices that are highly connected (neighbors)
-   - edge attributes that are likely to be accessed
-   - vertex attributes that are likely to be accessed
-
-- Example: We use an interaction graph to model connections
-   between people on Twitter. Every time
-
-- $C^d(B) = conductance, lower is better.$
-- $C^h(B) = cohesiveness, higher is better.$
-
 \end{alltt}
-
-$L(B) = \sqrt[2]{C^h(B) \cdot ( 1 - C^d(B))}$
 
@@
 attributes $q.A$ and traverses parts of the graph for the time range
-$q.T=[q.t_s,q.t_e]$. We denote the set of all attributes as $A$. A given block
+$q.T=[q.t_s,q.t_e]$. We denote the set of all attributes as $A$. Given a block
 $B$, we denote its time range as $B.T$, which is the union of the time ranges
@@
 \begin{equation}
-\mathcal{P} \leftarrow \mbox{argmax}_{\{\mathcal{P}: H(\mathcal{P}(B)) < \alpha\}} L(\mathcal{P}(B))
+\mathcal{P} \leftarrow \mbox{argmin}_{\{\mathcal{P}: H(\mathcal{P}(B)) < \alpha\}} L(\mathcal{P}(B))
 \end{equation}
@@
 \begin{equation}
-H(\mathcal{P}(B)) = (1-\frac{\sum_{a\in A} s(a)}{s(B)})\cdot |\mathcal{P}(B)| 
+H(\mathcal{P}(B)) = |\mathcal{P}(B)|\cdot\left(1-\frac{\sum_{a\in A} s(a)}{s(B)}\right) 
 \end{equation}

### 2014-04-16 notes/graph-adapt/main.tex
fix for eq 2
A first attempt at formulizing the query I/O

@@
 \usepackage{alltt}
+\usepackage{enumerate}
 
@@
 \begin{equation}
-H(\mathcal{P}(B)) = |\mathcal{P}(B)|\cdot\left(1-\frac{\sum_{a\in A} s(a)}{s(B)}\right) 
+H(\mathcal{P}(B)) = |\mathcal{P}(B)|\cdot\left(1-\frac{c(B)\cdot \sum_{a\in A} s(a)}{s(B)}\right) 
 \end{equation}
@@
 
+\subsection{Query I/0}
+
+\noindent 
+We have a set of queries:
+
+$$
+Q = \left\{ q_0, \dots,  q_n \right\}
+$$
+
+\noindent 
+And that the queries access set of attributes:
+$$
+Q.A = \left\{ q_0.A, \dots,  q_n.A \right\}
+$$
+
+\noindent 
+Assume that we are given a function $f$ that maps a query to the frequency the
+query is executed:
+$$
+f(q_i)
+$$
+
+To work through a concrete example, say we are given a schema \texttt{abcd}
+which we partition in two ways:
+\begin{enumerate}[(i)]
+\item \texttt{ab}, \texttt{ac}, and \texttt{d} 
+\item \texttt{abd} and \texttt{ac} 
+\end{enumerate}
+
+\noindent 
+and say we are given the following query attribute sets and frequencies:
+$$
+Q.A = \left\{ abd, ac, d \right\}\\
+$$
+$$
+\left\{ f(q_o) \mapsto 0.4,  f(q_1) \mapsto 0.5,  f(q_2) \mapsto 0.1 \right\}
+$$
+Then the I/O cost for (i) should be:
+$$
+L(\mathcal{P}(B))  = 0.4 \cdot (s(ab) + s(d)) + 0.5 \cdot s(ac) + 0.1 \cdot s(d) 
+$$
+Then the I/O cost for (ii) should be:
+$$
+L(\mathcal{P}(B))  = 0.4 \cdot s(abd) + 0.5 \cdot s(ac) + 0.1 \cdot s(abd) 
+$$
+\noindent 
+Generalizing, I think we want an equation like the following:
+$$
+L(\mathcal{P}(B)) = \sum_{q\in Q} \left( f(q) \cdot \sum_{a\in A} s(a) \right)
+$$
+\noindent 
+But, I don't think this is quite right. I have the following questions below:\\
+\textbf{Questions}:
+\begin{enumerate}
+\item The term in the second summation is supposed to be the sizes of all the
+sub-blocks accessed. I am not quite sure how to write this in our notation. 
+\item It might be the case that there are several possible sub-block that we
+  could access to satisfy a query. I think somewhere in the formulation, we need
+  to say that we access the \texttt{min} subblock that will give us the
+  attributes we want.
+\end{enumerate}
+
 \bibliographystyle{abbrv}

### 2014-04-18 notes/graph-adapt/main.tex
block size

@@
 of its temporal neighborlists. Let $s(a)$ denote the size of an attribute $a$.
-We also use the same notation for block size, that is $s(B)$ for a block $B$. 
-We use $c(B)$ to denote the number of entries in the temporal neighborlists
-within block $B$, that is the number of edges in the block.
+We use $c_n(B)$ to denote the number of temporal neighborlists within block
+$B$ and $c_e(B)$ to denote the total number of edges in the temporal
+neighborlists within the block. We overload the notation for block size and
+use $s(B)$ to denote the size of a block $B$. We have: 
+\begin{equation}
+s(B) = c_e(B) \cdot \left(16 + \sum_{a\in A} s(a)\right) + c_n(B) \cdot 12  
+\end{equation}
+Here, $16$ corresponds to the cost of storing the edge id and the timestamp,
+and $qw$ corresponds to the cost of storing the head vertex ($8$ bytes) plus
+the number of entries ($4$ bytes) for a temporal neighborlist.
 
@@
 \begin{equation}
-H(\mathcal{P}(B)) = |\mathcal{P}(B)|\cdot\left(1-\frac{c(B)\cdot \sum_{a\in A} s(a)}{s(B)}\right) 
+H(\mathcal{P}(B)) = |\mathcal{P}(B)|\cdot\left(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\right) 
 \end{equation}

### 2014-04-24 notes/graph-adapt/main.tex
another pass at query i/o

--- 2014-04-15 notes/graph-adapt/main.tex
@@
 \begin{equation}
-H(\mathcal{P}(B)) = |\mathcal{P}(B)|\cdot\left(1-\frac{\sum_{a\in A} s(a)}{s(B)}\right) 
+H(\mathcal{P}(B)) = |\mathcal{P}(B)|\cdot\left(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\right) 
 \end{equation}
@@
 
+\subsection{Query I/0}
+
+We are also given a function $f$ that maps a query to the frequency the
+query is executed during a particular time range $[t_s,t_e]$:
+$$
+f:  \text{query} \rightarrow \text{time} * \text{time} \rightarrow \text{frequency}
+$$
+
+\noindent 
+We are also given a function $m$ that maps a query to the set of sub-blocks that are
+accessed to satisfy that query during a particular time range. 
+$$
+m:  \text{query} \rightarrow \text{time} * \text{time} \rightarrow \text{block set}
+$$
+
+\noindent Define:
+$$
+\bigcup_{B'\in m(q, q.t_s, q.t_e)} B'.A = M
+$$
+
+\noindent 
+Generalizing, I think we want an equation like the following:
+$$
+L(\mathcal{P}(B)) = \sum_{q\in Q} \left( f(q, q.t_s, q.t_e) \cdot \left( \sum_{a \in M}
+  s(a) + \left\vert{m(q, q.t_s, q.t_e)}\right\vert \cdot c(B)\right) \right)
+$$
+
 \bibliographystyle{abbrv}

### 2014-04-24 code/test/SpanningTreeTest/src/Main.cpp
Added a spanning tree implementation.

@@ -1,0 +1,226 @@
+#include <intergdb/core/InteractionGraph.h>
+
+#include <cstdlib>
+#include <iostream>
+#include <climits>
+#include <vector>
+#include <map>
+
+#include <boost/lexical_cast.hpp>
+
+#include <boost/filesystem.hpp>
+
+using namespace std;
+using namespace intergdb::core;
+using boost::lexical_cast;
+
+typedef InteractionGraph<std::string,std::string> Graph;
+
+
+class Key
+{
+  public: 
+    VertexId s;
+    VertexId d;
+    Key(VertexId s, VertexId d)
+    {
+            this->s = s;
+            this->d = d;
+    }
+    bool operator<(const Key& k) const
+    {
+      int s_cmp = this->s < k.s;
+      if(s_cmp == 0)
+      {
+        return this->d < k.d;
+      }
+      return s_cmp < 0;
+    }
+};
+
+
+void createGraph(Graph &graph) 
+{
+    // The example from page 236 of 
+    // Algorithms in C by Sedgewick
+    graph.createVertex(0, "v0");
+    graph.createVertex(1, "v1");
+    graph.createVertex(2, "v2");
+    graph.createVertex(3, "v3");
+    graph.createVertex(4, "v4");
+    graph.createVertex(5, "v5");
+    graph.createVertex(6, "v6");
+    graph.createVertex(7, "v7");
+    Timestamp ts = 1.0;
+    graph.addEdge(3, 5, "1", ts); 
+    graph.addEdge(0, 2, "2", ts);
+    graph.addEdge(0, 7, "3", ts);
+    graph.addEdge(0, 1, "4", ts);
+    graph.addEdge(3, 4, "5", ts);
+    graph.addEdge(5, 4, "6", ts); 
+    graph.addEdge(0, 6, "7", ts);
+    graph.addEdge(7, 1, "8", ts);
+    graph.addEdge(7, 6, "9", ts);
+    graph.addEdge(7, 4, "10", ts);
+    graph.addEdge(6, 4, "11", ts);
+    graph.addEdge(0, 5, "12", ts);
+
+    graph.flush();     
+}
+
+
+void print_wt( map<VertexId,double> &wt) {
+    VertexId key;
+    double value;
+    cout << "wt: " << endl;
+    for ( auto iter : wt ) {
+        key=iter.first;
+        value=iter.second;
+        cout << "  wt[" << key << "]=" << value << endl;
+    }
[…]

### 2014-04-28 notes/graph-adapt/main.tex
query I/O model

@@
 \usepackage{enumerate}
+\usepackage[ruled,vlined]{algorithm2e}
+
+\renewcommand{\CommentSty}[1]{\textnormal{#1}}
+\DontPrintSemicolon
+\SetKwComment{tcp}{$\triangleright$ }{}
+\SetVlineSkip{0cm}
+\SetAlgoSkip{}
+
+\let\oldemptyset\emptyset
+\let\emptyset\varnothing
 
@@
 \conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
-
 \title{Adaptive Disk Storage for Interaction Graphs}
-
 \numberofauthors{1} %  in this sample file, there are a *total*
 
-\author{ }
+\author{}
 
@@
 \section{Cost Model}
-
 Let $Q$ be the query workload, where each query $q\in Q$ accesses a set of
@@
 relative storage overhead beyond a limit, say $1+\alpha$ times the original.
-If we represent the query I/O as $L(\mathcal{P}(B))$ and the relative storage
-overhead as $H(\mathcal{P}(B))$, our goal is to find:
+If we represent the query I/O as $L(\mathcal{P}, B)$ and the relative storage
+overhead as $H(\mathcal{P}, B)$, our goal is to find:
 \begin{equation}
-\mathcal{P} \leftarrow \mbox{argmin}_{\{\mathcal{P}: H(\mathcal{P}(B)) < \alpha\}} L(\mathcal{P}(B))
+\mathcal{P} \leftarrow \mbox{argmin}_{\{\mathcal{P}: H(\mathcal{P}(B)) < \alpha\}} L(\mathcal{P},B)
 \end{equation}
@@
 \begin{equation}
-H(\mathcal{P}(B)) = |\mathcal{P}(B)|\cdot\left(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\right) 
+H(\mathcal{P}, B) = |\mathcal{P}(B)|\cdot\left(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\right) 
 \end{equation}
 
-The Query I/O can be formalized as:\\
-TODO for Robert.\\
-We need to use the workload characteristics for the blocks time range as
-well as $\mathcal{P}(B)$ to formalize the query I/O.\\
+\subsection{Query I/O}
 
-\subsection{Query I/0}
+Let $m$ be a function that maps a query $q$ to the set of sub-blocks that are
+accessed to satisfy it for a relevant block $B$ under a given partitioning
+$\mathcal{P}$. For the case of non-overlapping attributes, we have:
+\begin{equation}
+m(\mathcal{P}, B, q) = \{B': B'\in \mathcal{P}(B) \wedge q.A \cap B'.A \ne \emptyset\}  
+\end{equation}
 
-We are also given a function $f$ that maps a query to the frequency the
-query is executed during a particular time range $[t_s,t_e]$:
-$$
-f:  \text{query} \rightarrow \text{time} * \text{time} \rightarrow \text{frequency}
-$$
+For the case of overlapping attributes, we use a simple heuristic to define the
+set of sub-blocks to be used for answering the query.
+Algorithm~\ref{alg:greedyM} captures it. The idea is to select sub-blocks that
+bring the highest relative marginal gain, which is defined as the size of
+attribute data that contributes to the query result, relative to the sub-block
+size. While computing the marginal gain, attributes that are covered by
+sub-blocks that are already selected are not considered.
+\begin{algorithm}[h]
+\scriptsize
+\caption{m-overlapping($\mathcal{P}, B, q$)}
+\label{alg:greedyM}
+\KwData{$\mathcal{P}$: partitioning function, $B$: block, $q$: query}
+$S\leftarrow \emptyset; R\leftarrow \emptyset$ \tcp*{Selected attributes; Resulting sub-blocks}
+\While(\tcp*[f]{While unselected attributes remain}){$S \subset q.A$}{
+  $B' \leftarrow \mbox{argmax}_{B'\in\mathcal{P}(B)\setminus R} \sum_{a\in B'.A \cap q.A \setminus S} \frac{c_e(B') \cdot s(a)}{s(B')}$
+  $S \leftarrow S \cup B'.A$\tcp*{Extend the selected attributes}
+  $R\leftarrow R \cup B'$\tcp*{Extend the selected sub-blocks}
+}
+\Return R \tcp*{Final set of sub-blocks covering the query attributes}
+\end{algorithm}
 
-\noindent 
-We are also given a function $m$ that maps a query to the set of sub-blocks that are
-accessed to satisfy that query during a particular time range. 
-$$
-m:  \text{query} \rightarrow \text{time} * \text{time} \rightarrow \text{block set}
-$$
+The query I/O cost for a block is then given by:
+\begin{equation}
+L(\mathcal{P}, B) = \sum_{q\in Q} \mathbf{1}(Q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} s(B')
+\end{equation}
 
-\noindent Define:
-$$
-\bigcup_{B'\in m(q, q.t_s, q.t_e)} B'.A = M
-$$
+\noindent The following are unused, but may be needed for the partitioning
+algorithm.
 
-\noindent 
-Generalizing, I think we want an equation like the following:
-$$
-L(\mathcal{P}(B)) = \sum_{q\in Q} \left( f(q, q.t_s, q.t_e) \cdot \left( \sum_{a \in M}
-  s(a) + \left\vert{m(q, q.t_s, q.t_e)}\right\vert \cdot c(B)\right) \right)
-$$
+Let $f: \mathbb{R}\times A \rightarrow \mathbb{N}$ be a function that maps a
+attribute-timestamp pair $(a, t)$ to the frequency of queries that access
+attribute $a$ for time $t$. More formally:
+\begin{equation}
+f(a,t) = \sum_{q\in Q} \textbf{1}(a\in q.A \wedge t\in q.T)
+\end{equation}
+
+Let $T$ represent a time range $[T.s, T.e]$. We denote the frequency of
+attribute access for the time interval $T$ for attribute $a$ as follows:
+\begin{equation}
+f(a, T) = \int_{T.s}^{T.e} f(a, t)\cdot dt =  \sum_{q\in Q} |q.T \cap T| 
+\end{equation}
 

### 2014-04-30 notes/graph-adapt/main.tex
added algorithm for partitioning with overlapping

--- 2014-04-24 notes/graph-adapt/main.tex
@@
 
+\begin{itemize}
+
+\item First, let's define a measure for the similarity between two
+partitions. I think this can be the sum of the size of attributes that
+they do not have in common. So, if we have partitions: abc, abd, and
+abe, then abc-abd will be more similar than abc-abe, if size(d) <
+size(e).
+
+\item The first partitioning is based on what queries we have seen.
+Every query gets its own sub-block. This is the "ideal" situation,
+because the I/O cost would be minimized for every query that we would
+have seen.
+
+\item We calculate the storage cost for the partitioning.
+
+\item If the storage cost is above some threshold, then we do a pairwise
+comparison of all sub-blocks to determine which two are the most
+similar, using (0).
+
+\item We combine the two most-similar sub-blocks, and repeat the process
+starting at (2) until we are satisfied with the storage cost.
+
+\item The result is our partitioned block.
+
+\end{itemize}
+
 \section{Cost Model}

### 2014-04-30 notes/graph-adapt/main.tex

--- 2014-04-28 notes/graph-adapt/main.tex
@@
 \section{Adaptation}
+
+\begin{itemize}
+
+\item First, let's define a measure for the similarity between two
+partitions. I think this can be the sum of the size of attributes that
+they do not have in common. So, if we have partitions: abc, abd, and
+abe, then abc-abd will be more similar than abc-abe, if size(d) <
+size(e).
+
+\item The first partitioning is based on what queries we have seen.
+Every query gets its own sub-block. This is the "ideal" situation,
+because the I/O cost would be minimized for every query that we would
+have seen.
+
+\item We calculate the storage cost for the partitioning.
+
+\item If the storage cost is above some threshold, then we do a pairwise
+comparison of all sub-blocks to determine which two are the most
+similar, using (0).
+
+\item We combine the two most-similar sub-blocks, and repeat the process
+starting at (2) until we are satisfied with the storage cost.
+
+\item The result is our partitioned block.
+
+\end{itemize}
 

### 2014-04-30 notes/graph-adapt/main.tex
switched to JLDBJ style

@@
-\documentclass{sig-alternate}
+\documentclass[twocolumn]{svjour3}          % twocolumn
+%
+\smartqed  % flush right qed marks, e.g. at end of proof
+%
 
@@
 \usepackage[ruled,vlined]{algorithm2e}
+\usepackage{xargs}
+\usepackage{amssymb}
 
@@
 
+% Insert the name of "your journal" with
+\journalname{VLDBJ}
+%
 \begin{document}
 
-\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
 \title{Adaptive Disk Storage for Interaction Graphs}
-\numberofauthors{1} %  in this sample file, there are a *total*
 
-\author{}
+%\titlerunning{Short form of title}        % if too long for running head
 
-\date{}
+\author{First Author         \and
+        Second Author %etc.
+}
+
+%\authorrunning{Short form of author list} % if too long for running head
+
+\institute{F. Author \at
+              first address \\
+              Tel.: +123-45-678910\\
+              Fax: +123-45-678910\\
+              \email{fauthor@example.com}           %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+           \and
+           S. Author \at
+              second address
+}
+
+\date{Received: date / Accepted: date}
+% The correct dates will be entered by the editor
 
 \maketitle
+
 \begin{abstract}
-TODO
+Insert your abstract here. Include keywords, PACS and mathematical
+subject classification numbers as needed.
+\keywords{First keyword \and Second keyword \and More}
+% \PACS{PACS code1 \and PACS code2 \and more}
+% \subclass{MSC code1 \and MSC code2 \and more}
 \end{abstract}
@@
 In an interaction graph database organized as blocks of temporal neighborlists
-(as in my earlier paper), it is important to store the edge and vertex
+(as in Gedik et al.~\cite{gedik14}), it is important to store the edge and vertex
 properties locally. For  instance, if the edge properties are kept away in a
 relational table, there will be almost no benefit to the locality
-optimizations performed for block organization (discussed in \cite{gedik14}), as we would have to go back and forth between the disk blocks to
+optimizations performed for block organization (discussed in Gedik et al.~\cite{gedik14}, as we would have to go back and forth between the disk blocks to
 access the edge attributes. 
@@
 \end{alltt}
-
-\section{Adaptation}
-
-\begin{itemize}
-
-\item First, let's define a measure for the similarity between two
-partitions. I think this can be the sum of the size of attributes that
-they do not have in common. So, if we have partitions: abc, abd, and
-abe, then abc-abd will be more similar than abc-abe, if size(d) <
-size(e).
-
-\item The first partitioning is based on what queries we have seen.
-Every query gets its own sub-block. This is the "ideal" situation,
-because the I/O cost would be minimized for every query that we would
-have seen.
-
-\item We calculate the storage cost for the partitioning.
-
-\item If the storage cost is above some threshold, then we do a pairwise
-comparison of all sub-blocks to determine which two are the most
-similar, using (0).
-
-\item We combine the two most-similar sub-blocks, and repeat the process
-starting at (2) until we are satisfied with the storage cost.
-
-\item The result is our partitioned block.
-
-\end{itemize}
 
@@
 
-\bibliographystyle{abbrv}
+\section{Adaptation}
+
+\subsection{Non-Overlapping Attributes}
+
+%\linprog{f(x)}{g_i(x) \leq 0}{\eqnlimit[1]{i}[n]}{h_j(x) = 0}{\eqnlimit[1]{j}[m]}{eq:linprog1}
+
+\subsection{Overlapping Attributes}
+
+\begin{itemize}
+
+\item First, let's define a measure for the similarity between two
+partitions. I think this can be the sum of the size of attributes that
+they do not have in common. So, if we have partitions: abc, abd, and
+abe, then abc-abd will be more similar than abc-abe, if size(d) <
+size(e).
+
+\item The first partitioning is based on what queries we have seen.
+Every query gets its own sub-block. This is the "ideal" situation,
+because the I/O cost would be minimized for every query that we would
+have seen.
+
+\item We calculate the storage cost for the partitioning.
+
+\item If the storage cost is above some threshold, then we do a pairwise
+comparison of all sub-blocks to determine which two are the most
+similar, using (0).
+
+\item We combine the two most-similar sub-blocks, and repeat the process
+starting at (2) until we are satisfied with the storage cost.
+
+\item The result is our partitioned block.
+
+\end{itemize}
+
+%\begin{acknowledgements}
+%If you'd like to thank anyone, place your comments here
+%and remove the percent signs.
+%\end{acknowledgements}
+
+\bibliographystyle{spmpsci}
 \bibliography{main}
@@
 \end{document}
+% end of file template.tex

### 2014-05-01 notes/graph-adapt/main.tex
Bad attempt at writing up the algorithm for partitioning
minor fix

--- 2014-04-30 notes/graph-adapt/main.tex
@@
 
-\begin{itemize}
-
-\item First, let's define a measure for the similarity between two
-partitions. I think this can be the sum of the size of attributes that
-they do not have in common. So, if we have partitions: abc, abd, and
-abe, then abc-abd will be more similar than abc-abe, if size(d) <
-size(e).
-
-\item The first partitioning is based on what queries we have seen.
-Every query gets its own sub-block. This is the "ideal" situation,
-because the I/O cost would be minimized for every query that we would
-have seen.
-
-\item We calculate the storage cost for the partitioning.
-
-\item If the storage cost is above some threshold, then we do a pairwise
-comparison of all sub-blocks to determine which two are the most
-similar, using (0).
-
-\item We combine the two most-similar sub-blocks, and repeat the process
-starting at (2) until we are satisfied with the storage cost.
-
-\item The result is our partitioned block.
-
-\end{itemize}
-
 \section{Cost Model}
@@
 \begin{equation}
-L(\mathcal{P}, B) = \sum_{q\in Q} \mathbf{1}(Q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} s(B')
+L(\mathcal{P}, B) = \sum_{q\in Q} \mathbf{1}(q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} s(B')
 \end{equation}

### 2014-05-01 notes/graph-adapt/main.tex

--- 2014-04-30 notes/graph-adapt/main.tex
@@
 \usepackage{amssymb}
+\usepackage{amsmath}
 
@@
 
-\author{First Author         \and
-        Second Author %etc.
+\author{Bu\u{g}ra Gedik         \and
+        Robert Soul\'{e}%etc.
 }
@@
 
-\institute{F. Author \at
-              first address \\
-              Tel.: +123-45-678910\\
-              Fax: +123-45-678910\\
-              \email{fauthor@example.com}           %  \\
-%             \emph{Present address:} of F. Author  %  if needed
+\institute{B. Gedik \at            
+              \email{bgedik@cs.bilkent.edu.tr}           %  \\
            \and
-           S. Author \at
-              second address
+           R. Soul\'{e} \at
+               \email{robert.soule@usi.ch}     
 }
@@
 sub-blocks that are already selected are not considered.
+%
 \begin{algorithm}[h]
@@
 \end{algorithm} 
-
+%
 The query I/O cost for a block is then given by:
 \begin{equation}
-L(\mathcal{P}, B) = \sum_{q\in Q} \mathbf{1}(Q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} s(B')
+L(\mathcal{P}, B) = \sum_{q\in Q} \mathbf{1}(q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} s(B')
 \end{equation}
@@
 %\linprog{f(x)}{g_i(x) \leq 0}{\eqnlimit[1]{i}[n]}{h_j(x) = 0}{\eqnlimit[1]{j}[m]}{eq:linprog1}
+
+ \begin{alignat*}{2}
+    \text{minimize: }   & L(\mathcal{P}, B)   \\
+    \text{subject to: } & H(\mathcal{P}, B) \leq \text{Max Storage}
+  \end{alignat*}
 
@@
 
+\begin{algorithm}[h]
+\scriptsize
+\caption{Algorithm for partitioning blocks into sub-blocks with overlapping attributes.}
+\label{alg:overlappingP}
+\KwData{$B$: block, $Q$: set of queries}
+$\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}
+
+\For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
+ $\mathcal{P}(B) = \mathcal{P}(B) \cup q.A$ 
+}
+
+\While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq \text{max}$}{
+    $min = \infty$ \; 
+    $x = \emptyset$ \;
+    $y = \emptyset$ \;
+    \For{$b_1 \in \mathcal{P}(B)$}{ 
+        \For{$b_2 \in \mathcal{P}(B)$}{ 
+            $\delta = (b_1 \cup b_2) - (b_1 \cap b_2)$ \; 
+           \If{$min < \delta $}{
+                 $\text{x} \leftarrow  b_1$\;
+                 $\text{y} \leftarrow b_2$\;
+                 $min \leftarrow \delta$
+            } 
+        }
+    }
+$ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - x - y$ \;
+ $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  + x \cup y$ \;
+}
+\Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
+\end{algorithm}
+
 %\begin{acknowledgements}

### 2014-05-07 docs/graph-adapt/main.tex
Re-organized directory structure
updated alternative algorithm

--- 2014-05-01 notes/graph-adapt/main.tex
@@
 
+\paragraph{First proposal}
+
 \begin{itemize}
@@
 they do not have in common. So, if we have partitions: abc, abd, and
-abe, then abc-abd will be more similar than abc-abe, if size(d) <
+abe, then abc-abd will be more similar than abc-abe, if size(d) $<$
 size(e).
@@
 
+\paragraph{Second proposal}
+
+\begin{itemize}
+
+\item The first partitioning is based on what queries we have seen.
+Every query gets its own sub-block. This is the "ideal" situation,
+because the I/O cost would be minimized for every query that we would
+have seen.
+
+\item We keep a mapping of queries to blocks and frequencies.
+
+\item If the storage cost is above some threshold, then we get the queries with
+  the lowest frequencies, and combine their blocks. We also add the frequencies,
+  so we are combining the blocks that are least likely to be accessed.
+
+\item We repeat until we are below the storage threshold.
+
+\item The result is our partitioned block.
+
+\end{itemize}
+
+\begin{algorithm}[h]
+\scriptsize
+\caption{Alternative algorithm for partitioning blocks into sub-blocks with overlapping attributes.}
+\label{alg:overlappingP2}
+\SetKwFunction{Enqueue}{Enqueue}
+\SetKwFunction{Dequeue}{Dequeue}
+
+\KwData{$B$: block, $Q$: set of queries}
+$\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}
+$pq \leftarrow \emptyset;$  \tcp*{priority queue;}
+
+\For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
+ $\mathcal{P}(B) = \mathcal{P}(B) \cup q.A$ \;
+ \Enqueue{$pq,\mathcal{P}(B), f(\mathcal{P}(B))$}\tcp*{put blocks in the queue;}
+}
+\While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
+  \text{max}$}{  
+  $x \leftarrow $ \Dequeue{$pq$}\;
+$y \leftarrow $ \Dequeue{$pq$}\tcp*{remove the two minimum;}
+  \Enqueue{$pq,x \cup y, f(x) + f(y)$}\tcp*{combine and insert;}    
+    }
+
+\For(\tcp*[f]{The elements in the queue are our partition}){$p \in pq$}{ 
+  $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) \cup p$ \;
+}
+\Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
+\end{algorithm}
+
 %\begin{acknowledgements}

### 2014-05-15 docs/graph-adapt/main.tex
updated problem description

@@
 % 
-Typical interaction graphs have data associated with their vertices and edges.
+Typical interaction graphs have data associated with their vertices and
+edges. We call this data \emph{attributes} or \emph{properties}.
 Most algorithms access this data as they traverse the graph. However, not all
@@
 optimizations performed for block organization (discussed in Gedik et al.~\cite{gedik14}, as we would have to go back and forth between the disk blocks to
-access the edge attributes. 
+access the edge attributes). 
 %
@@
 
-Recall that interaction graphs are temporal. As such, the co-access
-correlations for the attributes can be different for different temporal
-regions. It might also  be unknown at the insertion time and may be discovered
-later depending on the workload. This points to the need for adaptively
-optimizing the layout (somewhat similar to the H2O paper).
+An important characteristic of interaction graphs are that they are temporal.
+As such, the co-access correlations for the attributes can be different for
+different temporal regions. It might also be unknown at the insertion time and
+may be discovered later depending on the workload. This points to the need for
+adaptively optimizing the layout (somewhat similar to H2O \cite{alagiannis14}
+and HYRISE~\cite{grund10}).
 %
@@
  different subset of the attributes, but with a link between them in case a
-query needs to access both. Some attributes can be replicated as well. 
-More details to follow about all this...
+query needs to access both.
 
-\section{Example}
+\section{Problem Description}
 
-\begin{alltt}\scriptsize
-- Work through a concrete example of a database and a 
-   couple of queries that would change the layout on disk.
-\end{alltt}
+The problem is to determine a partitioning of the disk blocks into sub-blocks
+that would minimize query I/O. Of course, partitioning into sub-blocks results
+in an increase in storage cost. The storage cost must stay below some specified limit.
+
+There are two possibilities: \emph{overlapping}
+and \emph{non-overlapping} partitioning. In the overlapping case, attributes can
+be replicated in the sub-blocks. The non-overlapping case is a true partition.
+
+We start by defining the cost model, which formalizes the query I/O and storage costs.
 
@@
 
+\paragraph{\textbf{Other useful equations}}
+
 \noindent The following are unused, but may be needed for the partitioning
@@
 
-\section{Adaptation}
+\section{Partitioning}
+
+TODO: This section is still a work-in-progress. We are trying to find the algorithms.
 
@@
 
-%\linprog{f(x)}{g_i(x) \leq 0}{\eqnlimit[1]{i}[n]}{h_j(x) = 0}{\eqnlimit[1]{j}[m]}{eq:linprog1}
-
- \begin{alignat*}{2}
-    \text{minimize: }   & L(\mathcal{P}, B)   \\
-    \text{subject to: } & H(\mathcal{P}, B) \leq \text{Max Storage}
-  \end{alignat*}
+\textbf{Problem:} \emph{Find a true partitioning of attributes that minimizes
+  the query I/O and bounds the storage cost by some upper limit.}
 
 \subsection{Overlapping Attributes}
+
+\textbf{Problem:} \emph{Find an overlapping partitioning of attributes that minimizes
+  the query I/O and bounds the storage cost by some upper limit.}
+
+Below, we sketch two proposed algorithms. We are still working on these.
 

### 2014-06-05 docs/hasler/main.tex

--- 2014-05-01 notes/graph-adapt/main.tex
@@
-\documentclass{sig-alternate}
-
+\documentclass[11pt]{article}
+\usepackage[margin=1in]{geometry}
 \usepackage{alltt}
-\usepackage{enumerate}
-\usepackage[ruled,vlined]{algorithm2e}
-
-\renewcommand{\CommentSty}[1]{\textnormal{#1}}
-\DontPrintSemicolon
-\SetKwComment{tcp}{$\triangleright$ }{}
-\SetVlineSkip{0cm}
-\SetAlgoSkip{}
-
-\let\oldemptyset\emptyset
-\let\emptyset\varnothing
+\usepackage{balance}
+\usepackage{url}
 
@@
 
-\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
-\title{Adaptive Disk Storage for Interaction Graphs}
-\numberofauthors{1} %  in this sample file, there are a *total*
+% \begin{alltt}\scriptsize
+%  Concise project description
+% 2.1. Summary
+% Max. one page
+% 2.2. Project outline
+% Only a few pages
+% - Field of research
+% - Goals and objectives
+% - Significance of the project
+% 2.3. Work plan
+% \end{alltt}
 
-\author{}
+\section*{Adaptive Disk Storage for Interaction Graphs}
 
-\date{}
+\section{Summary}
 
-\maketitle
-\begin{abstract}
-TODO
-\end{abstract}
-
-\section{Introduction}
-An \emph{interaction graph} is an append-only graph, where new edges and
-vertices are added as time progresses. 
-%
+Graph databases are becoming increasingly popular for applications in the world-wide web, social networks, and telecommunications.
+An interaction graph is an append-only graph, where new edges and vertices are added as time progresses.
+A good example of an interaction graph is Twitter, where users represent vertices, and edges are mentions of other users
 
 %
-Typical interaction graphs have data associated with their vertices and edges.
-Most algorithms access this data as they traverse the graph. However, not all
-of the data is accessed by all the queries. Typically, there are correlations
-among the attributes accessed by different queries, such as Q1 and Q5
-accessing attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and
-so on.
-%
-
 In an interaction graph database organized as blocks of temporal neighborlists
-(as in my earlier paper), it is important to store the edge and vertex
+(as in Gedik et al.~\cite{gedik14}), it is important to store the edge and vertex
 properties locally. For  instance, if the edge properties are kept away in a
 relational table, there will be almost no benefit to the locality
-optimizations performed for block organization (discussed in \cite{gedik14}), as we would have to go back and forth between the disk blocks to
+optimizations performed for block organization (discussed in Gedik et al.~\cite{gedik14}, as we would have to go back and forth between the disk blocks to
 access the edge attributes.
@@
 later depending on the workload. This points to the need for adaptively
-optimizing the layout (somewhat similar to the H2O paper).
+optimizing the layout (somewhat similar to the H2O~\cite{alagiannis14}).
 %
@@
 
-\section{Example}
+\section{Project outline}
 
+\subsection{Field of research}
+
+Graph Database work:
 \begin{alltt}\scriptsize
-- Work through a concrete example of a database and a 
-   couple of queries that would change the layout on disk.
+- Google’s Pregel~\cite{malewicz10}
+- Apache Giraph~\cite{giraph}
+- GPS~\cite{salihoglu13}
+- Trinity~\cite{shao13}
+- PowerGraph~\cite{gonzalez12}
+- Prabhakaran et al.~\cite{prabhakaran12}
+- X-Stream~\cite{roy13}
+- Xie et al. ~\cite{xie13}
+- Neo4j ~\cite{neo4j}
+- HyperGraphDB ~\cite{hypergraphdb}
+- Tao (Facebook)~\cite{venkataramani12}
+- Kineograph~\cite{cheng12}
+-  Titan ~\cite{titan}
 \end{alltt}
 
-\section{Adaptation}
+Adaptive storage:
+\begin{alltt}\scriptsize
+H2O: A Hands-free Adaptive Store~cite{alagiannis14}
+http://stratos.seas.harvard.edu/publications/h2o-hands-free-adaptive-store
+\end{alltt}
 
-\section{Cost Model}
-Let $Q$ be the query workload, where each query $q\in Q$ accesses a set of
-attributes $q.A$ and traverses parts of the graph for the time range
-$q.T=[q.t_s,q.t_e]$. We denote the set of all attributes as $A$. Given a block
-$B$, we denote its time range as $B.T$, which is the union of the time ranges
-of its temporal neighborlists. Let $s(a)$ denote the size of an attribute $a$.
-We use $c_n(B)$ to denote the number of temporal neighborlists within block
-$B$ and $c_e(B)$ to denote the total number of edges in the temporal
-neighborlists within the block. We overload the notation for block size and
-use $s(B)$ to denote the size of a block $B$. We have: 
-\begin{equation}
-s(B) = c_e(B) \cdot \left(16 + \sum_{a\in A} s(a)\right) + c_n(B) \cdot 12  
-\end{equation}
-Here, $16$ corresponds to the cost of storing the edge id and the timestamp,
-and $qw$ corresponds to the cost of storing the head vertex ($8$ bytes) plus
-the number of entries ($4$ bytes) for a temporal neighborlist.
+\subsection{Goals and objectives}
+\subsection{Significance of the project}
 
-Our goal is to create an overlapping partitioning of a block, denoted by
-$\mathcal{P}(B)$. We call the partitions \emph{sub-blocks}. We have
-$\bigcup_{B'\in \mathcal{P}(B)} B'.A = A$. We aim to find the function
-$\mathcal{P}$ that minimizes the query I/O over $B$, while keeping the
-relative storage overhead beyond a limit, say $1+\alpha$ times the original.
-If we represent the query I/O as $L(\mathcal{P}, B)$ and the relative storage
-overhead as $H(\mathcal{P}, B)$, our goal is to find:
-\begin{equation}
-\mathcal{P} \leftarrow \mbox{argmin}_{\{\mathcal{P}: H(\mathcal{P}(B)) < \alpha\}} L(\mathcal{P},B)
-\end{equation}
+\section{Work plan}
 
-The storage overhead can be formalized as:
-\begin{equation}
-H(\mathcal{P}, B) = |\mathcal{P}(B)|\cdot\left(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\right) 
-\end{equation}
+The work described in this proposal consists of five threads of activity.
+\begin{itemize}
+\item \emph{Define cost model}
+\item \emph{Design partitioning algorithm}
+\item \emph{Implement prototype}
+\item \emph{Micro-benchmarks}
+\item \emph{Real work applications}
+\end{itemize}
 
-\subsection{Query I/O}
-
-Let $m$ be a function that maps a query $q$ to the set of sub-blocks that are
-accessed to satisfy it for a relevant block $B$ under a given partitioning
-$\mathcal{P}$. For the case of non-overlapping attributes, we have:
-\begin{equation}
-m(\mathcal{P}, B, q) = \{B': B'\in \mathcal{P}(B) \wedge q.A \cap B'.A \ne \emptyset\}  
-\end{equation}
-
-For the case of overlapping attributes, we use a simple heuristic to define the
-set of sub-blocks to be used for answering the query.
-Algorithm~\ref{alg:greedyM} captures it. The idea is to select sub-blocks that
-bring the highest relative marginal gain, which is defined as the size of
-attribute data that contributes to the query result, relative to the sub-block
-size. While computing the marginal gain, attributes that are covered by
-sub-blocks that are already selected are not considered.
-\begin{algorithm}[h]
-\scriptsize
-\caption{m-overlapping($\mathcal{P}, B, q$)}
-\label{alg:greedyM}
-\KwData{$\mathcal{P}$: partitioning function, $B$: block, $q$: query}
-$S\leftarrow \emptyset; R\leftarrow \emptyset$ \tcp*{Selected attributes; Resulting sub-blocks}
-\While(\tcp*[f]{While unselected attributes remain}){$S \subset q.A$}{
-  $B' \leftarrow \mbox{argmax}_{B'\in\mathcal{P}(B)\setminus R} \sum_{a\in B'.A \cap q.A \setminus S} \frac{c_e(B') \cdot s(a)}{s(B')}$
-  $S \leftarrow S \cup B'.A$\tcp*{Extend the selected attributes}
-  $R\leftarrow R \cup B'$\tcp*{Extend the selected sub-blocks}
+{\footnotesize
+\bibliographystyle{abbrv}
+\balance
+\bibliography{main}
 }
-\Return R \tcp*{Final set of sub-blocks covering the query attributes}
-\end{algorithm}
-
-The query I/O cost for a block is then given by:
-\begin{equation}
-L(\mathcal{P}, B) = \sum_{q\in Q} \mathbf{1}(q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} s(B')
-\end{equation}
-
-\noindent The following are unused, but may be needed for the partitioning
-algorithm.
-
-Let $f: \mathbb{R}\times A \rightarrow \mathbb{N}$ be a function that maps a
-attribute-timestamp pair $(a, t)$ to the frequency of queries that access
-attribute $a$ for time $t$. More formally:
-\begin{equation}
-f(a,t) = \sum_{q\in Q} \textbf{1}(a\in q.A \wedge t\in q.T)
-\end{equation}
-
-Let $T$ represent a time range $[T.s, T.e]$. We denote the frequency of
-attribute access for the time interval $T$ for attribute $a$ as follows:
-\begin{equation}
-f(a, T) = \int_{T.s}^{T.e} f(a, t)\cdot dt =  \sum_{q\in Q} |q.T \cap T| 
-\end{equation}
-
-\bibliographystyle{abbrv}
-\bibliography{main}
 

### 2014-06-05 docs/graph-adapt/main.tex
qw -> 12

@@
 An \emph{interaction graph} is an append-only graph, where new edges and
-vertices are added as time progresses. 
-%
+vertices are added as time progresses.%
 
-% 
-Typical interaction graphs have data associated with their vertices and
-edges. We call this data \emph{attributes} or \emph{properties}.
-Most algorithms access this data as they traverse the graph. However, not all
-of the data is accessed by all the queries. Typically, there are correlations
-among the attributes accessed by different queries, such as Q1 and Q5
-accessing attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and
-so on.
-%
+Typical interaction graphs have data associated with their vertices and edges.
+We call this data \emph{attributes} or \emph{properties}. Most algorithms
+access this data as they traverse the graph. However, not all of the data is
+accessed by all the queries. Typically, there are correlations among the
+attributes accessed by different queries, such as Q1 and Q5 accessing
+attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and so on.%
 
 In an interaction graph database organized as blocks of temporal neighborlists
-(as in Gedik et al.~\cite{gedik14}), it is important to store the edge and vertex
-properties locally. For  instance, if the edge properties are kept away in a
-relational table, there will be almost no benefit to the locality
-optimizations performed for block organization (discussed in Gedik et al.~\cite{gedik14}, as we would have to go back and forth between the disk blocks to
-access the edge attributes). 
-%
+(as in Gedik et al.~\cite{gedik14}), it is important to store the edge and
+vertex properties locally. For  instance, if the edge properties are kept away
+in a relational table, there will be almost no benefit to the locality
+optimizations performed for block organization (discussed in Gedik et
+al.~\cite{gedik14}, as we would have to go back and forth between the disk
+blocks to access the edge attributes).%
 
@@
 Unfortunately, there is no clear correspondence to a column-oriented database
-layout for the graph databases.
-%
+layout for the graph databases.%
 
@@
 adaptively optimizing the layout (somewhat similar to H2O \cite{alagiannis14}
-and HYRISE~\cite{grund10}).
-%
+and HYRISE~\cite{grund10}).%
 
-I imagine a solution to this problem, which I name the 'rail layout'. The idea
+I imagine a solution to this problem, which I name the `rail layout'. The idea
 is to start with large blocks that contain the entire data. As we learn about
@@
 that would minimize query I/O. Of course, partitioning into sub-blocks results
-in an increase in storage cost. The storage cost must stay below some specified limit.
+in an increase in storage cost. The storage cost must stay below some
+specified limit.
 
-There are two possibilities: \emph{overlapping}
-and \emph{non-overlapping} partitioning. In the overlapping case, attributes can
-be replicated in the sub-blocks. The non-overlapping case is a true partition.
+There are two possibilities: \emph{overlapping} and \emph{non-overlapping}
+partitioning. In the overlapping case, attributes can be replicated in the
+sub-blocks. The non-overlapping case is a true partition.
 
-We start by defining the cost model, which formalizes the query I/O and storage costs.
+We start by defining the cost model, which formalizes the query I/O and storage
+costs.
 
@@
 Here, $16$ corresponds to the cost of storing the edge id and the timestamp,
-and $qw$ corresponds to the cost of storing the head vertex ($8$ bytes) plus
+and $12$ corresponds to the cost of storing the head vertex ($8$ bytes) plus
 the number of entries ($4$ bytes) for a temporal neighborlist.

### 2014-06-05 docs/graph-adapt/main.tex
idea

@@
 
-Our goal is to create an overlapping partitioning of a block, denoted by
-$\mathcal{P}(B)$. We call the partitions \emph{sub-blocks}. We have
+Our goal is to create a potentially overlapping partitioning of a block,
+denoted by $\mathcal{P}(B)$. We call the partitions \emph{sub-blocks}. We have
 $\bigcup_{B'\in \mathcal{P}(B)} B'.A = A$. We aim to find the function
@@
 \begin{equation}
-L(\mathcal{P}, B) = \sum_{q\in Q} \mathbf{1}(q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} s(B')
+L(\mathcal{P}, B) = \sum_{q\in Q} \mathbf{1}(q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} \!\!s(B')
 \end{equation}
@@
   the query I/O and bounds the storage cost by some upper limit.}
+
+Idea1: Start with each attribute on its own, but as a set. At each step,
+consider  merging two existing sets, but pick the pair that results in the 
+maximum reduction in query cost per overhead removed.
 

### 2014-06-23 docs/graph-adapt/main.tex
authos
Added a more generalized overlapping attributes algorithm

@@
 \let\emptyset\varnothing
+
+\DeclareMathOperator{\dist}{dist}
 
@@
 
-\textbf{Problem:} \emph{Find a true partitioning of attributes that minimizes
+\paragraph*{Problem.}\emph{Find a true partitioning of attributes that minimizes
   the query I/O and bounds the storage cost by some upper limit.}
+
+\paragraph*{Approach.}
 
@@
 
-\textbf{Problem:} \emph{Find an overlapping partitioning of attributes that minimizes
+\paragraph*{Problem.}\emph{Find an overlapping partitioning of attributes that minimizes
   the query I/O and bounds the storage cost by some upper limit.}
 
-Below, we sketch two proposed algorithms. We are still working on these.
+\paragraph*{Approach.} We start the algorithm with a 
+ partitioning based on what queries we have seen.
+Every query gets its own sub-block. This is the``ideal'' partitioning,
+because the I/O cost would be minimized for every query that we would
+have seen. As input to the algorithm, we provide a function $\dist$, which
+is used to compute the distance between two blocks.
 
-\paragraph{First proposal}
+$$
+\dist: block \rightarrow block \rightarrow int
+$$
 
-\begin{itemize}
+Note that the $\dist$ function can be chosen arbitrarily.  Examples of useful
+$\dist$ functions are the frequencies with which the queries are accessed, or the Jaccard index, which measures the set-similarity of
+the attributes.
 
-\item First, let's define a measure for the similarity between two
-partitions. I think this can be the sum of the size of attributes that
-they do not have in common. So, if we have partitions: abc, abd, and
-abe, then abc-abd will be more similar than abc-abe, if size(d) $<$
-size(e).
-
-\item The first partitioning is based on what queries we have seen.
-Every query gets its own sub-block. This is the "ideal" situation,
-because the I/O cost would be minimized for every query that we would
-have seen.
-
-\item We calculate the storage cost for the partitioning.
-
-\item If the storage cost is above some threshold, then we do a pairwise
-comparison of all sub-blocks to determine which two are the most
-similar, using (0).
-
-\item We combine the two most-similar sub-blocks, and repeat the process
-starting at (2) until we are satisfied with the storage cost.
-
-\item The result is our partitioned block.
-
-\end{itemize}
+Using this function, the algorithm iteratively combines the two partitions that
+are closest together. Algorithm~\ref{alg:closestBF} shows the pseudocode for a brute-force approach
+  to finding the closest paris that runs in $O(n^2)$ time. However, there is
+  also a divide-and-conquer algorithm~\cite{cormen01} for solving closest pair
+  of points problem in $O(n \log n)$.
+After each combination of partitions, the algorithm calculate the storage cost
+for the partitioning. The algorithm stops when the  storage cost is below some
+specified threshold.  The result is the block partitioning.
 
 \begin{algorithm}[h]
+\SetKwFunction{closestPair}{closestPair}
 \scriptsize
@@
 
-\While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq \text{max}$}{
-    $min = \infty$ \; 
-    $x = \emptyset$ \;
-    $y = \emptyset$ \;
-    \For{$b_1 \in \mathcal{P}(B)$}{ 
-        \For{$b_2 \in \mathcal{P}(B)$}{ 
-            $\delta = (b_1 \cup b_2) - (b_1 \cap b_2)$ \; 
-           \If{$min < \delta $}{
-                 $\text{x} \leftarrow  b_1$\;
-                 $\text{y} \leftarrow b_2$\;
-                 $min \leftarrow \delta$
-            } 
-        }
-    }
+\While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
+  \text{max}$}{
+$\text{x}, \text{y}  \leftarrow  \closestPair{}  $\;
  $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - x - y$ \;
@@
 
-\paragraph{Second proposal}
+ \begin{algorithm}[h]
+\SetKwFunction{length}{length}
+\SetKwFunction{dist}{dist}
+ \scriptsize
+ \caption{Brute-force algorithm for closest points in $O(n^2)$.}
+ \label{alg:closestBF}
 
-\begin{itemize}
+$minDist  \gets \infty $ \;
+\For{$i \gets 1$ \textbf{to} $\length(P)-1$} {
+\For{$j \gets i+ 1$ \textbf{to} $\length(P)$} {
 
-\item The first partitioning is based on what queries we have seen.
-Every query gets its own sub-block. This is the "ideal" situation,
-because the I/O cost would be minimized for every query that we would
-have seen.
+  $p \gets P[i]$ \;
+  $q \gets P[j]$ \;
+  \If{$\dist(p,q) < minDist$} {
+    $minDist \gets \dist(p,q)$\;
+    $closestPair \gets (p,q)$\;
+  }
+}
+}
+\Return $closestPair$
 
-\item We keep a mapping of queries to blocks and frequencies.
+% \KwData{$B$: block, $Q$: set of queries}
 
-\item If the storage cost is above some threshold, then we get the queries with
-  the lowest frequencies, and combine their blocks. We also add the frequencies,
-  so we are combining the blocks that are least likely to be accessed.
+% $\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}
 
-\item We repeat until we are below the storage threshold.
+% \For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
+%  $\mathcal{P}(B) = \mathcal{P}(B) \cup q.A$ 
+% }
 
-\item The result is our partitioned block.
+% \While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
+%   \text{max}$}{
+% $\text{x}, \text{y}  \leftarrow  \closestPair{}  $\;
+%  $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - x - y$ \;
+%   $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  + x \cup y$ \;
 
-\end{itemize}
-
-\begin{algorithm}[h]
-\scriptsize
-\caption{Alternative algorithm for partitioning blocks into sub-blocks with overlapping attributes.}
-\label{alg:overlappingP2}
-\SetKwFunction{Enqueue}{Enqueue}
-\SetKwFunction{Dequeue}{Dequeue}
-
-\KwData{$B$: block, $Q$: set of queries}
-$\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}
-$pq \leftarrow \emptyset;$  \tcp*{priority queue;}
-
-\For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
- $\mathcal{P}(B) = \mathcal{P}(B) \cup q.A$ \;
- \Enqueue{$pq,\mathcal{P}(B), f(\mathcal{P}(B))$}\tcp*{put blocks in the queue;}
-}
-\While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
-  \text{max}$}{  
-  $x \leftarrow $ \Dequeue{$pq$}\;
-$y \leftarrow $ \Dequeue{$pq$}\tcp*{remove the two minimum;}
-  \Enqueue{$pq,x \cup y, f(x) + f(y)$}\tcp*{combine and insert;}    
-    }
-
-\For(\tcp*[f]{The elements in the queue are our partition}){$p \in pq$}{ 
-  $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) \cup p$ \;
-}
-\Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
+% \Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
  \end{algorithm}

### 2014-06-24 docs/graph-adapt/main.tex
non-overlapping alg

@@
 
-\author{Bu\u{g}ra Gedik         \and
-        Robert Soul\'{e}%etc.
+\author{Robert Soul\'{e}         \and
+        Bu\u{g}ra Gedik%etc.
 }
@@
 
-\paragraph*{Approach.}
+%$\mathcal{P}(B) \leftarrow \emptyset$ \tcp*{Initial partitioning}
 
-Idea1: Start with each attribute on its own, but as a set. At each step,
-consider  merging two existing sets, but pick the pair that results in the 
-maximum reduction in query cost per overhead removed.
+\begin{algorithm}[h]
+\scriptsize
+\caption{Algorithm for partitioning blocks into sub-blocks with non-overlapping attributes.}
+\label{alg:overlappingP}
+\KwData{$B$: block, $Q$: set of queries}
+$c^*\leftarrow \infty$ \tcp*{Lowest cost over all \# of partitions}
+\For(\tcp*[f]{For each possible \# of partitions}){$k=1$ to $|A|$}{
+   $R[i]\leftarrow \emptyset, \forall i\in [1..k]$ \tcp*{Initialize partitions}
+   \For(\tcp*[f]{For each attribute}){$a \in A$\textnormal{, in decr.\/ order of }$f(a)$}{
+      $c\leftarrow \infty$ \tcp*{Lowest cost over all assignments}  
+      $j\leftarrow -1$ \tcp*{Best partition assignment}
+      \For(\tcp*[f]{For each partition assignment}){$i\in [1..k]$} {
+         $R[i]\leftarrow R[i] \cup \{a\}$\tcp*{Assign attribute}
+         \If(\tcp*[f]{If size is not feasable}){$H(R, B, Q)\geq\alpha$}{
+            \textbf{continue}
+         }
+         \If(\tcp*[f]{If query cost is lower}){$L(R, B, Q)<c$}{
+            $c\leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
+            $j\leftarrow i$\tcp*{Update the best partition}
+         }
+         $R[i]\leftarrow R[i] \setminus a$\tcp*{Un-assign attribute}
+      }
+      $R[j]\leftarrow R[j] \cup a$\tcp*{Assign to best partition}
+   }
+   \If(\tcp*[f]{If query cost is lower}){$L(R, B, Q)<c^*$}{
+      $c^* \leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
+      $\mathcal{P}(B)\leftarrow R$\tcp*{Update the best partitioning}
+   }
+}
+\Return $\mathcal{P}(B)$ \tcp*{Final set of sub-blocks}
+\end{algorithm}
 

### 2014-06-26 docs/graph-adapt/main.tex
minor touches

@@
          }
-         $R[i]\leftarrow R[i] \setminus a$\tcp*{Un-assign attribute}
+         $R[i]\leftarrow R[i] \setminus \{a\}$\tcp*{Un-assign attribute}
       }
-      $R[j]\leftarrow R[j] \cup a$\tcp*{Assign to best partition}
+      $R[j]\leftarrow R[j] \cup \{a\}$\tcp*{Assign to best partition}
    }
-   \If(\tcp*[f]{If query cost is lower}){$L(R, B, Q)<c^*$}{
+   \If(\tcp*[f]{Lower cost}){$||R||=|A| \wedge L(R, B, Q)<c^*$}{
       $c^* \leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}

### 2014-07-02 docs/graph-adapt/main.tex
updated set notation
simulation setup

@@
    }
-   \If(\tcp*[f]{Lower cost}){$||R||=|A| \wedge L(R, B, Q)<c^*$}{
+   \If(\tcp*[f]{Has lower cost}){$||R||=|A| \wedge L(R, B, Q)<c^*$}{
       $c^* \leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}

### 2014-07-02 code/simulation/include/intergdb/simulation/ExperimentalData.h

@@ -1,0 +1,45 @@
+#ifndef INTERGDB_SIMULATION_EXPERIMENTALDATA_H
+#define INTERGDB_SIMULATION_EXPERIMENTALDATA_H
+
+#include <string>
+#include <fstream>
+#include <vector>
+#include <unordered_map>
+#include <boost/variant.hpp>
+
+namespace intergdb { namespace simulation {
+
+class ExperimentalData 
+{
+public:
+  typedef boost::variant<size_t, double, std::string> any_value; 
+public:
+  ExperimentalData(std::string const & name, 
+         std::string const & dirName = "expData");
+  void setDescription(std::string const & desc);
+  void setKeepValues(bool keep);
+  void addField(std::string const & field);
+  void open();
+  void addRecord();
+  void setFieldValue(int index, any_value const & value);
+  void setFieldValue(std::string const & name, any_value const & value);
+  void appendFieldValue(any_value const & value);
+  void close();
+  std::vector<std::vector<any_value> > const & getValues() const;
+private:
+  bool keepValues_;
+  size_t currentFieldIndex_;
+  std::string description_;
+  std::string name_;
+  std::string dirName_;
+  std::string const ext_;
+  std::string fileName_;
+  std::fstream file_;
+  std::vector<std::string> fieldNames_;
+  std::unordered_map<std::string, size_t> fieldNameIndices_;
+  std::vector<std::vector<any_value> > values_;
+};
+
+}}
+
+#endif /* INTERGDB_SIMULATION_EXPERIMENTALDATA_H */

### 2014-07-02 code/simulation/include/intergdb/simulation/Experiments.h

@@ -1,0 +1,17 @@
+#ifndef INTERGDB_SIMULATION_EXPERIMENTS_H
+#define INTERGDB_SIMULATION_EXPERIMENTS_H
+
+#include <intergdb/simulation/ExperimentalRun.h>
+
+namespace intergdb { namespace simulation
+{
+  class SampleExperiment : public ExperimentalRun 
+  {
+    void process() override;
+  };
+
+  // List other experiments here
+
+} } /* namespace */
+
+#endif /* INTERGDB_SIMULATION_EXPERIMENTS_H */

### 2014-07-02 code/simulation/src/intergdb/simulation/ExperimentalRun.cpp

@@ -1,0 +1,47 @@
+#include <intergdb/simulation/ExperimentalRun.h>
+
+#include <intergdb/simulation/AutoTimer.h>
+
+#include <boost/lexical_cast.hpp>
+
+#include <cxxabi.h>
+#include <typeinfo>
+#include <iostream>
+
+using namespace std;
+using namespace intergdb::simulation;
+
+string ExperimentalRun::getFullyQualifiedClassName() 
+{
+  int status;
+  char * realname = abi::__cxa_demangle(typeid(*this).name(), 0, 0, &status);
+    string cname = realname;
+  free(realname);
+  return cname;
+}
+
+string ExperimentalRun::getClassName() 
+{
+  string fqname = getFullyQualifiedClassName();
+  size_t loc = fqname.find_last_of("::");
+  if (loc==string::npos)
+    return fqname;
+  return fqname.substr(loc+1);
+}
+
+void ExperimentalRun::run()
+{
+  cout << "Experiment '" + getClassName() + "' started..." << endl;
+  AutoTimer timer;
+  timer.start();
+  try {
+    process();
+  } catch (exception const & e) {
+    cerr << "Exception caught, desc: " << e.what() << endl;
+  } catch (...) {
+    cerr << "Exception caught, desc: <unknown>" << endl;
+  }
+  timer.stop();
+  cout << "Experiment '" + getClassName() + "' ended (took " + 
+    boost::lexical_cast<string>(timer.getRealTimeInSeconds()) + " seconds)." << endl;
+}

### 2014-07-02 code/simulation/src/intergdb/simulation/Main.cpp

@@ -1,0 +1,14 @@
+#include <intergdb/simulation/Experiments.h>
+
+#include <cstdlib>
+
+using namespace std;
+using namespace intergdb;
+using namespace intergdb::simulation;
+
+int main()
+{
+    (SampleExperiment()).run();
+    return EXIT_SUCCESS;
+}
+

### 2014-07-02 code/simulation/src/intergdb/simulation/SampleExperiment.cpp

@@ -1,0 +1,33 @@
+#include <intergdb/simulation/Experiments.h>
+#include <intergdb/simulation/ExperimentalData.h>
+
+#include <random>
+#include <iostream>
+
+using namespace std;
+using namespace intergdb;
+using namespace intergdb::simulation;
+
+void SampleExperiment::process() 
+{
+  cerr << "This is a sample experiment with name: " 
+    << this->getClassName() << endl;
+
+  random_device rdev;
+  mt19937 rgen(rdev());
+  uniform_real_distribution<> udist(0, 10);
+
+  ExperimentalData exp("sample_exp");
+  exp.setDescription("This is a sample experiment");
+  exp.addField("block_size");
+  exp.addField("locality");
+  exp.setKeepValues(false);
+  exp.open();
+  for (size_t block_size : {1024, 2048, 4096}) {
+    exp.addRecord();
+    exp.setFieldValue("block_size", block_size);
+    double locality = udist(rgen);
+    exp.setFieldValue("locality", locality);
+  }
+  exp.close();
+};

### 2014-07-02 code/simulation/include/intergdb/simulation/AutoTimer.h
simulation setup

@@ -1,0 +1,25 @@
+
+#ifndef INTERGDB_SIMULATION_AUTOTIMER_H
+#define INTERGDB_SIMULATION_AUTOTIMER_H
+
+#include <boost/chrono.hpp>
+
+namespace intergdb { namespace simulation
+{
+    class AutoTimer
+    {
+    public:
+        AutoTimer();
+        AutoTimer(bool start);
+        void start();
+        void stop();
+        double getCPUTimeInSeconds();
+        double getRealTimeInSeconds();
+        double getCPUUtilization();
+    private:
+        boost::chrono::process_cpu_clock::time_point time_;
+        boost::chrono::process_cpu_clock::time_point end_;
+    };
+}}
+
+#endif /* INTERGDB_SIMULATION_AUTOTIMER_H */

### 2014-07-02 code/simulation/include/intergdb/simulation/ExperimentalRun.h

@@ -1,0 +1,20 @@
+#ifndef INTERGDB_SIMULATION_EXPERIMENTALRUN_H
+#define INTERGDB_SIMULATION_EXPERIMENTALRUN_H
+
+#include <string>
+
+namespace intergdb { namespace simulation
+{
+    class ExperimentalRun {
+    public:
+        virtual ~ExperimentalRun() {}
+        virtual void process() = 0;
+        void run();
+    protected:
+        std::string getClassName();
+        std::string getFullyQualifiedClassName();
+    };
+
+} } /* namespace */
+
+#endif /* INTERGDB_SIMULATION_EXPERIMENTALRUN_H */

### 2014-07-02 code/simulation/src/intergdb/simulation/AutoTimer.cpp

@@ -1,0 +1,51 @@
+#include <intergdb/simulation/AutoTimer.h>
+
+using namespace intergdb::simulation;
+using namespace boost::chrono;
+
+AutoTimer::AutoTimer()
+{}
+
+AutoTimer::AutoTimer(bool start)
+{
+  if (start) 
+    time_ = process_cpu_clock::now(); 
+}
+
+void AutoTimer::start()
+{
+  time_ = process_cpu_clock::now(); 
+}
+
+void AutoTimer::stop()
+{
+  end_ = process_cpu_clock::now();  
+}
+
+double AutoTimer::getCPUTimeInSeconds()
+{
+  process_cpu_clock::times timeTaken = (end_ - time_).count();
+  process_user_cpu_clock::duration userTimeTaken(timeTaken.user);
+  process_system_cpu_clock::duration systemTimeTaken(timeTaken.system);
+  return duration<double>(userTimeTaken).count() + 
+         duration<double>(systemTimeTaken).count();
+}
+
+double AutoTimer::getRealTimeInSeconds()
+{
+  process_cpu_clock::times timeTaken = (end_ - time_).count();
+  process_real_cpu_clock::duration realTimeTaken(timeTaken.real);
+  return duration<double>(realTimeTaken).count();
+}
+
+double AutoTimer::getCPUUtilization()
+{
+  process_cpu_clock::times timeTaken = (end_ - time_).count();
+  process_user_cpu_clock::duration userTimeTaken(timeTaken.user);
+  process_system_cpu_clock::duration systemTimeTaken(timeTaken.system);
+  process_real_cpu_clock::duration realTimeTaken(timeTaken.real);
+  return (duration<double>(userTimeTaken).count() + 
+      duration<double>(systemTimeTaken).count()) / 
+         duration<double>(realTimeTaken).count();
+}
+

### 2014-07-02 code/simulation/src/intergdb/simulation/ExperimentalData.cpp

@@ -1,0 +1,90 @@
+#include <intergdb/simulation/ExperimentalData.h>
+
+#include <boost/filesystem.hpp>
+#include <stdexcept>
+
+using namespace std;
+using namespace intergdb;
+using namespace intergdb::simulation;
+
+ExperimentalData::ExperimentalData(string const & expName, 
+  string const & dirName /*="."*/)
+  : keepValues_(false),
+    currentFieldIndex_(0),
+    name_ (expName),
+    dirName_ (dirName),
+    ext_(".dat"),
+    fileName_(name_+ext_)
+{} 
+
+void ExperimentalData::setDescription(std::string const & desc)
+{
+  description_ = desc;
+}
+
+void ExperimentalData::setKeepValues(bool keep) 
+{
+  keepValues_ = keep;
+}
+
+void ExperimentalData::addField(string const & field)
+{
+  fieldNameIndices_.insert(make_pair(field, fieldNames_.size()));
+    fieldNames_.push_back(field);
+}
+
+void ExperimentalData::open()
+{
+  using namespace boost::filesystem;
+  path dir(dirName_);
+  create_directory(dir);
+  file_.open((dirName_+"/"+fileName_).c_str(), ios::out);
+  if (!description_.empty())
+    file_ << "# " << description_ << endl;
+  file_ << "# ";
+  for (size_t i=0, iu=fieldNames_.size(); i<iu; ++i)
+    file_ << fieldNames_[i] << "\t";
+  file_.flush();
+}
+
+void ExperimentalData::addRecord()
+{
+  file_ << endl;
+  if (keepValues_) {
+    currentFieldIndex_ = 0;
+    values_.push_back(vector<any_value>(fieldNames_.size(), any_value()));
+  }
+}
+
+void ExperimentalData::setFieldValue(int fieldIndex, any_value const & value)
+{
+  file_ << value << "\t";
+  file_.flush();
+  if (keepValues_) 
+    values_.back()[static_cast<size_t>(fieldIndex)] = value;
+}
+
+void ExperimentalData::setFieldValue(std::string const & fieldName, any_value const & value)
+{
+  if (fieldNameIndices_.count(fieldName)==0) 
+    throw runtime_error("field '"+fieldName+"' not found!");
+  size_t index = fieldNameIndices_[fieldName];
+  setFieldValue(index, value);
+}
+
+void ExperimentalData::appendFieldValue(any_value const & value)
+{
[…]

### 2014-07-12 docs/graph-adapt/main.tex
added lp notes to paper

@@
 $\text{x}, \text{y}  \leftarrow  \closestPair{}  $\;
- $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - x - y$ \;
-  $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  + x \cup y$ \;
+ $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - \{x, y\}$ \;
+  $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  \cup \{ x \cup y \}$ \;
 }
@@
 
+\subsection{Optimization}
+
+\begin{enumerate}
+
+   \item If an attribute is in a query:
+\begin{alltt}
+    v_a1_q1 = 1
+    v_a2_q1 = 0
+    v_a1_q2 = 0
+    v_a2_q2 = 1
+\end{alltt}
+
+   \item If an attribute is in a partition:
+\begin{alltt}
+    v_a1_p1 = ?
+    v_a2_p1 = ?
+    v_a1_p2 = ?
+    v_a2_p2 = ?
+\end{alltt}
+
+ \item If an partition is used by a query
+
+\begin{alltt}
+    v_p1_q1 = ?
+    v_p1_q2 = ?
+    v_p2_q1 = ?
+    v_p2_q2 = ?
+\end{alltt}
+
+   \item If an attribute and partition are used by a query
+\begin{alltt} 
+   v_a1_p1_q1 = ?
+    v_a1_p1_q2 = ?
+    v_a2_p2_q1 = ?
+    v_a2_p2_q2 = ?
+\end{alltt}
+
+    \item add constraint that all attributes must appear in a partition
+\begin{alltt}
+    v_a1_p1 + v_a1_p2  = 1
+    v_a2_p1 + v_a2_p2 = 1
+\end{alltt}
+
+    \item the number of attributes use by query from a partition determines the
+      value of \texttt{v\_p1\_q1} (either 0 or 1)
+\begin{alltt}
+    (v_a1_q1 * v_a1_p1 + v_a2_q1 
+   * v_a2_p1) - v_p1_q1 >= 0
+\end{alltt}
+
+    \item To ensure that it is 0 or 1, we are going to add to the objective function, where K is some big value:
+\begin{alltt}
+   - K * v_p1_q1
+   - K * v_p1_q2
+   - K * v_p2_q1 
+   - K * v_p2_q2
+\end{alltt}
+
+    \item add constraint to make sure that \texttt{v\_a1\_p1} and \texttt{v\_p1\_q1} determine
+      \texttt{v\_a1\_p1\_q1}
+\begin{alltt}
+    v_a1_p1_q1 >= (v_a1_p1 + v_p1_q1) -1
+\end{alltt}
+
+    \item objective function : minimize the following, where the ?'s are constants
+
+   \item Here is what we want:
+\begin{alltt}
+   (? * v_a1_p1_q1 + ? * v_a2_p1_q1 + ?)
+  + (? * v_a1_p1_q2 + ? * v_a2_p1_q2 + ?)
+  + (? * v_a1_p2_q1 + ? * v_a2_p2_q1 + ?)
+  + (? * v_a1_p2_q2 + ? * v_a2_p2_q2 + ?)
+\end{alltt}
+\end{enumerate}
+
 %\begin{acknowledgements}

### 2014-07-12 docs/graph-adapt/main.tex
mixed integer linear program

--- 2014-07-02 docs/graph-adapt/main.tex
@@
 \begin{equation}
-s(B) = c_e(B) \cdot \left(16 + \sum_{a\in A} s(a)\right) + c_n(B) \cdot 12  
+s(B) = c_e(B) \cdot \left(16 + \sum_{a\in B.A} s(a)\right) + c_n(B) \cdot 12  
 \end{equation}
@@
 
+\subsection{Mixed Integer Linear Program Formulation}
+$x_{a,p}$: $1$ if attribute $a$ is in partition $p$, $0$ otherwise.\\
+$y_{p,q}$: $1$ if partition $p$ is used by query $q$, $0$ otherwise.\\
+$z_{a,p,q}$: $1$ if partition $p$ is used by query $q$, and attribute $a$ is in partition $p$, $0$ otherwise.\\
+$u_{p}$: $1$ if partition $p$ is assigned at least $1$ attribute, $0$ otherwise.\\
+The formulation for ($k$ partitions):
+
+% NOTE: In the simulation we will have query weights that will factor into the objective function
+\begin{eqnarray}
+\text{minimize} 
+    \Big(\sum_{q\in Q} \sum_{p=1}^{k} (16\cdot c_e(B) &+& 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
+    + \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q}\Big) &-& \Big(K \sum_{p=1}^{k} u_p + \sum_{q\in Q} y_{p,q}\Big) \nonumber\\
+\text{subject to}\hspace{2.75cm}&&\nonumber\\
+\forall_{a\in A}, 
+    && \sum_{p=1}^{k} x_{a,p} = 1\nonumber\\
+\forall_{\{p,q\}\in [1..k]\times Q}, 
+    &&  \sum_{a\in A} q(a)\cdot x_{a,p} - y_{p,q} \geq 0 \nonumber\\
+\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
+  && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\nonumber\\
+\forall_{p\in[1..k]},
+   && \sum_{a\in A} y_{a,p} - u_p \geq 0 \nonumber\\
+\Big(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\Big) && \sum_{p=1}^{k} u_p < \alpha \nonumber
+\end{eqnarray}
+
 \subsection{Overlapping Attributes}

### 2014-07-16 docs/graph-adapt/main.tex
mixed integer linear program updates
Removed my ill-formatted notes
made a minor change to the overhead formuation
changed how overhead constaints are handled

--- 2014-07-12 docs/graph-adapt/main.tex
@@
 \begin{equation}
-H(\mathcal{P}, B) = |\mathcal{P}(B)|\cdot\left(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\right) 
+H(\mathcal{P}, B) = (|\mathcal{P}(B)|-1)\cdot\left(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\right) 
 \end{equation}
@@
 $z_{a,p,q}$: $1$ if partition $p$ is used by query $q$, and attribute $a$ is in partition $p$, $0$ otherwise.\\
-$u_{p}$: $1$ if partition $p$ is assigned at least $1$ attribute, $0$ otherwise.\\
-The formulation for ($k$ partitions):
+
+Let $q(a)\equiv \mathbf{1}(a \in q.A)$.
 
@@
     \Big(\sum_{q\in Q} \sum_{p=1}^{k} (16\cdot c_e(B) &+& 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
-    + \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q}\Big) &-& \Big(K \sum_{p=1}^{k} u_p + \sum_{q\in Q} y_{p,q}\Big) \nonumber\\
+    + \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q}\Big) &-& \Big(K\cdot \sum_{p=1}^{k} u_p + \sum_{q\in Q} y_{p,q}\Big) \nonumber\\
 \text{subject to}\hspace{2.75cm}&&\nonumber\\
@@
   && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\nonumber\\
-\forall_{p\in[1..k]},
-   && \sum_{a\in A} y_{a,p} - u_p \geq 0 \nonumber\\
-\Big(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\Big) && \sum_{p=1}^{k} u_p < \alpha \nonumber
 \end{eqnarray}
+
+We will start with $k=|A|$ and check if the number of partitions used results
+on the overhead constraint being satisified, that is:
+$$\Big(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\Big) \cdot \Big(\sum_{p=1}^{k} u_p -1\Big) < \alpha$$
+
+If the constraint is satisfied, then we are done. Otherwise, we set
+$k=\sum_{p=1}^{k} u_p -1$ and repeat.
 

### 2014-07-16 docs/graph-adapt/main.tex
remove the u's from the objective as well
fixed the paper witht the new formulation
added back u's and overhead constaints with the new formulation

@@
 $z_{a,p,q}$: $1$ if partition $p$ is used by query $q$, and attribute $a$ is in partition $p$, $0$ otherwise.\\
+$u_{p}$: $1$ if partition $p$ is assigned at least $1$ attribute, $0$ otherwise.\\
 
@@
 \text{minimize}  
-    \Big(\sum_{q\in Q} \sum_{p=1}^{k} (16\cdot c_e(B) &+& 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
-    + \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q}\Big) &-& \Big(K\cdot \sum_{p=1}^{k} u_p + \sum_{q\in Q} y_{p,q}\Big) \nonumber\\
-\text{subject to}\hspace{2.75cm}&&\nonumber\\
+    \sum_{q\in Q} \sum_{p=1}^{k}(16\cdot c_e(B) &+& 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
+    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
+\text{subject to}&&\nonumber\\
 \forall_{a\in A}, 
@@
     &&  \sum_{a\in A} q(a)\cdot x_{a,p} - y_{p,q} \geq 0 \nonumber\\
+\forall_{\{p,q\}\in [1..k]\times Q}, 
+    &&  K\cdot y_{p,q} - \sum_{a\in A} q(a)\cdot x_{a,p}  \geq 0 \nonumber\\
 \forall_{\{a,p,q\}\in A\times [1..k]\times Q},
     && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\nonumber\\
+\forall_{p\in[1..k]},
+    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
+\forall_{p\in[1..k]},
+    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
+\Big(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\Big) &\cdot& \Big(\sum_{p=1}^{k} u_p -1\Big) \leq \alpha \nonumber
 \end{eqnarray}
-
-We will start with $k=|A|$ and check if the number of partitions used results
-on the overhead constraint being satisified, that is:
-$$\Big(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\Big) \cdot \Big(\sum_{p=1}^{k} u_p -1\Big) < \alpha$$
-
-If the constraint is satisfied, then we are done. Otherwise, we set
-$k=\sum_{p=1}^{k} u_p -1$ and repeat.
 

### 2014-07-19 docs/graph-adapt/main.tex
added ILP construction we use for expressing y and u constraints

@@
 \text{minimize}  
-    \sum_{q\in Q} \sum_{p=1}^{k}(16\cdot c_e(B) &+& 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
+    \sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
     &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
@@
     && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
-\Big(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\Big) &\cdot& \Big(\sum_{p=1}^{k} u_p -1\Big) \leq \alpha \nonumber
+ && \sum_{p=1}^{k} u_p -1 \leq \frac{\alpha}{
+1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}
+ } \nonumber
 \end{eqnarray}
+
+Here we use the following LP construction. Assume we have two variables,
+$\beta_1$ and $\beta_2$, where $\beta_2\in[0,1]$ and $\beta_1\geq 0$. We want to
+implement the following constraint: 
+$$
+\beta_2 = 
+\begin{cases}
+0 & \beta_1 = 0 \\
+1 & \beta_1 \geq 0 
+\end{cases}
+$$
+
+This could be expressed as a linear constraint as follows,  where $K$ is a
+large constant guaranteed to be larger  that $\beta_1$ for practical purposes:
+\begin{eqnarray}
+\hspace{2cm}&& \beta_1 - \beta_2 \geq 0\nonumber\\
+\hspace{2cm}&& K\cdot\beta_2 - \beta_1 \geq 0\nonumber
+\end{eqnarray}
+
+In our ILP formulation, we use this construction twice. The first use is shown
+in the second and third constrants, where $\beta_1=\sum_{a\in A} q(a)\cdot
+x_{a,p}$ and $\beta_2=y_{p,q}$. The second use is shown in the fifth and 
+sixth constaints, where $\beta_1=\sum_{a\in A} x_{a,p}$ and $\beta_2=u_p$.
 

### 2014-07-20 docs/paper-old/main.tex
moved paper to a more moduler format

--- 2014-07-19 docs/graph-adapt/main.tex
@@
 
-\institute{B. Gedik \at            
-              \email{bgedik@cs.bilkent.edu.tr}           %  \\
+\institute{B. Gedik \at \email{bgedik@cs.bilkent.edu.tr} 
            \and
-           R. Soul\'{e} \at
-               \email{robert.soule@usi.ch}     
+           R. Soul\'{e} \at \email{robert.soule@usi.ch}     
 }
@@
          $R[i]\leftarrow R[i] \cup \{a\}$\tcp*{Assign attribute}
-         \If(\tcp*[f]{If size is not feasable}){$H(R, B, Q)\geq\alpha$}{
-            \textbf{continue}
-         }
          \If(\tcp*[f]{If query cost is lower}){$L(R, B, Q)<c$}{
@@
    }
-   \If(\tcp*[f]{Has lower cost}){$||R||=|A| \wedge L(R, B, Q)<c^*$}{
+   \tcp{If solution is feasable and has lower cost}
+   \If{$H(R, B, Q)\geq\alpha \wedge L(R, B, Q)<c^*$}{
       $c^* \leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}

### 2014-07-20 docs/paper/text/solution.tex
moved paper to a more moduler format

--- 2014-07-20 docs/paper-old/main.tex
@@
-\documentclass[twocolumn]{svjour3}          % twocolumn
-%
-\smartqed  % flush right qed marks, e.g. at end of proof
-%
-
-\usepackage{alltt}
-\usepackage{enumerate}
-\usepackage[ruled,vlined]{algorithm2e}
-\usepackage{xargs}
-\usepackage{amssymb}
-\usepackage{amsmath}
-
-\renewcommand{\CommentSty}[1]{\textnormal{#1}}
-\DontPrintSemicolon
-\SetKwComment{tcp}{$\triangleright$ }{}
-\SetVlineSkip{0cm}
-\SetAlgoSkip{}
-
-\let\oldemptyset\emptyset
-\let\emptyset\varnothing
-
-\DeclareMathOperator{\dist}{dist}
-
-% Insert the name of "your journal" with
-\journalname{VLDBJ}
-%
-\begin{document}
-
-\title{Adaptive Disk Storage for Interaction Graphs}
-
-%\titlerunning{Short form of title}        % if too long for running head
-
-\author{Robert Soul\'{e}         \and
-        Bu\u{g}ra Gedik%etc.
-}
-
-%\authorrunning{Short form of author list} % if too long for running head
-
-\institute{B. Gedik \at \email{bgedik@cs.bilkent.edu.tr} 
-           \and
-           R. Soul\'{e} \at \email{robert.soule@usi.ch}     
-}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
-
-\begin{abstract}
-Insert your abstract here. Include keywords, PACS and mathematical
-subject classification numbers as needed.
-\keywords{First keyword \and Second keyword \and More}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
-\end{abstract}
-
-\section{Introduction}
-An \emph{interaction graph} is an append-only graph, where new edges and
-vertices are added as time progresses.%
-
-Typical interaction graphs have data associated with their vertices and edges.
-We call this data \emph{attributes} or \emph{properties}. Most algorithms
-access this data as they traverse the graph. However, not all of the data is
-accessed by all the queries. Typically, there are correlations among the
-attributes accessed by different queries, such as Q1 and Q5 accessing
-attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and so on.%
-
-In an interaction graph database organized as blocks of temporal neighborlists
-(as in Gedik et al.~\cite{gedik14}), it is important to store the edge and
-vertex properties locally. For  instance, if the edge properties are kept away
-in a relational table, there will be almost no benefit to the locality
-optimizations performed for block organization (discussed in Gedik et
-al.~\cite{gedik14}, as we would have to go back and forth between the disk
-blocks to access the edge attributes).%
-
-On the other hand, putting all the edge attributes into the disk blocks
-containing the graph structure is expected to cause significant overhead when
-only a few of these attributes are read. This is somewhat similar to the 
-problem with row-oriented databases, where the entire row needs to be accessed
-from the disk despite the query needing only a small fraction of it.
-Unfortunately, there is no clear correspondence to a column-oriented database
-layout for the graph databases.%
-
-An important characteristic of interaction graphs are that they are temporal.
-As such, the co-access correlations for the attributes can be different for
-different temporal regions. It might also be unknown at the insertion time and
-may be discovered later depending on the workload. This points to the need for
-adaptively optimizing the layout (somewhat similar to H2O \cite{alagiannis14}
-and HYRISE~\cite{grund10}).%
-
-I imagine a solution to this problem, which I name the `rail layout'. The idea
-is to start with large blocks that contain the entire data. As we learn about
-the access properties for different time regions, we might start splitting
-such blocks into smaller blocks that run parallel to each other, almost like
-having two or more graph databases for certain time regions, each containing a
- different subset of the attributes, but with a link between them in case a
-query needs to access both.
-
-\section{Problem Description}
-
-The problem is to determine a partitioning of the disk blocks into sub-blocks
-that would minimize query I/O. Of course, partitioning into sub-blocks results
-in an increase in storage cost. The storage cost must stay below some
-specified limit.
-
-There are two possibilities: \emph{overlapping} and \emph{non-overlapping}
-partitioning. In the overlapping case, attributes can be replicated in the
-sub-blocks. The non-overlapping case is a true partition.
-
-We start by defining the cost model, which formalizes the query I/O and storage
-costs.
-
-\section{Cost Model}
-Let $Q$ be the query workload, where each query $q\in Q$ accesses a set of
-attributes $q.A$ and traverses parts of the graph for the time range
-$q.T=[q.t_s,q.t_e]$. We denote the set of all attributes as $A$. Given a block
-$B$, we denote its time range as $B.T$, which is the union of the time ranges
-of its temporal neighborlists. Let $s(a)$ denote the size of an attribute $a$.
-We use $c_n(B)$ to denote the number of temporal neighborlists within block
-$B$ and $c_e(B)$ to denote the total number of edges in the temporal
-neighborlists within the block. We overload the notation for block size and
-use $s(B)$ to denote the size of a block $B$. We have: 
-\begin{equation}
-s(B) = c_e(B) \cdot \left(16 + \sum_{a\in B.A} s(a)\right) + c_n(B) \cdot 12  
-\end{equation}
-Here, $16$ corresponds to the cost of storing the edge id and the timestamp,
-and $12$ corresponds to the cost of storing the head vertex ($8$ bytes) plus
-the number of entries ($4$ bytes) for a temporal neighborlist.
-
-Our goal is to create a potentially overlapping partitioning of a block,
-denoted by $\mathcal{P}(B)$. We call the partitions \emph{sub-blocks}. We have
-$\bigcup_{B'\in \mathcal{P}(B)} B'.A = A$. We aim to find the function
-$\mathcal{P}$ that minimizes the query I/O over $B$, while keeping the
-relative storage overhead beyond a limit, say $1+\alpha$ times the original.
-If we represent the query I/O as $L(\mathcal{P}, B)$ and the relative storage
-overhead as $H(\mathcal{P}, B)$, our goal is to find:
-\begin{equation}
-\mathcal{P} \leftarrow \mbox{argmin}_{\{\mathcal{P}: H(\mathcal{P}(B)) < \alpha\}} L(\mathcal{P},B)
-\end{equation}
-
-The storage overhead can be formalized as:
-\begin{equation}
-H(\mathcal{P}, B) = (|\mathcal{P}(B)|-1)\cdot\left(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\right) 
-\end{equation}
-
-\subsection{Query I/O}
-
-Let $m$ be a function that maps a query $q$ to the set of sub-blocks that are
-accessed to satisfy it for a relevant block $B$ under a given partitioning
-$\mathcal{P}$. For the case of non-overlapping attributes, we have:
-\begin{equation}
-m(\mathcal{P}, B, q) = \{B': B'\in \mathcal{P}(B) \wedge q.A \cap B'.A \ne \emptyset\}  
-\end{equation}
-
-For the case of overlapping attributes, we use a simple heuristic to define the
-set of sub-blocks to be used for answering the query.
-Algorithm~\ref{alg:greedyM} captures it. The idea is to select sub-blocks that
-bring the highest relative marginal gain, which is defined as the size of
-attribute data that contributes to the query result, relative to the sub-block
-size. While computing the marginal gain, attributes that are covered by
-sub-blocks that are already selected are not considered.
-%
-\begin{algorithm}[h]
-\scriptsize
-\caption{m-overlapping($\mathcal{P}, B, q$)}
-\label{alg:greedyM}
-\KwData{$\mathcal{P}$: partitioning function, $B$: block, $q$: query}
-$S\leftarrow \emptyset; R\leftarrow \emptyset$ \tcp*{Selected attributes; Resulting sub-blocks}
-\While(\tcp*[f]{While unselected attributes remain}){$S \subset q.A$}{
-  $B' \leftarrow \mbox{argmax}_{B'\in\mathcal{P}(B)\setminus R} \sum_{a\in B'.A \cap q.A \setminus S} \frac{c_e(B') \cdot s(a)}{s(B')}$
-  $S \leftarrow S \cup B'.A$\tcp*{Extend the selected attributes}
-  $R\leftarrow R \cup B'$\tcp*{Extend the selected sub-blocks}
-}
-\Return R \tcp*{Final set of sub-blocks covering the query attributes}
-\end{algorithm} 
-%
-The query I/O cost for a block is then given by:
-\begin{equation}
-L(\mathcal{P}, B) = \sum_{q\in Q} \mathbf{1}(q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} \!\!s(B')
-\end{equation}
-
-\paragraph{\textbf{Other useful equations}}
-
-\noindent The following are unused, but may be needed for the partitioning
-algorithm.
-
-Let $f: \mathbb{R}\times A \rightarrow \mathbb{N}$ be a function that maps a
-attribute-timestamp pair $(a, t)$ to the frequency of queries that access
-attribute $a$ for time $t$. More formally:
-\begin{equation}
-f(a,t) = \sum_{q\in Q} \textbf{1}(a\in q.A \wedge t\in q.T)
-\end{equation}
-
-Let $T$ represent a time range $[T.s, T.e]$. We denote the frequency of
-attribute access for the time interval $T$ for attribute $a$ as follows:
-\begin{equation}
-f(a, T) = \int_{T.s}^{T.e} f(a, t)\cdot dt =  \sum_{q\in Q} |q.T \cap T| 
-\end{equation}
-
 \section{Partitioning}
-
-TODO: This section is still a work-in-progress. We are trying to find the algorithms.
+This section is still a work-in-progress. We are trying to find the algorithms.
 
@@
 
-\begin{algorithm}[h]
+\begin{algorithm}[ht]
 \scriptsize
@@
 
-\begin{algorithm}[h]
+\begin{algorithm}[ht]
 \SetKwFunction{closestPair}{closestPair}
@@
 \caption{Algorithm for partitioning blocks into sub-blocks with overlapping attributes.}
-\label{alg:overlappingP}
+\label{alg:closestPair}
 \KwData{$B$: block, $Q$: set of queries}
@@
 
- \begin{algorithm}[h]
+ \begin{algorithm}[ht]
 \SetKwFunction{length}{length}
@@
  \label{alg:closestBF}
-
 $minDist  \gets \infty $ \;
@@
 \Return $closestPair$
-
 % \KwData{$B$: block, $Q$: set of queries}
-
 % $\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}
-
 % \For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
@@
 % }
-
 % \While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
@@
 %   $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  + x \cup y$ \;
-
 % \Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
  \end{algorithm}
-
-%\begin{acknowledgements}
-%If you'd like to thank anyone, place your comments here
-%and remove the percent signs.
-%\end{acknowledgements}
-
-\bibliographystyle{spmpsci}
-\bibliography{main}
-
-\end{document}
-% end of file template.tex

### 2014-07-20 docs/paper/text/introduction.tex
moved paper to a more moduler format

--- 2014-07-16 docs/graph-adapt/main.tex
@@
-\documentclass[twocolumn]{svjour3}          % twocolumn
-%
-\smartqed  % flush right qed marks, e.g. at end of proof
-%
-
-\usepackage{alltt}
-\usepackage{enumerate}
-\usepackage[ruled,vlined]{algorithm2e}
-\usepackage{xargs}
-\usepackage{amssymb}
-\usepackage{amsmath}
-
-\renewcommand{\CommentSty}[1]{\textnormal{#1}}
-\DontPrintSemicolon
-\SetKwComment{tcp}{$\triangleright$ }{}
-\SetVlineSkip{0cm}
-\SetAlgoSkip{}
-
-\let\oldemptyset\emptyset
-\let\emptyset\varnothing
-
-\DeclareMathOperator{\dist}{dist}
-
-% Insert the name of "your journal" with
-\journalname{VLDBJ}
-%
-\begin{document}
-
-\title{Adaptive Disk Storage for Interaction Graphs}
-
-%\titlerunning{Short form of title}        % if too long for running head
-
-\author{Robert Soul\'{e}         \and
-        Bu\u{g}ra Gedik%etc.
-}
-
-%\authorrunning{Short form of author list} % if too long for running head
-
-\institute{B. Gedik \at            
-              \email{bgedik@cs.bilkent.edu.tr}           %  \\
-           \and
-           R. Soul\'{e} \at
-               \email{robert.soule@usi.ch}     
-}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
-
-\begin{abstract}
-Insert your abstract here. Include keywords, PACS and mathematical
-subject classification numbers as needed.
-\keywords{First keyword \and Second keyword \and More}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
-\end{abstract}
-
 \section{Introduction}
@@
 query needs to access both.
-
-\section{Problem Description}
-
-The problem is to determine a partitioning of the disk blocks into sub-blocks
-that would minimize query I/O. Of course, partitioning into sub-blocks results
-in an increase in storage cost. The storage cost must stay below some
-specified limit.
-
-There are two possibilities: \emph{overlapping} and \emph{non-overlapping}
-partitioning. In the overlapping case, attributes can be replicated in the
-sub-blocks. The non-overlapping case is a true partition.
-
-We start by defining the cost model, which formalizes the query I/O and storage
-costs.
-
-\section{Cost Model}
-Let $Q$ be the query workload, where each query $q\in Q$ accesses a set of
-attributes $q.A$ and traverses parts of the graph for the time range
-$q.T=[q.t_s,q.t_e]$. We denote the set of all attributes as $A$. Given a block
-$B$, we denote its time range as $B.T$, which is the union of the time ranges
-of its temporal neighborlists. Let $s(a)$ denote the size of an attribute $a$.
-We use $c_n(B)$ to denote the number of temporal neighborlists within block
-$B$ and $c_e(B)$ to denote the total number of edges in the temporal
-neighborlists within the block. We overload the notation for block size and
-use $s(B)$ to denote the size of a block $B$. We have: 
-\begin{equation}
-s(B) = c_e(B) \cdot \left(16 + \sum_{a\in B.A} s(a)\right) + c_n(B) \cdot 12  
-\end{equation}
-Here, $16$ corresponds to the cost of storing the edge id and the timestamp,
-and $12$ corresponds to the cost of storing the head vertex ($8$ bytes) plus
-the number of entries ($4$ bytes) for a temporal neighborlist.
-
-Our goal is to create a potentially overlapping partitioning of a block,
-denoted by $\mathcal{P}(B)$. We call the partitions \emph{sub-blocks}. We have
-$\bigcup_{B'\in \mathcal{P}(B)} B'.A = A$. We aim to find the function
-$\mathcal{P}$ that minimizes the query I/O over $B$, while keeping the
-relative storage overhead beyond a limit, say $1+\alpha$ times the original.
-If we represent the query I/O as $L(\mathcal{P}, B)$ and the relative storage
-overhead as $H(\mathcal{P}, B)$, our goal is to find:
-\begin{equation}
-\mathcal{P} \leftarrow \mbox{argmin}_{\{\mathcal{P}: H(\mathcal{P}(B)) < \alpha\}} L(\mathcal{P},B)
-\end{equation}
-
-The storage overhead can be formalized as:
-\begin{equation}
-H(\mathcal{P}, B) = (|\mathcal{P}(B)|-1)\cdot\left(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\right) 
-\end{equation}
-
-\subsection{Query I/O}
-
-Let $m$ be a function that maps a query $q$ to the set of sub-blocks that are
-accessed to satisfy it for a relevant block $B$ under a given partitioning
-$\mathcal{P}$. For the case of non-overlapping attributes, we have:
-\begin{equation}
-m(\mathcal{P}, B, q) = \{B': B'\in \mathcal{P}(B) \wedge q.A \cap B'.A \ne \emptyset\}  
-\end{equation}
-
-For the case of overlapping attributes, we use a simple heuristic to define the
-set of sub-blocks to be used for answering the query.
-Algorithm~\ref{alg:greedyM} captures it. The idea is to select sub-blocks that
-bring the highest relative marginal gain, which is defined as the size of
-attribute data that contributes to the query result, relative to the sub-block
-size. While computing the marginal gain, attributes that are covered by
-sub-blocks that are already selected are not considered.
-%
-\begin{algorithm}[h]
-\scriptsize
-\caption{m-overlapping($\mathcal{P}, B, q$)}
-\label{alg:greedyM}
-\KwData{$\mathcal{P}$: partitioning function, $B$: block, $q$: query}
-$S\leftarrow \emptyset; R\leftarrow \emptyset$ \tcp*{Selected attributes; Resulting sub-blocks}
-\While(\tcp*[f]{While unselected attributes remain}){$S \subset q.A$}{
-  $B' \leftarrow \mbox{argmax}_{B'\in\mathcal{P}(B)\setminus R} \sum_{a\in B'.A \cap q.A \setminus S} \frac{c_e(B') \cdot s(a)}{s(B')}$
-  $S \leftarrow S \cup B'.A$\tcp*{Extend the selected attributes}
-  $R\leftarrow R \cup B'$\tcp*{Extend the selected sub-blocks}
-}
-\Return R \tcp*{Final set of sub-blocks covering the query attributes}
-\end{algorithm} 
-%
-The query I/O cost for a block is then given by:
-\begin{equation}
-L(\mathcal{P}, B) = \sum_{q\in Q} \mathbf{1}(q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} \!\!s(B')
-\end{equation}
-
-\paragraph{\textbf{Other useful equations}}
-
-\noindent The following are unused, but may be needed for the partitioning
-algorithm.
-
-Let $f: \mathbb{R}\times A \rightarrow \mathbb{N}$ be a function that maps a
-attribute-timestamp pair $(a, t)$ to the frequency of queries that access
-attribute $a$ for time $t$. More formally:
-\begin{equation}
-f(a,t) = \sum_{q\in Q} \textbf{1}(a\in q.A \wedge t\in q.T)
-\end{equation}
-
-Let $T$ represent a time range $[T.s, T.e]$. We denote the frequency of
-attribute access for the time interval $T$ for attribute $a$ as follows:
-\begin{equation}
-f(a, T) = \int_{T.s}^{T.e} f(a, t)\cdot dt =  \sum_{q\in Q} |q.T \cap T| 
-\end{equation}
-
-\section{Partitioning}
-
-TODO: This section is still a work-in-progress. We are trying to find the algorithms.
-
-\subsection{Non-Overlapping Attributes}
-
-\paragraph*{Problem.}\emph{Find a true partitioning of attributes that minimizes
-  the query I/O and bounds the storage cost by some upper limit.}
-
-%$\mathcal{P}(B) \leftarrow \emptyset$ \tcp*{Initial partitioning}
-
-\begin{algorithm}[h]
-\scriptsize
-\caption{Algorithm for partitioning blocks into sub-blocks with non-overlapping attributes.}
-\label{alg:overlappingP}
-\KwData{$B$: block, $Q$: set of queries}
-$c^*\leftarrow \infty$ \tcp*{Lowest cost over all \# of partitions}
-\For(\tcp*[f]{For each possible \# of partitions}){$k=1$ to $|A|$}{
-   $R[i]\leftarrow \emptyset, \forall i\in [1..k]$ \tcp*{Initialize partitions}
-   \For(\tcp*[f]{For each attribute}){$a \in A$\textnormal{, in decr.\/ order of }$f(a)$}{
-      $c\leftarrow \infty$ \tcp*{Lowest cost over all assignments}  
-      $j\leftarrow -1$ \tcp*{Best partition assignment}
-      \For(\tcp*[f]{For each partition assignment}){$i\in [1..k]$} {
-         $R[i]\leftarrow R[i] \cup \{a\}$\tcp*{Assign attribute}
-         \If(\tcp*[f]{If size is not feasable}){$H(R, B, Q)\geq\alpha$}{
-            \textbf{continue}
-         }
-         \If(\tcp*[f]{If query cost is lower}){$L(R, B, Q)<c$}{
-            $c\leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
-            $j\leftarrow i$\tcp*{Update the best partition}
-         }
-         $R[i]\leftarrow R[i] \setminus \{a\}$\tcp*{Un-assign attribute}
-      }
-      $R[j]\leftarrow R[j] \cup \{a\}$\tcp*{Assign to best partition}
-   }
-   \If(\tcp*[f]{Has lower cost}){$||R||=|A| \wedge L(R, B, Q)<c^*$}{
-      $c^* \leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
-      $\mathcal{P}(B)\leftarrow R$\tcp*{Update the best partitioning}
-   }
-}
-\Return $\mathcal{P}(B)$ \tcp*{Final set of sub-blocks}
-\end{algorithm}
-
-\subsection{Mixed Integer Linear Program Formulation}
-$x_{a,p}$: $1$ if attribute $a$ is in partition $p$, $0$ otherwise.\\
-$y_{p,q}$: $1$ if partition $p$ is used by query $q$, $0$ otherwise.\\
-$z_{a,p,q}$: $1$ if partition $p$ is used by query $q$, and attribute $a$ is in partition $p$, $0$ otherwise.\\
-$u_{p}$: $1$ if partition $p$ is assigned at least $1$ attribute, $0$ otherwise.\\
-
-Let $q(a)\equiv \mathbf{1}(a \in q.A)$.
-
-% NOTE: In the simulation we will have query weights that will factor into the objective function
-\begin{eqnarray}
-\text{minimize}  
-    \sum_{q\in Q} \sum_{p=1}^{k}(16\cdot c_e(B) &+& 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
-    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
-\text{subject to}&&\nonumber\\
-\forall_{a\in A}, 
-    && \sum_{p=1}^{k} x_{a,p} = 1\nonumber\\
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  \sum_{a\in A} q(a)\cdot x_{a,p} - y_{p,q} \geq 0 \nonumber\\
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  K\cdot y_{p,q} - \sum_{a\in A} q(a)\cdot x_{a,p}  \geq 0 \nonumber\\
-\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
-    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\nonumber\\
-\forall_{p\in[1..k]},
-    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
-\forall_{p\in[1..k]},
-    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
-\Big(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\Big) &\cdot& \Big(\sum_{p=1}^{k} u_p -1\Big) \leq \alpha \nonumber
-\end{eqnarray}
-
-\subsection{Overlapping Attributes}
-
-\paragraph*{Problem.}\emph{Find an overlapping partitioning of attributes that minimizes
-  the query I/O and bounds the storage cost by some upper limit.}
-
-\paragraph*{Approach.} We start the algorithm with a 
- partitioning based on what queries we have seen.
-Every query gets its own sub-block. This is the``ideal'' partitioning,
-because the I/O cost would be minimized for every query that we would
-have seen. As input to the algorithm, we provide a function $\dist$, which
-is used to compute the distance between two blocks.
-
-$$
-\dist: block \rightarrow block \rightarrow int
-$$
-
-Note that the $\dist$ function can be chosen arbitrarily.  Examples of useful
-$\dist$ functions are the frequencies with which the queries are accessed, or the Jaccard index, which measures the set-similarity of
-the attributes.
-
-Using this function, the algorithm iteratively combines the two partitions that
-are closest together. Algorithm~\ref{alg:closestBF} shows the pseudocode for a brute-force approach
-  to finding the closest paris that runs in $O(n^2)$ time. However, there is
-  also a divide-and-conquer algorithm~\cite{cormen01} for solving closest pair
-  of points problem in $O(n \log n)$.
-After each combination of partitions, the algorithm calculate the storage cost
-for the partitioning. The algorithm stops when the  storage cost is below some
-specified threshold.  The result is the block partitioning.
-
-\begin{algorithm}[h]
-\SetKwFunction{closestPair}{closestPair}
-\scriptsize
-\caption{Algorithm for partitioning blocks into sub-blocks with overlapping attributes.}
-\label{alg:overlappingP}
-\KwData{$B$: block, $Q$: set of queries}
-$\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}
-
-\For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
- $\mathcal{P}(B) = \mathcal{P}(B) \cup q.A$ 
-}
-
-\While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
-  \text{max}$}{
-$\text{x}, \text{y}  \leftarrow  \closestPair{}  $\;
- $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - \{x, y\}$ \;
-  $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  \cup \{ x \cup y \}$ \;
-}
-\Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
-\end{algorithm}
-
- \begin{algorithm}[h]
-\SetKwFunction{length}{length}
-\SetKwFunction{dist}{dist}
- \scriptsize
- \caption{Brute-force algorithm for closest points in $O(n^2)$.}
- \label{alg:closestBF}
-
-$minDist  \gets \infty $ \;
-\For{$i \gets 1$ \textbf{to} $\length(P)-1$} {
-\For{$j \gets i+ 1$ \textbf{to} $\length(P)$} {
-
-  $p \gets P[i]$ \;
-  $q \gets P[j]$ \;
-  \If{$\dist(p,q) < minDist$} {
-    $minDist \gets \dist(p,q)$\;
-    $closestPair \gets (p,q)$\;
-  }
-}
-}
-\Return $closestPair$
-
-% \KwData{$B$: block, $Q$: set of queries}
-
-% $\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}
-
-% \For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
-%  $\mathcal{P}(B) = \mathcal{P}(B) \cup q.A$ 
-% }
-
-% \While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
-%   \text{max}$}{
-% $\text{x}, \text{y}  \leftarrow  \closestPair{}  $\;
-%  $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - x - y$ \;
-%   $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  + x \cup y$ \;
-
-% \Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
- \end{algorithm}
-
-%\begin{acknowledgements}
-%If you'd like to thank anyone, place your comments here
-%and remove the percent signs.
-%\end{acknowledgements}
-
-\bibliographystyle{spmpsci}
-\bibliography{main}
-
-\end{document}
-% end of file template.tex

### 2014-07-20 docs/paper/text/problem.tex
moved paper to a more moduler format

--- 2014-07-16 docs/graph-adapt/main.tex
@@
-\documentclass[twocolumn]{svjour3}          % twocolumn
-%
-\smartqed  % flush right qed marks, e.g. at end of proof
-%
-
-\usepackage{alltt}
-\usepackage{enumerate}
-\usepackage[ruled,vlined]{algorithm2e}
-\usepackage{xargs}
-\usepackage{amssymb}
-\usepackage{amsmath}
-
-\renewcommand{\CommentSty}[1]{\textnormal{#1}}
-\DontPrintSemicolon
-\SetKwComment{tcp}{$\triangleright$ }{}
-\SetVlineSkip{0cm}
-\SetAlgoSkip{}
-
-\let\oldemptyset\emptyset
-\let\emptyset\varnothing
-
-\DeclareMathOperator{\dist}{dist}
-
-% Insert the name of "your journal" with
-\journalname{VLDBJ}
-%
-\begin{document}
-
-\title{Adaptive Disk Storage for Interaction Graphs}
-
-%\titlerunning{Short form of title}        % if too long for running head
-
-\author{Robert Soul\'{e}         \and
-        Bu\u{g}ra Gedik%etc.
-}
-
-%\authorrunning{Short form of author list} % if too long for running head
-
-\institute{B. Gedik \at            
-              \email{bgedik@cs.bilkent.edu.tr}           %  \\
-           \and
-           R. Soul\'{e} \at
-               \email{robert.soule@usi.ch}     
-}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
-
-\begin{abstract}
-Insert your abstract here. Include keywords, PACS and mathematical
-subject classification numbers as needed.
-\keywords{First keyword \and Second keyword \and More}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
-\end{abstract}
-
-\section{Introduction}
-An \emph{interaction graph} is an append-only graph, where new edges and
-vertices are added as time progresses.%
-
-Typical interaction graphs have data associated with their vertices and edges.
-We call this data \emph{attributes} or \emph{properties}. Most algorithms
-access this data as they traverse the graph. However, not all of the data is
-accessed by all the queries. Typically, there are correlations among the
-attributes accessed by different queries, such as Q1 and Q5 accessing
-attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and so on.%
-
-In an interaction graph database organized as blocks of temporal neighborlists
-(as in Gedik et al.~\cite{gedik14}), it is important to store the edge and
-vertex properties locally. For  instance, if the edge properties are kept away
-in a relational table, there will be almost no benefit to the locality
-optimizations performed for block organization (discussed in Gedik et
-al.~\cite{gedik14}, as we would have to go back and forth between the disk
-blocks to access the edge attributes).%
-
-On the other hand, putting all the edge attributes into the disk blocks
-containing the graph structure is expected to cause significant overhead when
-only a few of these attributes are read. This is somewhat similar to the 
-problem with row-oriented databases, where the entire row needs to be accessed
-from the disk despite the query needing only a small fraction of it.
-Unfortunately, there is no clear correspondence to a column-oriented database
-layout for the graph databases.%
-
-An important characteristic of interaction graphs are that they are temporal.
-As such, the co-access correlations for the attributes can be different for
-different temporal regions. It might also be unknown at the insertion time and
-may be discovered later depending on the workload. This points to the need for
-adaptively optimizing the layout (somewhat similar to H2O \cite{alagiannis14}
-and HYRISE~\cite{grund10}).%
-
-I imagine a solution to this problem, which I name the `rail layout'. The idea
-is to start with large blocks that contain the entire data. As we learn about
-the access properties for different time regions, we might start splitting
-such blocks into smaller blocks that run parallel to each other, almost like
-having two or more graph databases for certain time regions, each containing a
- different subset of the attributes, but with a link between them in case a
-query needs to access both.
-
 \section{Problem Description}
@@
 
-\section{Cost Model}
+\subsection{Cost Model}
 Let $Q$ be the query workload, where each query $q\in Q$ accesses a set of
@@
 
-\subsection{Query I/O}
+\subsubsection{Query I/O}
 
@@
 %
-\begin{algorithm}[h]
+\begin{algorithm}[ht]
 \scriptsize
@@
 \end{equation}
-
-\section{Partitioning}
-
-TODO: This section is still a work-in-progress. We are trying to find the algorithms.
-
-\subsection{Non-Overlapping Attributes}
-
-\paragraph*{Problem.}\emph{Find a true partitioning of attributes that minimizes
-  the query I/O and bounds the storage cost by some upper limit.}
-
-%$\mathcal{P}(B) \leftarrow \emptyset$ \tcp*{Initial partitioning}
-
-\begin{algorithm}[h]
-\scriptsize
-\caption{Algorithm for partitioning blocks into sub-blocks with non-overlapping attributes.}
-\label{alg:overlappingP}
-\KwData{$B$: block, $Q$: set of queries}
-$c^*\leftarrow \infty$ \tcp*{Lowest cost over all \# of partitions}
-\For(\tcp*[f]{For each possible \# of partitions}){$k=1$ to $|A|$}{
-   $R[i]\leftarrow \emptyset, \forall i\in [1..k]$ \tcp*{Initialize partitions}
-   \For(\tcp*[f]{For each attribute}){$a \in A$\textnormal{, in decr.\/ order of }$f(a)$}{
-      $c\leftarrow \infty$ \tcp*{Lowest cost over all assignments}  
-      $j\leftarrow -1$ \tcp*{Best partition assignment}
-      \For(\tcp*[f]{For each partition assignment}){$i\in [1..k]$} {
-         $R[i]\leftarrow R[i] \cup \{a\}$\tcp*{Assign attribute}
-         \If(\tcp*[f]{If size is not feasable}){$H(R, B, Q)\geq\alpha$}{
-            \textbf{continue}
-         }
-         \If(\tcp*[f]{If query cost is lower}){$L(R, B, Q)<c$}{
-            $c\leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
-            $j\leftarrow i$\tcp*{Update the best partition}
-         }
-         $R[i]\leftarrow R[i] \setminus \{a\}$\tcp*{Un-assign attribute}
-      }
-      $R[j]\leftarrow R[j] \cup \{a\}$\tcp*{Assign to best partition}
-   }
-   \If(\tcp*[f]{Has lower cost}){$||R||=|A| \wedge L(R, B, Q)<c^*$}{
-      $c^* \leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
-      $\mathcal{P}(B)\leftarrow R$\tcp*{Update the best partitioning}
-   }
-}
-\Return $\mathcal{P}(B)$ \tcp*{Final set of sub-blocks}
-\end{algorithm}
-
-\subsection{Mixed Integer Linear Program Formulation}
-$x_{a,p}$: $1$ if attribute $a$ is in partition $p$, $0$ otherwise.\\
-$y_{p,q}$: $1$ if partition $p$ is used by query $q$, $0$ otherwise.\\
-$z_{a,p,q}$: $1$ if partition $p$ is used by query $q$, and attribute $a$ is in partition $p$, $0$ otherwise.\\
-$u_{p}$: $1$ if partition $p$ is assigned at least $1$ attribute, $0$ otherwise.\\
-
-Let $q(a)\equiv \mathbf{1}(a \in q.A)$.
-
-% NOTE: In the simulation we will have query weights that will factor into the objective function
-\begin{eqnarray}
-\text{minimize}  
-    \sum_{q\in Q} \sum_{p=1}^{k}(16\cdot c_e(B) &+& 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
-    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
-\text{subject to}&&\nonumber\\
-\forall_{a\in A}, 
-    && \sum_{p=1}^{k} x_{a,p} = 1\nonumber\\
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  \sum_{a\in A} q(a)\cdot x_{a,p} - y_{p,q} \geq 0 \nonumber\\
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  K\cdot y_{p,q} - \sum_{a\in A} q(a)\cdot x_{a,p}  \geq 0 \nonumber\\
-\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
-    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\nonumber\\
-\forall_{p\in[1..k]},
-    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
-\forall_{p\in[1..k]},
-    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
-\Big(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\Big) &\cdot& \Big(\sum_{p=1}^{k} u_p -1\Big) \leq \alpha \nonumber
-\end{eqnarray}
-
-\subsection{Overlapping Attributes}
-
-\paragraph*{Problem.}\emph{Find an overlapping partitioning of attributes that minimizes
-  the query I/O and bounds the storage cost by some upper limit.}
-
-\paragraph*{Approach.} We start the algorithm with a 
- partitioning based on what queries we have seen.
-Every query gets its own sub-block. This is the``ideal'' partitioning,
-because the I/O cost would be minimized for every query that we would
-have seen. As input to the algorithm, we provide a function $\dist$, which
-is used to compute the distance between two blocks.
-
-$$
-\dist: block \rightarrow block \rightarrow int
-$$
-
-Note that the $\dist$ function can be chosen arbitrarily.  Examples of useful
-$\dist$ functions are the frequencies with which the queries are accessed, or the Jaccard index, which measures the set-similarity of
-the attributes.
-
-Using this function, the algorithm iteratively combines the two partitions that
-are closest together. Algorithm~\ref{alg:closestBF} shows the pseudocode for a brute-force approach
-  to finding the closest paris that runs in $O(n^2)$ time. However, there is
-  also a divide-and-conquer algorithm~\cite{cormen01} for solving closest pair
-  of points problem in $O(n \log n)$.
-After each combination of partitions, the algorithm calculate the storage cost
-for the partitioning. The algorithm stops when the  storage cost is below some
-specified threshold.  The result is the block partitioning.
-
-\begin{algorithm}[h]
-\SetKwFunction{closestPair}{closestPair}
-\scriptsize
-\caption{Algorithm for partitioning blocks into sub-blocks with overlapping attributes.}
-\label{alg:overlappingP}
-\KwData{$B$: block, $Q$: set of queries}
-$\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}
-
-\For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
- $\mathcal{P}(B) = \mathcal{P}(B) \cup q.A$ 
-}
-
-\While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
-  \text{max}$}{
-$\text{x}, \text{y}  \leftarrow  \closestPair{}  $\;
- $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - \{x, y\}$ \;
-  $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  \cup \{ x \cup y \}$ \;
-}
-\Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
-\end{algorithm}
-
- \begin{algorithm}[h]
-\SetKwFunction{length}{length}
-\SetKwFunction{dist}{dist}
- \scriptsize
- \caption{Brute-force algorithm for closest points in $O(n^2)$.}
- \label{alg:closestBF}
-
-$minDist  \gets \infty $ \;
-\For{$i \gets 1$ \textbf{to} $\length(P)-1$} {
-\For{$j \gets i+ 1$ \textbf{to} $\length(P)$} {
-
-  $p \gets P[i]$ \;
-  $q \gets P[j]$ \;
-  \If{$\dist(p,q) < minDist$} {
-    $minDist \gets \dist(p,q)$\;
-    $closestPair \gets (p,q)$\;
-  }
-}
-}
-\Return $closestPair$
-
-% \KwData{$B$: block, $Q$: set of queries}
-
-% $\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}
-
-% \For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
-%  $\mathcal{P}(B) = \mathcal{P}(B) \cup q.A$ 
-% }
-
-% \While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
-%   \text{max}$}{
-% $\text{x}, \text{y}  \leftarrow  \closestPair{}  $\;
-%  $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - x - y$ \;
-%   $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  + x \cup y$ \;
-
-% \Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
- \end{algorithm}
-
-%\begin{acknowledgements}
-%If you'd like to thank anyone, place your comments here
-%and remove the percent signs.
-%\end{acknowledgements}
-
-\bibliographystyle{spmpsci}
-\bibliography{main}
-
-\end{document}
-% end of file template.tex

### 2014-07-20 docs/paper/text/solution.tex
added ILP solution for the overlapping case

@@
 
-%$\mathcal{P}(B) \leftarrow \emptyset$ \tcp*{Initial partitioning}
+\subsubsection{Mixed Integer Linear Program Formulation}
+$x_{a,p}$: $1$ if attribute $a$ is in partition $p$, $0$ otherwise.\\
+$y_{p,q}$: $1$ if partition $p$ is used by query $q$, $0$ otherwise.\\
+$z_{a,p,q}$: $1$ if partition $p$ is used by query $q$, and attribute $a$ is in partition $p$, $0$ otherwise.\\
+$u_{p}$: $1$ if partition $p$ is assigned at least $1$ attribute, $0$ otherwise.\\
+
+Let $q(a)\equiv \mathbf{1}(a \in q.A)$.
+
+% NOTE: In the simulation we will have query weights that will factor into the objective function
+\begin{eqnarray}
+\text{minimize}  
+    \sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
+    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
+\text{subject to}&&\nonumber\\
+\forall_{a\in A}, 
+    && \sum_{p=1}^{k} x_{a,p} = 1\nonumber\\
+\forall_{\{p,q\}\in [1..k]\times Q}, 
+    &&  \sum_{a\in A} q(a)\cdot x_{a,p} - y_{p,q} \geq 0 \nonumber\\
+\forall_{\{p,q\}\in [1..k]\times Q}, 
+    &&  K\cdot y_{p,q} - \sum_{a\in A} q(a)\cdot x_{a,p}  \geq 0 \nonumber\\
+\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
+    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\nonumber\\
+\forall_{p\in[1..k]},
+    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
+\forall_{p\in[1..k]},
+    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
+&& \sum_{p=1}^{k} u_p -1 \leq \frac{\alpha}
+  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}} \nonumber
+\end{eqnarray}
+
+Here we use the following LP construction. Assume we have two variables,
+$\beta_1$ and $\beta_2$, where $\beta_2\in[0,1]$ and $\beta_1\geq 0$. We want to
+implement the following constraint: 
+$$
+\beta_2 = 
+\begin{cases}
+0 & \beta_1 = 0 \\
+1 & \beta_1 \geq 0 
+\end{cases}
+$$
+
+This could be expressed as a linear constraint as follows,  where $K$ is a
+large constant guaranteed to be larger  that $\beta_1$ for practical purposes:
+\begin{eqnarray}
+\hspace{2cm}&& \beta_1 - \beta_2 \geq 0\nonumber\\
+\hspace{2cm}&& K\cdot\beta_2 - \beta_1 \geq 0\nonumber
+\end{eqnarray}
+
+In our ILP formulation, we use this construction twice. The first use is shown
+in the second and third constrants, where $\beta_1=\sum_{a\in A} q(a)\cdot
+x_{a,p}$ and $\beta_2=y_{p,q}$. The second use is shown in the fifth and 
+sixth constaints, where $\beta_1=\sum_{a\in A} x_{a,p}$ and $\beta_2=u_p$.
+
+\subsubsection{Heuristic Solution}
+
+\paragraph*{Approach.} TODO
 
@@
 
-\subsection{Mixed Integer Linear Program Formulation}
-$x_{a,p}$: $1$ if attribute $a$ is in partition $p$, $0$ otherwise.\\
-$y_{p,q}$: $1$ if partition $p$ is used by query $q$, $0$ otherwise.\\
-$z_{a,p,q}$: $1$ if partition $p$ is used by query $q$, and attribute $a$ is in partition $p$, $0$ otherwise.\\
-$u_{p}$: $1$ if partition $p$ is assigned at least $1$ attribute, $0$ otherwise.\\
+\subsection{Overlapping Attributes}
 
-Let $q(a)\equiv \mathbf{1}(a \in q.A)$.
+\paragraph*{Problem.}\emph{Find an overlapping partitioning of attributes that minimizes
+  the query I/O and bounds the storage cost by some upper limit.}
 
+\subsubsection{Mixed Integer Linear Program Formulation}
 % NOTE: In the simulation we will have query weights that will factor into the objective function
@@
 \forall_{a\in A}, 
-    && \sum_{p=1}^{k} x_{a,p} = 1\nonumber\\
+    && \sum_{p=1}^{k} x_{a,p} > 0\nonumber\\
+\forall_{a\in A}, 
+    && \sum_{p=1}^{k} x_{a,p} \leq k\nonumber\\
+\forall_{\{a,q\}\in A\times Q},
+    &&  \sum_{p=1}^{k} z_{a,p,q} \geq q(a) \nonumber\\
 \forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  \sum_{a\in A} q(a)\cdot x_{a,p} - y_{p,q} \geq 0 \nonumber\\
+    &&  \sum_{a\in A} z_{a,p,q} - y_{p,q} \geq 0 \nonumber\\
 \forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  K\cdot y_{p,q} - \sum_{a\in A} q(a)\cdot x_{a,p}  \geq 0 \nonumber\\
+    &&  K\cdot y_{p,q} - \sum_{a\in A} z_{a,p,q}  \geq 0 \nonumber\\
 \forall_{\{a,p,q\}\in A\times [1..k]\times Q}, 
-    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\nonumber\\
+    && z_{a,p,q} - (y_{p,q} + q(a)) \geq -1 \nonumber\\
 \forall_{p\in[1..k]},
@@
     && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
- && \sum_{p=1}^{k} u_p -1 \leq \frac{\alpha}{
-1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}
- } \nonumber
+&& \sum_{p=1}^{k} u_p -1 \leq \frac{\alpha}
+  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}} \nonumber
 \end{eqnarray}
 
-Here we use the following LP construction. Assume we have two variables,
-$\beta_1$ and $\beta_2$, where $\beta_2\in[0,1]$ and $\beta_1\geq 0$. We want to
-implement the following constraint: 
-$$
-\beta_2 = 
-\begin{cases}
-0 & \beta_1 = 0 \\
-1 & \beta_1 \geq 0 
-\end{cases}
-$$
-
-This could be expressed as a linear constraint as follows,  where $K$ is a
-large constant guaranteed to be larger  that $\beta_1$ for practical purposes:
-\begin{eqnarray}
-\hspace{2cm}&& \beta_1 - \beta_2 \geq 0\nonumber\\
-\hspace{2cm}&& K\cdot\beta_2 - \beta_1 \geq 0\nonumber
-\end{eqnarray}
-
-In our ILP formulation, we use this construction twice. The first use is shown
-in the second and third constrants, where $\beta_1=\sum_{a\in A} q(a)\cdot
-x_{a,p}$ and $\beta_2=y_{p,q}$. The second use is shown in the fifth and 
-sixth constaints, where $\beta_1=\sum_{a\in A} x_{a,p}$ and $\beta_2=u_p$.
-
-\subsection{Overlapping Attributes}
-
-\paragraph*{Problem.}\emph{Find an overlapping partitioning of attributes that minimizes
-  the query I/O and bounds the storage cost by some upper limit.}
+\subsubsection{Heuristic Solution}
 

### 2014-07-20 docs/paper/text/problem.tex
added the detailed description of the ILP for the case of non-overlapping partitions

@@
 
-\paragraph{\textbf{Other useful equations}}
-
-\noindent The following are unused, but may be needed for the partitioning
-algorithm.
-
-Let $f: \mathbb{R}\times A \rightarrow \mathbb{N}$ be a function that maps a
-attribute-timestamp pair $(a, t)$ to the frequency of queries that access
-attribute $a$ for time $t$. More formally:
-\begin{equation}
-f(a,t) = \sum_{q\in Q} \textbf{1}(a\in q.A \wedge t\in q.T)
-\end{equation}
-
-Let $T$ represent a time range $[T.s, T.e]$. We denote the frequency of
-attribute access for the time interval $T$ for attribute $a$ as follows:
-\begin{equation}
-f(a, T) = \int_{T.s}^{T.e} f(a, t)\cdot dt =  \sum_{q\in Q} |q.T \cap T| 
-\end{equation}
+%\paragraph{\textbf{Other useful equations}}
+%\noindent The following are unused, but may be needed for the partitioning
+%algorithm.
+%
+%Let $f: \mathbb{R}\times A \rightarrow \mathbb{N}$ be a function that maps a
+%attribute-timestamp pair $(a, t)$ to the frequency of queries that access
+%attribute $a$ for time $t$. More formally:
+%\begin{equation}
+%f(a,t) = \sum_{q\in Q} \textbf{1}(a\in q.A \wedge t\in q.T)
+%\end{equation}
+%
+%Let $T$ represent a time range $[T.s, T.e]$. We denote the frequency of
+%attribute access for the time interval $T$ for attribute $a$ as follows:
+%\begin{equation}
+%f(a, T) = \int_{T.s}^{T.e} f(a, t)\cdot dt =  \sum_{q\in Q} |q.T \cap T| 
+%\end{equation}

### 2014-07-20 docs/paper/text/solution.tex
added the detailed description of the ILP for the case of non-overlapping partitions

@@
+\clearpage
+\newpage
 \section{Partitioning}
@@
 
-\paragraph*{Problem.}\emph{Find a true partitioning of attributes that minimizes
-  the query I/O and bounds the storage cost by some upper limit.}
+\paragraph*{Problem.} \emph{Find a true partitioning of attributes that
+minimizes the query I/O and bounds the storage cost by some upper limit.}
 
-\subsubsection{Mixed Integer Linear Program Formulation}
-$x_{a,p}$: $1$ if attribute $a$ is in partition $p$, $0$ otherwise.\\
-$y_{p,q}$: $1$ if partition $p$ is used by query $q$, $0$ otherwise.\\
-$z_{a,p,q}$: $1$ if partition $p$ is used by query $q$, and attribute $a$ is in partition $p$, $0$ otherwise.\\
-$u_{p}$: $1$ if partition $p$ is assigned at least $1$ attribute, $0$ otherwise.\\
+\subsubsection{Integer Linear Program Formulation}
+We present an ILP formuation of the problem. For this purpose, we define a 
+number of binary ($0$ or $1$) variables: 
+\begin{itemize}
+\item $x_{a,p}$: $1$ if attribute $a$ is in partition $p$, $0$ otherwise.
+\item $y_{p,q}$: $1$ if partition $p$ is used by query $q$, $0$ otherwise.
+\item $z_{a,p,q}$: $1$ if partition $p$ is used by query $q$ and attribute $a$
+is in partition $p$, $0$ otherwise.
+\item $u_{p}$: $1$ if partition $p$ is assigned at least $1$ attribute, $0$ otherwise.
+\end{itemize}
 
-Let $q(a)\equiv \mathbf{1}(a \in q.A)$.
+Each of these variables serve an important purpose:
+\begin{itemize}
+\item $x$s define the attribute to partition assignment.
+\item $y$s help formulate the objective function's I/O contributions
+that are on a per-partition basis.
+\item $z$s help formulate the objective function's I/O contributions
+that are on a per-attribute basis.
+\item $u$s help formulate the storage overhead requirement.
+\end{itemize}
+
+We also define a helper notation for representing whether a variable is
+accessed by a query or not: $q(a)\equiv \mathbf{1}(a \in q.A)$, where
+$\mathbf{1}$ is the indicator function. We assume that there are a maximum of
+$k$ partitions. $k$ can be taken as $|A|$, as there cannot be more partitions
+than there are attributes.
+
+We are now ready to state the ILP formulation. We start with the objective
+function, that is the total query I/O cost:
+\begin{eqnarray}
+\sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot
+y_{p,q}\nonumber\\ 
+&+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q}\label{eq:no-obj}
+\end{eqnarray}
+
+In Eq.~\ref{eq:no-obj}, we simply sum for each query and each partiton to add
+the I/O cost of reading in the structural information found in a sub-block, if
+the partition is used by the query. We then sum over each attribute as well,
+and add the I/O cost of reading in the attributes. Note that $z_{a,p,q}$ could
+have been  replaced with $x_{a,p}\cdot y_{p,q}$, but that would make the
+objective function non-linear.
+
+Our first constraint is that each attribute gets assigned to one and only one
+partition. Formally:
+\begin{eqnarray}
+\forall_{a\in A}, \sum_{p=1}^{k} x_{a,p} = 1
+\end{eqnarray}
+
+Our second constraint is that, if a query contains an attribute assigned to a
+partition, then that partition is used by the query. In essence, we want to
+state: $\forall_{\{p,q\}\in [1..k]\times Q}, y_{p,q} = \mathbf{1}(\sum_{a\in
+A} q(a)\cdot x_{a,p}>0)$. In order to formulate this we use the following ILP 
+construction:
+
+Assume we have two variables, $\beta_1$ and $\beta_2$, where $\beta_2\in[0,1]$
+and $\beta_1\geq 0$. We want to implement the following constraint: $\beta_2 =
+\mathbf{1}(\beta_1 > 0)$.
+
+This could be expressed as a linear constraint as follows, where $K$ is a
+large constant guaranteed to be larger than $\beta_1$ for practical purposes:
+\begin{eqnarray}
+&& \beta_1 - \beta_2 \geq 0\nonumber\\
+&& K\cdot\beta_2 - \beta_1 \geq 0\nonumber
+\end{eqnarray}
+
+We now apply this construction to our second constraint, where
+$\beta_1=\sum_{a\in A} q(a)\cdot x_{a,p}$ and $\beta_2=y_{p,q}$. This results
+in the following linear constraints:
+\begin{eqnarray}
+\forall_{\{p,q\}\in [1..k]\times Q}, 
+    &&  \sum_{a\in A} q(a)\cdot x_{a,p} - y_{p,q} \geq 0 \nonumber\\
+\forall_{\{p,q\}\in [1..k]\times Q}, 
+    &&  K\cdot y_{p,q} - \sum_{a\in A} q(a)\cdot x_{a,p}  \geq 0 
+\end{eqnarray}
+
+Our third constraint is that, if an attribute $a$ is assigned to a partition
+$p$, and partition $p$ is used by a query $q$, then the corresponding $z$
+variable must be set to $1$. It must be $0$ in all other cases. That is, we
+want: $\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
+z_{a,p,q}=\mathbf{1}(x_{a,p} = y_{p,q} = 1)$. We express this as a linear 
+constraint, as follows:
+\begin{eqnarray}
+\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
+    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\label{eq:no-z}
+\end{eqnarray}
+
+In Eq.~\ref{eq:no-z}, when the $x$ and $y$ variable are both $1$, the  $z$
+variable is simply forced to be $1$. Otherwise, the $z$ variable can be either
+$0$ or $1$, but since the $z$ variables appear in the objective fucntion as
+positive terms, the solver will set them to $0$, which is what we want.
+
+Our fourth constraint is that, if a partition is non-empty, then its
+corresponding $u$ variable must be set to $0$. It must be $0$ otherwise. In
+other words,  we want $u_p = \mathbf{1}(\sum_{a\in A} x_{a,p}>0)$. This is
+expressed as linear constraints, as follows:
+\begin{eqnarray}
+\forall_{p\in[1..k]},
+    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
+\forall_{p\in[1..k]},
+    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \label{eq:no-u}
+\end{eqnarray}
+
+Eq.~\ref{eq:no-u} uses the same construction as the second constraint, where
+$\beta_1=\sum_{a\in A} x_{a,p}$ and $\beta_2=u_p$.
+
+Our fifth, and the last, constraint deals with the storage overhead. We want to
+ make sure that the storage overhead does not go over $\alpha$. The storage
+overhead depends on the number of partitions used. That means that the only 
+ILP variables it depends on is the $u$s. In particular, the number of
+partitions used is given by $\sum_{p=1}^{k} u_p$. This results in the
+following linear constraint:
+\begin{equation}
+\sum_{p=1}^{k} u_p - 1 \leq \frac{\alpha}
+  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}}
+\end{equation}
 
 % NOTE: In the simulation we will have query weights that will factor into the objective function
+\begin{figure}[!t]
+\begin{mdframed}
 \begin{eqnarray}
@@
 \end{eqnarray}
+\end{mdframed}
+\caption{ILP formulation for the non-overlapping partitioning}
+\label{fig:no-ilp}
+\end{figure}
 
-Here we use the following LP construction. Assume we have two variables,
-$\beta_1$ and $\beta_2$, where $\beta_2\in[0,1]$ and $\beta_1\geq 0$. We want to
-implement the following constraint: 
-$$
-\beta_2 = 
-\begin{cases}
-0 & \beta_1 = 0 \\
-1 & \beta_1 \geq 0 
-\end{cases}
-$$
-
-This could be expressed as a linear constraint as follows,  where $K$ is a
-large constant guaranteed to be larger  that $\beta_1$ for practical purposes:
-\begin{eqnarray}
-\hspace{2cm}&& \beta_1 - \beta_2 \geq 0\nonumber\\
-\hspace{2cm}&& K\cdot\beta_2 - \beta_1 \geq 0\nonumber
-\end{eqnarray}
-
-In our ILP formulation, we use this construction twice. The first use is shown
-in the second and third constrants, where $\beta_1=\sum_{a\in A} q(a)\cdot
-x_{a,p}$ and $\beta_2=y_{p,q}$. The second use is shown in the fifth and 
-sixth constaints, where $\beta_1=\sum_{a\in A} x_{a,p}$ and $\beta_2=u_p$.
+The final ILP formulation for the non-overlapping partitioning is given in
+Figure~\ref{fig:no-ilp}.
 
@@
 
-\paragraph*{Approach.} TODO
+\paragraph*{Approach.$\,$}
+TODO
 
@@
 
+\clearpage
+\newpage
 \subsection{Overlapping Attributes}
 
-\paragraph*{Problem.}\emph{Find an overlapping partitioning of attributes that minimizes
+\paragraph*{Problem.} \emph{Find an overlapping partitioning of attributes that minimizes
   the query I/O and bounds the storage cost by some upper limit.}
 
-\subsubsection{Mixed Integer Linear Program Formulation}
+\subsubsection{Integer Linear Program Formulation}
 % NOTE: In the simulation we will have query weights that will factor into the objective function

### 2014-07-20 docs/paper/text/solution.tex
minor fix to the overlapping case
minor fix to the non-overlapping case ILP formulation text

@@
 
-\paragraph*{Problem.} \emph{Find a true partitioning of attributes that
+\paragraph*{Problem.$\,$} \emph{Find a true partitioning of attributes that
 minimizes the query I/O and bounds the storage cost by some upper limit.}
@@
 
-Each of these variables serve an important purpose:
+Each of these variables serve a purpose:
 \begin{itemize}
-\item $x$s define the attribute to partition assignment.
-\item $y$s help formulate the objective function's I/O contributions
-that are on a per-partition basis.
-\item $z$s help formulate the objective function's I/O contributions
-that are on a per-attribute basis.
+\item $x$s define the attribute-to-partition assignment.
+\item $y$s help formulate the query I/O contribution
+of partitions, excluding their assigned attributes.
+\item $z$s help formulate the query I/O contribution
+of partitions, only considering their assigned attributes.
 \item $u$s help formulate the storage overhead requirement.
@@
 We are now ready to state the ILP formulation. We start with the objective
-function, that is the total query I/O cost:
+function, that is the total query I/O:
 \begin{eqnarray}
@@
 
-In Eq.~\ref{eq:no-obj}, we simply sum for each query and each partiton to add
-the I/O cost of reading in the structural information found in a sub-block, if
-the partition is used by the query. We then sum over each attribute as well,
-and add the I/O cost of reading in the attributes. Note that $z_{a,p,q}$ could
-have been  replaced with $x_{a,p}\cdot y_{p,q}$, but that would make the
-objective function non-linear.
+In Eq.~\ref{eq:no-obj}, we simply sum for each query and each partition, and
+add the I/O cost of reading in the structural information found in a
+sub-block, if the partition is used by the query. We then sum over each
+attribute as well, and add the I/O cost of reading in the attributes. Note
+that $z_{a,p,q}$ could have been  replaced with $x_{a,p}\cdot y_{p,q}$, but
+that would make the objective function non-linear.
 
-Our first constraint is that each attribute gets assigned to one and only one
-partition. Formally:
+We are now ready to state our constraints. Our first constraint is that, each
+attribute must be assigned to a single partition. Formally:
 \begin{eqnarray}
@@
 
-Our second constraint is that, if a query contains an attribute assigned to a
-partition, then that partition is used by the query. In essence, we want to
-state: $\forall_{\{p,q\}\in [1..k]\times Q}, y_{p,q} = \mathbf{1}(\sum_{a\in
-A} q(a)\cdot x_{a,p}>0)$. In order to formulate this we use the following ILP 
-construction:
+Our second constraint is that, if a query $q$ contains an attribute $a$
+assigned to a partition $p$, then partition $p$ is used by the query, i.e.,
+$y_{p,q}=1$. In essence, we want to state: $\forall_{\{p,q\}\in [1..k]\times
+Q}, y_{p,q} = \mathbf{1}(\sum_{a\in A} q(a)\cdot x_{a,p}>0)$.
 
+In order to formulate this constraint, we use the following ILP  construction:
 Assume we have two variables, $\beta_1$ and $\beta_2$, where $\beta_2\in[0,1]$
 and $\beta_1\geq 0$. We want to implement the following constraint: $\beta_2 =
-\mathbf{1}(\beta_1 > 0)$.
-
-This could be expressed as a linear constraint as follows, where $K$ is a
-large constant guaranteed to be larger than $\beta_1$ for practical purposes:
+\mathbf{1}(\beta_1 > 0)$. This could be expressed as a linear constraint as
+follows, where $K$ is a large constant guaranteed to be larger than $\beta_1$
+for practical purposes:
 \begin{eqnarray}
@@
 $p$, and partition $p$ is used by a query $q$, then the corresponding $z$
-variable must be set to $1$. It must be $0$ in all other cases. That is, we
-want: $\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
-z_{a,p,q}=\mathbf{1}(x_{a,p} = y_{p,q} = 1)$. We express this as a linear 
-constraint, as follows:
+variable must be set to $1$. That is, we want: $\forall_{\{a,p,q\}\in A\times
+[1..k]\times Q}, z_{a,p,q}=\mathbf{1}(x_{a,p} = y_{p,q} = 1)$. We express this
+as a linear  constraint, as follows:
 \begin{eqnarray}
@@
 
-In Eq.~\ref{eq:no-z}, when the $x$ and $y$ variable are both $1$, the  $z$
+In Eq.~\ref{eq:no-z}, when the $x$ and $y$ variables are both $1$, the  $z$
 variable is simply forced to be $1$. Otherwise, the $z$ variable can be either
@@
 Our fourth constraint is that, if a partition is non-empty, then its
-corresponding $u$ variable must be set to $0$. It must be $0$ otherwise. In
-other words,  we want $u_p = \mathbf{1}(\sum_{a\in A} x_{a,p}>0)$. This is
+corresponding $u$ variable must be set to $0$. In other words,  we want
+$\forall_{p\in[1..k]}, u_p = \mathbf{1}(\sum_{a\in A} x_{a,p}>0)$. This is
 expressed as linear constraints, as follows:
@@
 \forall_{\{a,p,q\}\in A\times [1..k]\times Q}, 
-    && z_{a,p,q} - (y_{p,q} + q(a)) \geq -1 \nonumber\\
+    && z_{a,p,q} - (x_{a,p} + y_{p,q} + q(a)) \geq -2 \nonumber\\
 \forall_{p\in[1..k]},

### 2014-07-29 code/liboptimizer/include/solver.h
Got the liboptimizer and test code compiling again. I either forgot to commit changes before, or there was a merge conflict.

@@ -1,2 +1,15 @@
+#pragma once
 
-int solve(int argc, char *argv[]);
+#include <intergdb/common/QueryWorkload.h>
+
+using namespace intergdb::common;
+
+class Solver
+{
+public:
+    Solver() {}
+    int solve(QueryWorkload * workload);
+
+};
+
+

### 2014-07-29 code/liboptimizer/src/solver.cpp
Got the liboptimizer and test code compiling again. I either forgot to commit changes before, or there was a merge conflict.

@@ -22,110 +22,6 @@ extern "C" {
 }
 #endif
 
-#include <intergdb/common/QueryWorkload.h>
-
-using namespace intergdb::common;
-
-
-/*
-
-    (1) If an attribute is in a query:
-    v_a1_q1 = 1
-    v_a2_q1 = 0
-    v_a1_q2 = 0
-    v_a2_q2 = 1
-
-    (2) If an attribute is in a partition:
-    v_a1_p1 = ?
-    v_a2_p1 = ?
-    v_a1_p2 = ?
-    v_a2_p2 = ?
-
-   (3) If an partition is used by a query
-    v_p1_q1 = ?
-    v_p1_q2 = ?
-    v_p2_q1 = ?
-    v_p2_q2 = ?
-
-   (4) If an attribute and partition are used by a query
-    v_a1_p1_q1 = ?
-    v_a1_p1_q2 = ?
-    v_a2_p2_q1 = ?
-    v_a2_p2_q2 = ?
-
-    (5) add constraint that all attributes must appear in a partition
-    v_a1_p1 + v_a1_p2  = 1
-    v_a2_p1 + v_a2_p2 = 1
-
-
-    (6) the number of attributes use by query from a partition determines the value of v_p1_q1 (either 0 or 1)
-    (v_a1_q1 * v_a1_p1 + v_a2_q1 * v_a2_p1) - v_p1_q1 >= 0
-
-    To ensure that it is 0 or 1, we are going to add to the objective function, where K is some big value:
-   - K * v_p1_q1
-   - K * v_p1_q2
-   - K * v_p2_q1 
-   - K * v_p2_q2
-
-    (7) add constraint to make sure that v_a1_p1 and v_p1_q1 determine  v_a1_p1_q_1
-    v_a1_p1_q1 >= (v_a1_p1 + v_p1_q1) -1
-
-    (8)  objective function : minimize the following, where the ?'s are constants
-
-   Here is what we want:
-   (? * v_a1_p1_q1 + ? * v_a2_p1_q1 + ?)
-  + (? * v_a1_p1_q2 + ? * v_a2_p1_q2 + ?)
-  + (? * v_a1_p2_q1 + ? * v_a2_p2_q1 + ?)
-  + (? * v_a1_p2_q2 + ? * v_a2_p2_q2 + ?)
-
-  - We need a function for computing the partitions
-  - m function: given a list of partitions, and a query, 
-    it tells us the list of functions that need to be accessed.
-
[…]

### 2014-07-29 docs/paper/text/solution.tex
minor fix to the non-overlapping case ILP formulation text
added lp example for overlapping case
added examples for the overlapping scenario

@@
 
-\subsubsection{Integer Linear Program Formulation}
+\subsubsection{Integer Linear Program Formulation}\label{subsubsec:nov-ilp}
 We present an ILP formuation of the problem. For this purpose, we define a 
@@
 && \beta_1 - \beta_2 \geq 0\nonumber\\
-&& K\cdot\beta_2 - \beta_1 \geq 0\nonumber
+&& K\cdot\beta_2 - \beta_1 \geq 0\label{eq:beta-ilp}
 \end{eqnarray}
@@
 \begin{equation}
-\sum_{p=1}^{k} u_p - 1 \leq \frac{\alpha}
+\sum_{p=1}^{k} u_p \leq 1 + \frac{\alpha}
   {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}}
@@
 \text{minimize}  
-    \sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
+    \sum_{q\in Q} \sum_{p=1}^{k} \!\!&&\!\! (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
     &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
@@
     && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
-&& \sum_{p=1}^{k} u_p -1 \leq \frac{\alpha}
+&& \sum_{p=1}^{k} u_p \leq 1 + \frac{\alpha}
   {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}} \nonumber
@@
 \caption{ILP formulation for the non-overlapping partitioning}
-\label{fig:no-ilp}
+\label{fig:nov-ilp}
 \end{figure}
@@
 The final ILP formulation for the non-overlapping partitioning is given in
-Figure~\ref{fig:no-ilp}.
+Figure~\ref{fig:nov-ilp}.
 
@@
 
-\paragraph*{Problem.} \emph{Find an overlapping partitioning of attributes that minimizes
-  the query I/O and bounds the storage cost by some upper limit.}
+\paragraph*{Problem.$\,$} \emph{Find an overlapping partitioning of attributes
+that minimizes the query I/O and bounds the storage cost by some upper
+limit.}
 
 \subsubsection{Integer Linear Program Formulation}
+We present an ILP formuation of the problem, as we did for the case of
+non-overlapping partitions in Section~\ref{subsubsec:nov-ilp}. We use the same
+set of variables and the same objective function. However, the formulation of
+the constraints differ.
+
+Our first constraint is that, each attribute must be assigned to at leat one
+partition. Formally:
+\begin{eqnarray}
+\forall_{a\in A}, \sum_{p=1}^{k} x_{a,p} \geq 1
+\end{eqnarray}
+
+As our second constraint, we require that for each attribute used by a query,
+there needs to be a partition that is used by that query and that contains the
+attribute in question. Formally:
+\begin{eqnarray}
+\forall_{\{a,q\}\in A\times Q}, \sum_{p=1}^{k} z_{a,p,q} \geq q(a) 
+\end{eqnarray}
+
+\begin{sloppypar}
+As our third constraint, we require that if a query is using an attribute from
+a partition, then that partition must contain the attribute. I.e., we need
+to link the $z$ variables with the $x$ variables as
+$\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, (z_{a,p,q} = 1) \implies 
+(x_{a,p} = 1)$. This can be state as linear constraints:
+\begin{eqnarray}
+\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, x_{a,p} - z_{a,p,q} \geq 0 
+\end{eqnarray}
+\end{sloppypar}
+
+As our fourth constraint, we require that if a query is using at least one
+attribute from a partition, then that partition must be used by the query.
+I.e., we need to link the $z$ variables with the $y$ variables as
+$\forall_{\{p,q\}\in [1..k]\times Q}, y_{p,q} = \mathbf{1}(\sum_{a\in A}
+z_{a,p,q}>0)$. As before, we use the ILP construction from
+Eq.~\ref{eq:beta-ilp} for this, where $\beta_2=y_{p,q}$ and $\beta_1 =
+\sum_{a\in A} z_{a,p,q}$. We get:
+\begin{eqnarray}
+\forall_{\{p,q\}\in [1..k]\times Q}, 
+    &&  \sum_{a\in A} z_{a,p,q} - y_{p,q} \geq 0 \nonumber\\
+\forall_{\{p,q\}\in [1..k]\times Q}, 
+    &&  K\cdot y_{p,q} - \sum_{a\in A} z_{a,p,q} \geq 0 
+\end{eqnarray}
+
+Our fifth constraint is that, if an attribute $a$ is assigned to a partition
+$p$, and partition $p$ is used by a query $q$, then the corresponding $z$
+variable must be set to $1$. This is same as the formulation for the
+non-overlapping case from Eq.~\ref{eq:no-z}.
+
+Our sixth constraint is that, if a partition is non-empty, then its
+corresponding $u$ variable must be set to $0$. Again, this is same as the
+formulation for the non-overlapping case from Eq.~\ref{eq:no-u}.
+
+Our seventh, and the last, constraint deals with the storage overhead. However,
+the storage overhead formulation for the overlapping case is different from 
+the one for the non-overlapping. This is because the overhead does not merely
+depend on the number of partitions, as attributes might have to be loaded
+multiple times from different partitions. As a result, we express the overhead
+using base variables as in the objective function. Formally:
+\begin{eqnarray}
+\sum_{p=1}^{k} (16\cdot c_e(B) &+& 12 \cdot c_n(B)) \cdot u_p  \nonumber \\ 
++ \sum_{a\in A} s(a) \!\!&\cdot&\!\! c_n(B)\cdot x_{a,p} \leq s(B)\cdot (1+\alpha)
+\end{eqnarray}
+
 % NOTE: In the simulation we will have query weights that will factor into the objective function
+\begin{figure}[!t]
+\begin{mdframed}
 \begin{eqnarray}
 \text{minimize}  
-    \sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
+    \sum_{q\in Q} \sum_{p=1}^{k} \!\!&&\!\! (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
     &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
@@
 \forall_{a\in A}, 
-    && \sum_{p=1}^{k} x_{a,p} > 0\nonumber\\
-\forall_{a\in A}, 
-    && \sum_{p=1}^{k} x_{a,p} \leq k\nonumber\\
+    && \sum_{p=1}^{k} x_{a,p} \geq 1\nonumber\\
 \forall_{\{a,q\}\in A\times Q},
     &&  \sum_{p=1}^{k} z_{a,p,q} \geq q(a) \nonumber\\
+\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, 
+    && x_{a,p} - z_{a,p,q} \geq 0 \nonumber\\
 \forall_{\{p,q\}\in [1..k]\times Q}, 
@@
 \forall_{\{a,p,q\}\in A\times [1..k]\times Q}, 
-    && z_{a,p,q} - (x_{a,p} + y_{p,q} + q(a)) \geq -2 \nonumber\\
+    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1 \nonumber\\
 \forall_{p\in[1..k]},
@@
     && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
-&& \sum_{p=1}^{k} u_p -1 \leq \frac{\alpha}
-  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}} \nonumber
+\sum_{p=1}^{k} (16\cdot c_e(B) &+& 12 \cdot c_n(B)) \cdot u_p  \nonumber \\ 
++ \sum_{a\in A} s(a) \!\!&\cdot&\!\! c_n(B)\cdot x_{a,p} \leq s(B)\cdot (1+\alpha)\nonumber
 \end{eqnarray}
+\end{mdframed}
+\caption{ILP formulation for the overlapping partitioning}
+\label{fig:ov-ilp}
+\end{figure}
+
+The final ILP formulation for the overlapping partitioning is given in
+Figure~\ref{fig:ov-ilp}.
 
@@
 
-\paragraph*{Approach.} We start the algorithm with a 
+\paragraph*{Approach.$\,$} We start the algorithm with a 
  partitioning based on what queries we have seen.

### 2014-08-06 docs/paper/text/introduction.tex
Checkpoint. Working on writing an introduction.

@@
 \section{Introduction}
+
+Query workload and disk layout have a significant impact on database
+performance.  For table-based, relational databases, this fact has led database
+designers to develop alternative approaches for storage layout: row-oriented
+storage is more efficient when queries access many attributes from a small
+number of records, and column-oriented storage is more efficient when queries
+access a small number of attributes from many records.
+
+Recently, there has been increased interest in graph databases in general, and
+\emph{interaction graph} databases in particular.  An \emph{interaction graph}
+is an append-only graph, where new edges and vertices are added as time
+progresses.  Thus, interaction graph encode the relationships over time between
+entities in their structure, making them ideal for use in social networking,
+telecommunications, or modeling the world-wide web.
+
+Unfortunately, although interaction graph databases, like relational databases,
+are the target of diverse query workloads, there is no clear correspondence to a
+row-oriented or column-oriented storage layout.
+Typical interaction graphs have data (i.e., \emph{attributes} ) associated with
+their vertices and edges. Most algorithms access this data as they traverse the
+graph.
+
+In an interaction graph database organized as blocks of temporal neighbor lists
+(as in Gedik et al.~\cite{gedik14}), it is important to store the edge and
+vertex attributes locally. For  instance, if the edge properties were kept separately
+in a relational table, there will be almost no benefit to the locality
+optimizations performed for block organization (discussed in Gedik et
+al.~\cite{gedik14}, as we would have to go back and forth between the disk
+blocks to access the edge attributes).%
+
+On the other hand, putting all the edge attributes into the disk blocks
+containing the graph structure is expected to cause significant overhead when
+only a few of these attributes are read. This is somewhat similar to the
+
+% However, not all of the data is
+% accessed by all the queries. Typically, there are correlations among the
+% attributes accessed by different queries, such as Q1 and Q5 accessing
+% attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and so on.%
+
+Typically, a graph database stores entities as vertices,
+relationships as edges, and additional information in labels that may be
+associated with either vertices or edges.
+
+They store entities as vertices,
+relationships as edges, and additional information in labels that may be
+associated with either vertices or edges.
+
 An \emph{interaction graph} is an append-only graph, where new edges and
@@
 
-In an interaction graph database organized as blocks of temporal neighborlists
-(as in Gedik et al.~\cite{gedik14}), it is important to store the edge and
-vertex properties locally. For  instance, if the edge properties are kept away
-in a relational table, there will be almost no benefit to the locality
-optimizations performed for block organization (discussed in Gedik et
-al.~\cite{gedik14}, as we would have to go back and forth between the disk
-blocks to access the edge attributes).%
+% In an interaction graph database organized as blocks of temporal neighbor lists
+% (as in Gedik et al.~\cite{gedik14}), it is important to store the edge and
+% vertex properties locally. For  instance, if the edge properties are kept away
+% in a relational table, there will be almost no benefit to the locality
+% optimizations performed for block organization (discussed in Gedik et
+% al.~\cite{gedik14}, as we would have to go back and forth between the disk
+% blocks to access the edge attributes).%
 

### 2014-08-07 docs/cidr2015/text/main.tex
Moving paper into ACM SIG format.

--- 2014-07-29 docs/paper/text/solution.tex
@@
-\clearpage
-\newpage
+\input{config}
+
+%\documentclass{sig-alternate}
+
+\begin{document}
+%
+% --- Author Metadata here ---
+\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
+%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
+%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
+% --- End of Author Metadata ---
+
+\title{Adaptive Disk Storage for Interaction Graphs}
+
+\numberofauthors{2} %  in this sample file, there are a *total*
+% of EIGHT authors. SIX appear on the 'first-page' (for formatting
+% reasons) and the remaining two appear in the \additionalauthors section.
+%
+\author{
+% You can go ahead and credit any number of authors here,
+% e.g. one 'row of three' or two rows (consisting of one row of three
+% and a second row of one, two or three).
+%
+% The command \alignauthor (no curly braces needed) should
+% precede each author name, affiliation/snail-mail address and
+% e-mail address. Additionally, tag each line of
+% affiliation/address with \affaddr, and tag the
+% e-mail address with \email.
+%
+% 1st. author
+\alignauthor
+Robert Soul\'{e}\\
+       \affaddr{University of Lugano}\\
+       \affaddr{1932 Wallamaloo Lane}\\
+       \affaddr{Wallamaloo, New Zealand}\\
+       \email{robert.soule@usi.ch}
+% 2nd. author
+\alignauthor
+Bu\u{g}ra Gedik\\
+       \affaddr{Bilkent University}\\
+       \affaddr{P.O. Box 1212}\\
+       \affaddr{Dublin, Ohio 43017-6221}\\
+       \email{bgedik@cs.bilkent.edu.tr}
+}
+
+\maketitle
+\begin{abstract}
+TODO
+\end{abstract}
+
+% A category with the (minimum) three required fields
+\category{H.4}{Information Systems Applications}{Miscellaneous}
+%A category including the fourth, optional field follows...
+\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
+
+\terms{Theory}
+
+\keywords{ACM proceedings, \LaTeX, text tagging}
+
+\section{Introduction}
+
+Query workload and disk layout have a significant impact on database
+performance.  For table-based, relational databases, this fact has led database
+designers to develop alternative approaches for storage layout: row-oriented
+storage is more efficient when queries access many attributes from a small
+number of records, and column-oriented storage is more efficient when queries
+access a small number of attributes from many records.
+
+Recently, there has been increased interest in graph databases in general, and
+\emph{interaction graph} databases in particular.  An \emph{interaction graph}
+is an append-only graph, where new edges and vertices are added as time
+progresses.  Thus, interaction graph encode the relationships over time between
+entities in their structure, making them ideal for use in social networking,
+telecommunications, or modeling the world-wide web.
+
+Unfortunately, although interaction graph databases, like relational databases,
+are the target of diverse query workloads, there is no clear correspondence to a
+row-oriented or column-oriented storage layout.
+Typical interaction graphs have data (i.e., \emph{attributes} ) associated with
+their vertices and edges. Most algorithms access this data as they traverse the
+graph.
+
+In an interaction graph database organized as blocks of temporal neighbor lists
+(as in Gedik et al.~\cite{gedik14}), it is important to store the edge and
+vertex attributes locally. For  instance, if the edge properties were kept separately
+in a relational table, there will be almost no benefit to the locality
+optimizations performed for block organization (discussed in Gedik et
+al.~\cite{gedik14}, as we would have to go back and forth between the disk
+blocks to access the edge attributes).%
+
+On the other hand, putting all the edge attributes into the disk blocks
+containing the graph structure is expected to cause significant overhead when
+only a few of these attributes are read. This is somewhat similar to the
+
+% However, not all of the data is
+% accessed by all the queries. Typically, there are correlations among the
+% attributes accessed by different queries, such as Q1 and Q5 accessing
+% attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and so on.%
+
+Typically, a graph database stores entities as vertices,
+relationships as edges, and additional information in labels that may be
+associated with either vertices or edges.
+
+They store entities as vertices,
+relationships as edges, and additional information in labels that may be
+associated with either vertices or edges.
+
+An \emph{interaction graph} is an append-only graph, where new edges and
+vertices are added as time progresses.%
+
+Typical interaction graphs have data associated with their vertices and edges.
+We call this data \emph{attributes} or \emph{properties}. Most algorithms
+access this data as they traverse the graph. However, not all of the data is
+accessed by all the queries. Typically, there are correlations among the
+attributes accessed by different queries, such as Q1 and Q5 accessing
+attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and so on.%
+
+% In an interaction graph database organized as blocks of temporal neighbor lists
+% (as in Gedik et al.~\cite{gedik14}), it is important to store the edge and
+% vertex properties locally. For  instance, if the edge properties are kept away
+% in a relational table, there will be almost no benefit to the locality
+% optimizations performed for block organization (discussed in Gedik et
+% al.~\cite{gedik14}, as we would have to go back and forth between the disk
+% blocks to access the edge attributes).%
+
+On the other hand, putting all the edge attributes into the disk blocks
+containing the graph structure is expected to cause significant overhead when
+only a few of these attributes are read. This is somewhat similar to the 
+problem with row-oriented databases, where the entire row needs to be accessed
+from the disk despite the query needing only a small fraction of it.
+Unfortunately, there is no clear correspondence to a column-oriented database
+layout for the graph databases.%
+
+An important characteristic of interaction graphs are that they are temporal.
+As such, the co-access correlations for the attributes can be different for
+different temporal regions. It might also be unknown at the insertion time and
+may be discovered later depending on the workload. This points to the need for
+adaptively optimizing the layout (somewhat similar to H2O \cite{alagiannis14}
+and HYRISE~\cite{grund10}).%
+
+I imagine a solution to this problem, which I name the `rail layout'. The idea
+is to start with large blocks that contain the entire data. As we learn about
+the access properties for different time regions, we might start splitting
+such blocks into smaller blocks that run parallel to each other, almost like
+having two or more graph databases for certain time regions, each containing a
+ different subset of the attributes, but with a link between them in case a
+query needs to access both.
+
+\section{Problem Description}
+
+The problem is to determine a partitioning of the disk blocks into sub-blocks
+that would minimize query I/O. Of course, partitioning into sub-blocks results
+in an increase in storage cost. The storage cost must stay below some
+specified limit.
+
+There are two possibilities: \emph{overlapping} and \emph{non-overlapping}
+partitioning. In the overlapping case, attributes can be replicated in the
+sub-blocks. The non-overlapping case is a true partition.
+
+We start by defining the cost model, which formalizes the query I/O and storage
+costs.
+
+\subsection{Cost Model}
+Let $Q$ be the query workload, where each query $q\in Q$ accesses a set of
+attributes $q.A$ and traverses parts of the graph for the time range
+$q.T=[q.t_s,q.t_e]$. We denote the set of all attributes as $A$. Given a block
+$B$, we denote its time range as $B.T$, which is the union of the time ranges
+of its temporal neighborlists. Let $s(a)$ denote the size of an attribute $a$.
+We use $c_n(B)$ to denote the number of temporal neighborlists within block
+$B$ and $c_e(B)$ to denote the total number of edges in the temporal
+neighborlists within the block. We overload the notation for block size and
+use $s(B)$ to denote the size of a block $B$. We have: 
+\begin{equation}
+s(B) = c_e(B) \cdot \Big(16 + \sum_{a\in B.A} s(a)\Big) + c_n(B) \cdot 12  
+\end{equation}
+Here, $16$ corresponds to the cost of storing the edge id and the timestamp,
+and $12$ corresponds to the cost of storing the head vertex ($8$ bytes) plus
+the number of entries ($4$ bytes) for a temporal neighborlist.
+
+Our goal is to create a potentially overlapping partitioning of a block,
+denoted by $\mathcal{P}(B)$. We call the partitions \emph{sub-blocks}. We have
+$\bigcup_{B'\in \mathcal{P}(B)} B'.A = A$. We aim to find the function
+$\mathcal{P}$ that minimizes the query I/O over $B$, while keeping the
+relative storage overhead beyond a limit, say $1+\alpha$ times the original.
+If we represent the query I/O as $L(\mathcal{P}, B)$ and the relative storage
+overhead as $H(\mathcal{P}, B)$, our goal is to find:
+\begin{equation}
+\mathcal{P} \leftarrow \mbox{argmin}_{\{\mathcal{P}: H(\mathcal{P}(B)) < \alpha\}} L(\mathcal{P},B)
+\end{equation}
+
+The storage overhead can be formalized as follows for the non-overlapping case:
+\begin{equation}
+H(\mathcal{P}, B) = (|\mathcal{P}(B)|-1)\cdot\Big(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\Big) 
+\end{equation}
+
+For the general case (including overlapping), we can formulate the storage
+overhead as follows:
+\begin{equation}
+H(\mathcal{P}, B) = \frac{\sum_{B'\in \mathcal{P}(B)} s(B')}{s(B)} - 1 
+\end{equation}
+
+%TODO: note about why we have two different cases
+
+\subsubsection{Query I/O}
+
+Let $m$ be a function that maps a query $q$ to the set of sub-blocks that are
+accessed to satisfy it for a relevant block $B$ under a given partitioning
+$\mathcal{P}$. For the case of non-overlapping attributes, we have:
+\begin{equation}
+m(\mathcal{P}, B, q) = \{B': B'\in \mathcal{P}(B) \wedge q.A \cap B'.A \ne \emptyset\}  
+\end{equation}
+
+For the case of overlapping attributes, we use a simple heuristic to define the
+set of sub-blocks to be used for answering the query.
+Algorithm~\ref{alg:greedyM} captures it. The idea is to select sub-blocks that
+bring the highest relative marginal gain, which is defined as the size of
+attribute data that contributes to the query result, relative to the sub-block
+size. While computing the marginal gain, attributes that are covered by
+sub-blocks that are already selected are not considered.
+%
+\begin{algorithm}[ht]
+\scriptsize
+\caption{m-overlapping($\mathcal{P}, B, q$)}
+\label{alg:greedyM}
+\KwData{$\mathcal{P}$: partitioning function, $B$: block, $q$: query}
+$S\leftarrow \emptyset; R\leftarrow \emptyset$ \tcp*{Selected attributes; Resulting sub-blocks}
+\While(\tcp*[f]{While unselected attributes remain}){$S \subset q.A$}{
+  $B' \leftarrow \mbox{argmax}_{B'\in\mathcal{P}(B)\setminus R} \sum_{a\in B'.A \cap q.A \setminus S} \frac{c_e(B') \cdot s(a)}{s(B')}$
+  $S \leftarrow S \cup B'.A$\tcp*{Extend the selected attributes}
+  $R\leftarrow R \cup B'$\tcp*{Extend the selected sub-blocks}
+}
+\Return R \tcp*{Final set of sub-blocks covering the query attributes}
+\end{algorithm} 
+%
+The query I/O cost for a block is then given by:
+\begin{equation}
+L(\mathcal{P}, B) = \sum_{q\in Q} \mathbf{1}(q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} \!\!s(B')
+\end{equation}
+
+%\paragraph{\textbf{Other useful equations}}
+%\noindent The following are unused, but may be needed for the partitioning
+%algorithm.
+%
+%Let $f: \mathbb{R}\times A \rightarrow \mathbb{N}$ be a function that maps a
+%attribute-timestamp pair $(a, t)$ to the frequency of queries that access
+%attribute $a$ for time $t$. More formally:
+%\begin{equation}
+%f(a,t) = \sum_{q\in Q} \textbf{1}(a\in q.A \wedge t\in q.T)
+%\end{equation}
+%
+%Let $T$ represent a time range $[T.s, T.e]$. We denote the frequency of
+%attribute access for the time interval $T$ for attribute $a$ as follows:
+%\begin{equation}
+%f(a, T) = \int_{T.s}^{T.e} f(a, t)\cdot dt =  \sum_{q\in Q} |q.T \cap T| 
+%\end{equation}
+
 \section{Partitioning}
@@
 
-\clearpage
-\newpage
 \subsection{Overlapping Attributes}
@@
  \end{algorithm}
+
+\section{Related Work}
+\section{Conclusion}
+
+%
+% The following two commands are all you need in the
+% initial runs of your .tex file to
+% produce the bibliography for the citations in your paper.
+\bibliographystyle{abbrv}
+\bibliography{main}  % sigproc.bib is the name of the Bibliography in this case
+
+\end{document}

### 2014-08-07 docs/cidr2015/text/main.tex
smoother introduction

@@
 
-\numberofauthors{2} %  in this sample file, there are a *total*
-% of EIGHT authors. SIX appear on the 'first-page' (for formatting
-% reasons) and the remaining two appear in the \additionalauthors section.
-%
+\numberofauthors{2}
+
 \author{
-% You can go ahead and credit any number of authors here,
-% e.g. one 'row of three' or two rows (consisting of one row of three
-% and a second row of one, two or three).
-%
-% The command \alignauthor (no curly braces needed) should
-% precede each author name, affiliation/snail-mail address and
-% e-mail address. Additionally, tag each line of
-% affiliation/address with \affaddr, and tag the
-% e-mail address with \email.
-%
 % 1st. author
@@
        \affaddr{University of Lugano}\\
-       \affaddr{1932 Wallamaloo Lane}\\
-       \affaddr{Wallamaloo, New Zealand}\\
+       \affaddr{Faculty of Informatics}\\
+       \affaddr{Lugano 6904, Switzerland}\\
        \email{robert.soule@usi.ch}
@@
        \affaddr{Bilkent University}\\
-       \affaddr{P.O. Box 1212}\\
-       \affaddr{Dublin, Ohio 43017-6221}\\
+       \affaddr{Department of Computer Engineering}\\
+       \affaddr{Bilkent, Ankara 06800, Turkey}
        \email{bgedik@cs.bilkent.edu.tr}
@@
 
-Query workload and disk layout have a significant impact on database
-performance.  For table-based, relational databases, this fact has led database
-designers to develop alternative approaches for storage layout: row-oriented
-storage is more efficient when queries access many attributes from a small
-number of records, and column-oriented storage is more efficient when queries
-access a small number of attributes from many records.
+Graph databases, and \emph{interaction graph} databases in
+particular, are a critical component for many of today's most popular
+applications.  Graph databases store entities as vertices, relationships as
+edges, and have data associated with either vertices or edges, called
+\emph{attributes} or \emph{properties}.  Interaction graphs differ from general
+graphs in that they are \emph{append-only}, and edges are assigned a
+\emph{temporal value}.  Thus, interaction graphs encode the relationships over
+time between entities in their structure, making them ideal for use in social
+networking, telecommunications, or modeling the world-wide web.
 
-Recently, there has been increased interest in graph databases in general, and
-\emph{interaction graph} databases in particular.  An \emph{interaction graph}
-is an append-only graph, where new edges and vertices are added as time
-progresses.  Thus, interaction graph encode the relationships over time between
-entities in their structure, making them ideal for use in social networking,
-telecommunications, or modeling the world-wide web.
+To query an interaction graph, most algorithms traverse the graph structure to
+access the relevant attributes.  Frequently, there are correlations among the
+attributes accessed by different queries. For example, queries $q_1$ and $q_5$
+might access attributes $a_1$ and $a_2$, while queries $q_2$, $q_3$ and $q_4$
+access attributes $a_3$ and $a_4$. Because interaction graphs are temporal, the
+co-access correlations for the attributes can vary for different
+temporal regions.  Moreover, the co-access correlations might be unknown at the
+insertion time, but be discovered later, when the workload is known.
 
-Unfortunately, although interaction graph databases, like relational databases,
-are the target of diverse query workloads, there is no clear correspondence to a
-row-oriented or column-oriented storage layout.
-Typical interaction graphs have data (i.e., \emph{attributes} ) associated with
-their vertices and edges. Most algorithms access this data as they traverse the
-graph.
+It is widely recognized that query workload and disk layout have a significant
+impact on database performance~\cite{alagiannis14,grund10,stonebraker05}.  For
+table-based, relational databases, this fact has led database designers to
+develop alternative approaches for storage layout: row-oriented storage is more
+efficient when queries access many attributes from a small number of records,
+and column-oriented storage is more efficient when queries access a small number
+of attributes from many records~\cite{stonebraker05}.  Unfortunately, although
+interaction graph databases, like relational databases, are the target of
+diverse query workloads, there is no clear correspondence to a row-oriented or
+column-oriented storage layout.
 
-In an interaction graph database organized as blocks of temporal neighbor lists
-(as in Gedik et al.~\cite{gedik14}), it is important to store the edge and
-vertex attributes locally. For  instance, if the edge properties were kept separately
-in a relational table, there will be almost no benefit to the locality
-optimizations performed for block organization (discussed in Gedik et
-al.~\cite{gedik14}, as we would have to go back and forth between the disk
-blocks to access the edge attributes).%
+Typically, an interaction graph database is organized as blocks of temporal
+neighbor lists.  Attributes can be stored in two ways, either separately (e.g., in a relational
+table), or locally with the temporal neighbor lists.  If they are stored
+separately, then the graph database cannot take advantage of locality
+optimizations performed for block organization.  As discussed in Gedik et
+al.~\cite{gedik14}, the database must go back and forth between the disk blocks
+to access the edge attributes.  On the other hand, if attributes are stored
+locally in the disk blocks containing the graph structure, then there can be
+significant overhead due to disk I/O if only a few attributes are needed to
+answer a query.
 
-On the other hand, putting all the edge attributes into the disk blocks
-containing the graph structure is expected to cause significant overhead when
-only a few of these attributes are read. This is somewhat similar to the
+This paper presents an algorithm for adaptively optimizing disk block storage
+for interaction graphs. The key idea behind the approach is a novel storage
+scheme, called a \emph{rail layout}.  Initially, attributes are stored with
+their neighbor lists in large blocks. Over time, the blocks are split into
+smaller-blocks that contain the neighbor lists, but only subsets of the
+attributes, depending on the query workload. Intuitively, this is almost like
+having two or more graph databases for certain time regions, each containing a
+different subset of the attributes, but with a link between them in case a query
+needs to access both.
 
-% However, not all of the data is
-% accessed by all the queries. Typically, there are correlations among the
-% attributes accessed by different queries, such as Q1 and Q5 accessing
-% attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and so on.%
-
-Typically, a graph database stores entities as vertices,
-relationships as edges, and additional information in labels that may be
-associated with either vertices or edges.
-
-They store entities as vertices,
-relationships as edges, and additional information in labels that may be
-associated with either vertices or edges.
-
-An \emph{interaction graph} is an append-only graph, where new edges and
-vertices are added as time progresses.%
-
-Typical interaction graphs have data associated with their vertices and edges.
-We call this data \emph{attributes} or \emph{properties}. Most algorithms
-access this data as they traverse the graph. However, not all of the data is
-accessed by all the queries. Typically, there are correlations among the
-attributes accessed by different queries, such as Q1 and Q5 accessing
-attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and so on.%
-
-% In an interaction graph database organized as blocks of temporal neighbor lists
-% (as in Gedik et al.~\cite{gedik14}), it is important to store the edge and
-% vertex properties locally. For  instance, if the edge properties are kept away
-% in a relational table, there will be almost no benefit to the locality
-% optimizations performed for block organization (discussed in Gedik et
-% al.~\cite{gedik14}, as we would have to go back and forth between the disk
-% blocks to access the edge attributes).%
-
-On the other hand, putting all the edge attributes into the disk blocks
-containing the graph structure is expected to cause significant overhead when
-only a few of these attributes are read. This is somewhat similar to the 
-problem with row-oriented databases, where the entire row needs to be accessed
-from the disk despite the query needing only a small fraction of it.
-Unfortunately, there is no clear correspondence to a column-oriented database
-layout for the graph databases.%
-
-An important characteristic of interaction graphs are that they are temporal.
-As such, the co-access correlations for the attributes can be different for
-different temporal regions. It might also be unknown at the insertion time and
-may be discovered later depending on the workload. This points to the need for
-adaptively optimizing the layout (somewhat similar to H2O \cite{alagiannis14}
-and HYRISE~\cite{grund10}).%
-
-I imagine a solution to this problem, which I name the `rail layout'. The idea
-is to start with large blocks that contain the entire data. As we learn about
-the access properties for different time regions, we might start splitting
-such blocks into smaller blocks that run parallel to each other, almost like
-having two or more graph databases for certain time regions, each containing a
- different subset of the attributes, but with a link between them in case a
-query needs to access both.
+% This points to the need for
+% adaptively optimizing the layout (somewhat similar to H2O \cite{alagiannis14}
+% and HYRISE~\cite{grund10}).%
 
@@
 
-\subsubsection{Heuristic Solution}
-
-\paragraph*{Approach.$\,$}
-TODO
-
-\begin{algorithm}[ht]
-\scriptsize
-\caption{Algorithm for partitioning blocks into sub-blocks with non-overlapping attributes.}
-\label{alg:overlappingP}
-\KwData{$B$: block, $Q$: set of queries}
-$c^*\leftarrow \infty$ \tcp*{Lowest cost over all \# of partitions}
-\For(\tcp*[f]{For each possible \# of partitions}){$k=1$ to $|A|$}{
-   $R[i]\leftarrow \emptyset, \forall i\in [1..k]$ \tcp*{Initialize partitions}
-   \For(\tcp*[f]{For each attribute}){$a \in A$\textnormal{, in decr.\/ order of }$f(a)$}{
-      $c\leftarrow \infty$ \tcp*{Lowest cost over all assignments}  
-      $j\leftarrow -1$ \tcp*{Best partition assignment}
-      \For(\tcp*[f]{For each partition assignment}){$i\in [1..k]$} {
-         $R[i]\leftarrow R[i] \cup \{a\}$\tcp*{Assign attribute}
-         \If(\tcp*[f]{If query cost is lower}){$L(R, B, Q)<c$}{
-            $c\leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
-            $j\leftarrow i$\tcp*{Update the best partition}
-         }
-         $R[i]\leftarrow R[i] \setminus \{a\}$\tcp*{Un-assign attribute}
-      }
-      $R[j]\leftarrow R[j] \cup \{a\}$\tcp*{Assign to best partition}
-   }
-   \tcp{If solution is feasable and has lower cost}
-   \If{$H(R, B, Q)\geq\alpha \wedge L(R, B, Q)<c^*$}{
-      $c^* \leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
-      $\mathcal{P}(B)\leftarrow R$\tcp*{Update the best partitioning}
-   }
-}
-\Return $\mathcal{P}(B)$ \tcp*{Final set of sub-blocks}
-\end{algorithm}
-
 \subsection{Overlapping Attributes}
@@
 
-\subsubsection{Heuristic Solution}
+\section{Evaluation}
+\section{Related Work}
 
-\paragraph*{Approach.$\,$} We start the algorithm with a 
- partitioning based on what queries we have seen.
-Every query gets its own sub-block. This is the``ideal'' partitioning,
-because the I/O cost would be minimized for every query that we would
-have seen. As input to the algorithm, we provide a function $\dist$, which
-is used to compute the distance between two blocks.
-
-$$
-\dist: block \rightarrow block \rightarrow int
-$$
-
-Note that the $\dist$ function can be chosen arbitrarily.  Examples of useful
-$\dist$ functions are the frequencies with which the queries are accessed, or the Jaccard index, which measures the set-similarity of
-the attributes.
-
-Using this function, the algorithm iteratively combines the two partitions that
-are closest together. Algorithm~\ref{alg:closestBF} shows the pseudocode for a brute-force approach
-  to finding the closest paris that runs in $O(n^2)$ time. However, there is
-  also a divide-and-conquer algorithm~\cite{cormen01} for solving closest pair
-  of points problem in $O(n \log n)$.
-After each combination of partitions, the algorithm calculate the storage cost
-for the partitioning. The algorithm stops when the  storage cost is below some
-specified threshold.  The result is the block partitioning.
-
-\begin{algorithm}[ht]
-\SetKwFunction{closestPair}{closestPair}
-\scriptsize
-\caption{Algorithm for partitioning blocks into sub-blocks with overlapping attributes.}
-\label{alg:closestPair}
-\KwData{$B$: block, $Q$: set of queries}
-$\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}
-
-\For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
- $\mathcal{P}(B) = \mathcal{P}(B) \cup q.A$ 
-}
-
-\While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
-  \text{max}$}{
-$\text{x}, \text{y}  \leftarrow  \closestPair{}  $\;
- $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - \{x, y\}$ \;
-  $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  \cup \{ x \cup y \}$ \;
-}
-\Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
-\end{algorithm}
-
- \begin{algorithm}[ht]
-\SetKwFunction{length}{length}
-\SetKwFunction{dist}{dist}
- \scriptsize
- \caption{Brute-force algorithm for closest points in $O(n^2)$.}
- \label{alg:closestBF}
-$minDist  \gets \infty $ \;
-\For{$i \gets 1$ \textbf{to} $\length(P)-1$} {
-\For{$j \gets i+ 1$ \textbf{to} $\length(P)$} {
-
-  $p \gets P[i]$ \;
-  $q \gets P[j]$ \;
-  \If{$\dist(p,q) < minDist$} {
-    $minDist \gets \dist(p,q)$\;
-    $closestPair \gets (p,q)$\;
-  }
-}
-}
-\Return $closestPair$
-% \KwData{$B$: block, $Q$: set of queries}
-% $\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}
-% \For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
-%  $\mathcal{P}(B) = \mathcal{P}(B) \cup q.A$ 
-% }
-% \While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
-%   \text{max}$}{
-% $\text{x}, \text{y}  \leftarrow  \closestPair{}  $\;
-%  $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - x - y$ \;
-%   $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  + x \cup y$ \;
-% \Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
- \end{algorithm}
-
-\section{Related Work}
+\begin{alltt}\scriptsize
+- H2O \cite{alagiannis14}
+- HYRISE~\cite{grund10}).%
+\end{alltt}
 \section{Conclusion}

### 2014-08-07 docs/workshop/text/abstract.tex
Renamed cidr to workshop

This paper presents an algorithm for adaptively optimizing disk block storage for interaction graphs.  An interaction graph is an append-only, temporal graph structure that encodes the relationships between entities over time. The algorithm makes use of a novel storage scheme, called a rail layout, that partitions disk blocks into smaller blocks that contain the graph structure, but only a subset of attributes. In the ideal case, the database need only read the attributes relevant to answering a particular query. The paper presents an integer linear program formulation to determine an optimal partitioning of attributes for a given query workload. Simulations show that the rail layout scheme significantly reduces latency due to query I/O for interaction graphs.

### 2014-08-14 code/libcommon/include/intergdb/common/Cost.h
Added new classes for partitioning, and computing the io and storage costs

@@ -1,0 +1,20 @@
+#pragma once
+
+#include <intergdb/common/Partitioning.h>
+#include <intergdb/common/QueryWorkload.h>
+
+namespace intergdb { namespace common
+{
+
+class Partitioning;
+class QueryWorkload;
+
+class Cost
+{
+public:
+    Cost() {}
+    int io(Partitioning partitioning, QueryWorkload workload) ;
+    int storage(Partitioning partitioning, QueryWorkload workload);
+};
+
+} }

### 2014-08-14 code/libcommon/include/intergdb/common/Partition.h

@@ -13,7 +13,7 @@ class Partition
 {
 public:
   Partition() {}
-  void addAttribute(Attribute * attribute) { attributes_.add(attribute); }
+  void addAttribute(Attribute * attribute) { attributes_.insert(attribute); }
   void removeAttribute(Attribute * attribute) { attributes_.erase(attribute); }
   std::unordered_set<Attribute *> const & getAttributes() const { return attributes_; }
 private:

### 2014-08-14 code/liboptimizer/src/Solver.cpp
Renamed solver to Solver

@@ -13,7 +13,7 @@
 #include <stdio.h>
 #include <assert.h>
 
-#include <solver.h>
+#include <Solver.h>
 #include <iostream>
 
 #include <intergdb/common/SystemConstants.h>

### 2014-08-14 code/test/OptimizerTest/src/Main.cpp
Renamed solver to Solver

@@ -6,7 +6,7 @@
 #include <intergdb/common/QueryWorkload.h>
 #include <intergdb/common/Query.h>
 
-#include <solver.h>
+#include <Solver.h>
 
 using namespace std;
 using namespace intergdb::common;

### 2014-08-14 code/liboptimizer/include/HeuristicNonOverlapping.h
Dummy implementations for other solvers

@@ -1,0 +1,18 @@
+#pragma once
+
+#include <intergdb/common/QueryWorkload.h>
+#include <intergdb/common/Partitioning.h>
+
+#include <Solver.h>
+
+using namespace intergdb::common;
+
+class HeuristicNonOverlapping : public Solver
+{
+public:
+    HeuristicNonOverlapping() { }
+    ~HeuristicNonOverlapping() { }
+    int solve(QueryWorkload * workload, Partitioning * partitioning);
+};
+
+

### 2014-08-14 code/liboptimizer/include/HeuristicOverlapping.h

@@ -1,0 +1,18 @@
+#pragma once
+
+#include <intergdb/common/QueryWorkload.h>
+#include <intergdb/common/Partitioning.h>
+
+#include <Solver.h>
+
+using namespace intergdb::common;
+
+class HeuristicOverlapping : public Solver
+{
+public:
+    HeuristicOverlapping() { }
+    ~HeuristicOverlapping() { }
+    int solve(QueryWorkload * workload, Partitioning * partitioning);
+};
+
+

### 2014-08-14 code/liboptimizer/include/PartitionPerAttribute.h

@@ -1,0 +1,18 @@
+#pragma once
+
+#include <intergdb/common/QueryWorkload.h>
+#include <intergdb/common/Partitioning.h>
+
+#include <Solver.h>
+
+using namespace intergdb::common;
+
+class PartitionPerAttribute : public Solver
+{
+public:
+    PartitionPerAttribute() { }
+    ~PartitionPerAttribute() { }
+    int solve(QueryWorkload * workload, Partitioning * partitioning);
+};
+
+

### 2014-08-14 code/liboptimizer/src/PartitionPerAttribute.cpp

@@ -1,0 +1,15 @@
+
+#include <assert.h>
+#include <iostream>
+
+#include <PartitionPerAttribute.h>
+
+#include <intergdb/common/SystemConstants.h>
+
+using namespace std;
+
+int PartitionPerAttribute::solve(QueryWorkload * workload, Partitioning * partitioning) 
+{
+    assert(false);
+    return 0;
+}

### 2014-08-14 code/liboptimizer/src/SolverFactory.cpp

@@ -2,7 +2,10 @@
 #include <SolverFactory.h>
 #include <OptimalOverlapping.h>
 #include <OptimalNonOverlapping.h>
-
+#include <SinglePartition.h>
+#include <PartitionPerAttribute.h>
+#include <HeuristicOverlapping.h>
+#include <HeuristicNonOverlapping.h>
 
 SolverFactory* SolverFactory::factory_ = NULL;
 
@@ -26,22 +29,22 @@ Solver * SolverFactory::makeOptimalNonOverlapping()
 
 Solver * SolverFactory::makeHeuristicOverlapping()
 {
-    return NULL;
+    return new HeuristicOverlapping();
 }
 
 Solver * SolverFactory:: makeHeuristicNonOverlapping()
 {
-    return NULL;
+    return new HeuristicNonOverlapping();
 }
 
 Solver * SolverFactory::makeSinglePartition()
 {
-    return NULL;
+    return new SinglePartition();
 }
 
 Solver * SolverFactory::makePartitionPerAttribute()
 {
-    return NULL;
+    return new PartitionPerAttribute();
 }
 
 

### 2014-08-14 code/liboptimizer/include/SinglePartition.h
Dummy implementations for other solvers

@@ -1,0 +1,18 @@
+#pragma once
+
+#include <intergdb/common/QueryWorkload.h>
+#include <intergdb/common/Partitioning.h>
+
+#include <Solver.h>
+
+using namespace intergdb::common;
+
+class SinglePartition : public Solver
+{
+public:
+    SinglePartition() { }
+    ~SinglePartition() { }
+    int solve(QueryWorkload * workload, Partitioning * partitioning);
+};
+
+

### 2014-08-14 code/liboptimizer/src/HeuristicNonOverlapping.cpp

@@ -1,0 +1,15 @@
+
+#include <assert.h>
+#include <iostream>
+
+#include <HeuristicNonOverlapping.h>
+
+#include <intergdb/common/SystemConstants.h>
+
+using namespace std;
+
+int HeuristicNonOverlapping::solve(QueryWorkload * workload, Partitioning * partitioning) 
+{
+    assert(false);
+    return 0;
+}

### 2014-08-14 code/liboptimizer/src/HeuristicOverlapping.cpp

@@ -1,0 +1,15 @@
+
+#include <assert.h>
+#include <iostream>
+
+#include <HeuristicOverlapping.h>
+
+#include <intergdb/common/SystemConstants.h>
+
+using namespace std;
+
+int HeuristicOverlapping::solve(QueryWorkload * workload, Partitioning * partitioning) 
+{
+    assert(false);
+    return 0;
+}

### 2014-08-14 code/liboptimizer/src/SinglePartition.cpp

@@ -1,0 +1,15 @@
+
+#include <assert.h>
+#include <iostream>
+
+#include <SinglePartition.h>
+
+#include <intergdb/common/SystemConstants.h>
+
+using namespace std;
+
+int SinglePartition::solve(QueryWorkload * workload, Partitioning * partitioning) 
+{
+    assert(false);
+    return 0;
+}

### 2014-08-14 code/liboptimizer/include/OptimalCommon.h
Solver interface change to return a partitioning.

@@ -43,7 +43,7 @@ class OptimalCommon : public Solver
 public:
     OptimalCommon() {}
     virtual ~OptimalCommon() {}
-    int solve(QueryWorkload * workload, Partitioning * partitioning);
+    Partitioning * solve(QueryWorkload * workload);
 
 protected:
     virtual int constraints(var_env *e, gurobi_ctx *ctx, QueryWorkload * workload) = 0;   

### 2014-08-14 code/liboptimizer/include/SinglePartition.h

@@ -12,7 +12,7 @@ class SinglePartition : public Solver
 public:
     SinglePartition() { }
     ~SinglePartition() { }
-    int solve(QueryWorkload * workload, Partitioning * partitioning);
+    Partitioning * solve(QueryWorkload * workload);
 };
 
 

### 2014-08-14 code/liboptimizer/src/HeuristicOverlapping.cpp

@@ -8,8 +8,8 @@
 
 using namespace std;
 
-int HeuristicOverlapping::solve(QueryWorkload * workload, Partitioning * partitioning) 
+Partitioning * HeuristicOverlapping::solve(QueryWorkload * workload) 
 {
     assert(false);
-    return 0;
+    return NULL;
 }

### 2014-08-14 code/liboptimizer/src/PartitionPerAttribute.cpp

@@ -8,8 +8,8 @@
 
 using namespace std;
 
-int PartitionPerAttribute::solve(QueryWorkload * workload, Partitioning * partitioning) 
+Partitioning * PartitionPerAttribute::solve(QueryWorkload * workload) 
 {
     assert(false);
-    return 0;
+    return NULL;
 }

### 2014-08-14 code/liboptimizer/include/HeuristicNonOverlapping.h
Solver interface change to return a partitioning.

@@ -12,7 +12,7 @@ class HeuristicNonOverlapping : public Solver
 public:
     HeuristicNonOverlapping() { }
     ~HeuristicNonOverlapping() { }
-    int solve(QueryWorkload * workload, Partitioning * partitioning);
+    Partitioning * solve(QueryWorkload * workload);
 };
 
 

### 2014-08-14 code/liboptimizer/include/HeuristicOverlapping.h

@@ -12,7 +12,7 @@ class HeuristicOverlapping : public Solver
 public:
     HeuristicOverlapping() { }
     ~HeuristicOverlapping() { }
-    int solve(QueryWorkload * workload, Partitioning * partitioning);
+    Partitioning * solve(QueryWorkload * workload);
 };
 
 

### 2014-08-14 code/liboptimizer/include/PartitionPerAttribute.h

@@ -12,7 +12,7 @@ class PartitionPerAttribute : public Solver
 public:
     PartitionPerAttribute() { }
     ~PartitionPerAttribute() { }
-    int solve(QueryWorkload * workload, Partitioning * partitioning);
+    Partitioning * solve(QueryWorkload * workload);
 };
 
 

### 2014-08-14 code/liboptimizer/include/Solver.h

@@ -9,7 +9,7 @@ class Solver
 {
 public:
     virtual ~Solver() { }
-    virtual int solve(QueryWorkload * workload, Partitioning * partitioning) = 0;  
+    virtual Partitioning * solve(QueryWorkload * workload) = 0;  
 };
 
 

### 2014-08-14 code/liboptimizer/src/HeuristicNonOverlapping.cpp

@@ -8,8 +8,8 @@
 
 using namespace std;
 
-int HeuristicNonOverlapping::solve(QueryWorkload * workload, Partitioning * partitioning) 
+Partitioning * HeuristicNonOverlapping::solve(QueryWorkload * workload) 
 {
     assert(false);
-    return 0;
+    return NULL;
 }

### 2014-08-14 code/liboptimizer/src/OptimalCommon.cpp

@@ -236,7 +236,7 @@ void OptimalCommon::cleanup(var_env *e, gurobi_ctx *ctx)
 }
 
 
-int OptimalCommon::solve(QueryWorkload * workload, Partitioning * partitioning) 
+Partitioning * OptimalCommon::solve(QueryWorkload * workload)
 {
     int error = 0;
     gurobi_ctx ctx;
@@ -281,6 +281,6 @@ QUIT:
     }
 
     cleanup(&e, &ctx);
-    return 0;
+    return NULL;
 }
 

### 2014-08-14 code/liboptimizer/src/SinglePartition.cpp

@@ -8,8 +8,8 @@
 
 using namespace std;
 
-int SinglePartition::solve(QueryWorkload * workload, Partitioning * partitioning) 
+Partitioning * SinglePartition::solve(QueryWorkload * workload) 
 {
     assert(false);
-    return 0;
+    return NULL;
 }

### 2014-08-14 code/test/OptimizerTest/src/Main.cpp

@@ -36,16 +36,16 @@ int main()
     cerr << "This is a test program for the solver." << endl;
 
     QueryWorkload workload; 
-    Partitioning partitioning; 
 
-    Solver * s = SolverFactory::instance()->makeOptimalOverlapping();    
+    Solver * solver = SolverFactory::instance()->makeOptimalOverlapping();    
 
     nov_ex1(&workload);
 
     cerr << workload.toString() << endl;
 
-    s->solve(&workload, 0);
-    delete(s);
+    Partitioning * partitioning = solver->solve(&workload);
+    delete(solver);
+    delete(partitioning);
     return 0;    
 }
  

### 2014-08-14 code/liboptimizer/src/SinglePartition.cpp
Implementations of the easy partitionings

@@ -10,6 +10,11 @@ using namespace std;
 
 Partitioning * SinglePartition::solve(QueryWorkload * workload) 
 {
-    assert(false);
-    return NULL;
+    Partition * partition = new Partition();
+    for (auto & attribute : workload->getAttributes()) {
+        partition->addAttribute(&attribute);
+    }
+    Partitioning * partitioning = new Partitioning();
+    partitioning->addPartition(partition);
+    return partitioning;
 }

### 2014-08-14 code/liboptimizer/src/PartitionPerAttribute.cpp
Implementations of the easy partitionings

@@ -10,6 +10,13 @@ using namespace std;
 
 Partitioning * PartitionPerAttribute::solve(QueryWorkload * workload) 
 {
-    assert(false);
-    return NULL;
+    Partitioning * partitioning = new Partitioning();
+    Partition * partition;
+    for (auto & attribute : workload->getAttributes()) {
+        partition = new Partition();
+        partition->addAttribute(&attribute);
+        partitioning->addPartition(partition);
+    }  
+    return partitioning;
 }
+   

### 2014-08-14 docs/paper/text/evaluation.tex
merged the text from the workshop. fixed bibtex bugs

\section{Evaluation}

\subsection{I/O comparison}
\begin{alltt}\scriptsize
optimal vs single partition vs. attribute per partition
Compute the I/O for different workloads,
3 values averaged at the end
Also compute the storage...
\end{alltt}

\subsection{LP Solver scalability}
\begin{alltt}\scriptsize
# of attrbutes vs solving time:
\end{alltt}

### 2014-08-15 docs/paper/text/evaluation.tex
list of experiments

@@
-\section{Evaluation}
+\begin{verbatim}
+1) RunningTimeVsNumAttributes:
+  x axis: # attributes
+  y axis: running time
+  series: optimal-nov, optimal-ov, heuristic-nov, heuristic-ov
 
-\subsection{I/O comparison}
-\begin{alltt}\scriptsize
-optimal vs single partition vs. attribute per partition
-Compute the I/O for different workloads,
-3 values averaged at the end
-Also compute the storage...
-\end{alltt}
+2) RunningTimeVsNumQueryKinds:
+  x axis: num query kinds
+  y axis: running time
+  series: optimal-nov, optimal-ov, heuristic-nov, heuristic-ov
 
-\subsection{LP Solver scalability}
-\begin{alltt}\scriptsize
-# of attrbutes vs solving time:
-\end{alltt}
+3) QueryIOVsNumAttributes:
+  x axis: # attributes
+  y axis: query IO
+  series: optimal-nov, optimal-ov, heuristic-nov, heuristic-ov, single-block, per-attribute
+
+4) StorageOverheadVsNumAttributes:
+  x axis: # attributes
+  y axis: storage overhead
+  series: optimal-nov, optimal-ov, heuristic-nov, heuristic-ov, single-block, per-attribute
+
+5) QueryIOVsStorageOverheadThreshold:
+  x axis: query overhead threshold
+  y axis: query IO
+  series: optimal-nov, optimal-ov, heuristic-nov, heuristic-ov, single-block, per-attribute
+
+6) StorageOverheadVsStorageOverheadThreshold:
+  x axis: query overhead threshold
+  y axis: storage overhead
+  series: optimal-nov, optimal-ov, heuristic-nov, heuristic-ov, single-block, per-attribute
+
+7) QueryIOVsNumQueryKinds:
+  x axis: num query kinds
+  y axis: query IO
+  series: optimal-nov, optimal-ov, heuristic-nov, heuristic-ov, single-block, per-attribute
+
+8) StorageOverheadVsNumQueryKinds:
+  x axis: num query kinds
+  y axis: storage overhead
+  series: optimal-nov, optimal-ov, heuristic-nov, heuristic-ov, single-block, per-attribute
+
+9) QueryIOVsAttributeSizeSkew:
+  x axis: attribute size skew
+  y axis: query IO
+  series: optimal-nov, optimal-ov, heuristic-nov, heuristic-ov, single-block, per-attribute
+
+10) StorageOverheadVsAttributeSizeSkew:
+  x axis: attribute size skew
+  y axis: storage overhead
+  series: optimal-nov, optimal-ov, heuristic-nov, heuristic-ov, single-block, per-attribute
+
+11) QueryIOVsQueryLength:
+  x axis: mean query length
+  y axis: query IO
+  series: optimal-nov, optimal-ov, heuristic-nov, heuristic-ov, single-block, per-attribute
+
+12) StorageOverheadVsQueryLength:
+  x axis: mean query length
+  y axis: storage overhead
+  series: optimal-nov, optimal-ov, heuristic-nov, heuristic-ov, single-block, per-attribute
+
+13) QueryIOVsQueryFreqSkew:
+  x axis: query freq skew
+  y axis: query IO
+  series: optimal-nov, optimal-ov, heuristic-nov, heuristic-ov, single-block, per-attribute
+
+14) StorageOverheadVsQueryFreqSkew:
+  x axis: query freq skew
+  y axis: storage overhead
+  series: optimal-nov, optimal-ov, heuristic-nov, heuristic-ov, single-block, per-attribute
+\end{verbatim}

### 2014-08-15 code/liboptimizer/include/solver.h
list of experiments

@@ -2,7 +2,6 @@
 
 #include <intergdb/common/QueryWorkload.h>
 
-
 using namespace intergdb::common;
 
 #define EDGE_ID_SIZE 8

### 2014-08-15 docs/paper/text/introduction.tex

--- 2014-07-20 docs/paper/text/introduction.tex
@@
-\section{Introduction}
-An \emph{interaction graph} is an append-only graph, where new edges and
-vertices are added as time progresses.%
+Graph databases, and \emph{interaction graph} databases in
+particular, are a critical component for many of today's most popular
+applications.  Graph databases store entities as vertices, relationships as
+edges, and have data associated with either vertices or edges, called
+\emph{attributes} or \emph{properties}.  Interaction graphs differ from general
+graphs in that they are \emph{append-only}, and edges are assigned a
+\emph{temporal value}.  Thus, interaction graphs encode the relationships over
+time between entities in their structure, making them ideal for use in social
+networking, telecommunications, or modeling the world-wide web.
 
-Typical interaction graphs have data associated with their vertices and edges.
-We call this data \emph{attributes} or \emph{properties}. Most algorithms
-access this data as they traverse the graph. However, not all of the data is
-accessed by all the queries. Typically, there are correlations among the
-attributes accessed by different queries, such as Q1 and Q5 accessing
-attributes a and b, and Q2, Q3, Q4 accessing attributes c and d, and so on.%
+To query an interaction graph, most algorithms traverse the graph structure to
+access the relevant attributes.  Frequently, there are correlations among the
+attributes accessed by different queries. For example, queries $q_1$ and $q_5$
+might access attributes $a_1$ and $a_2$, while queries $q_2$, $q_3$ and $q_4$
+access attributes $a_3$ and $a_4$. Because interaction graphs are temporal, the
+co-access correlations for the attributes can vary for different
+temporal regions.  Moreover, the co-access correlations might be unknown at the
+insertion time, but be discovered later, when the workload is known.
 
-In an interaction graph database organized as blocks of temporal neighborlists
-(as in Gedik et al.~\cite{gedik14}), it is important to store the edge and
-vertex properties locally. For  instance, if the edge properties are kept away
-in a relational table, there will be almost no benefit to the locality
-optimizations performed for block organization (discussed in Gedik et
-al.~\cite{gedik14}, as we would have to go back and forth between the disk
-blocks to access the edge attributes).%
+It is widely recognized that query workload and disk layout have a significant
+impact on database performance~\cite{alagiannis14,grund10,stonebraker05}.  For
+table-based, relational databases, this fact has led database designers to
+develop alternative approaches for storage layout: row-oriented storage is more
+efficient when queries access many attributes from a small number of records,
+and column-oriented storage is more efficient when queries access a small number
+of attributes from many records~\cite{stonebraker05}.  Unfortunately, although
+interaction graph databases, like relational databases, are the target of
+diverse query workloads, there is no clear correspondence to a row-oriented or
+column-oriented storage layout.
 
-On the other hand, putting all the edge attributes into the disk blocks
-containing the graph structure is expected to cause significant overhead when
-only a few of these attributes are read. This is somewhat similar to the 
-problem with row-oriented databases, where the entire row needs to be accessed
-from the disk despite the query needing only a small fraction of it.
-Unfortunately, there is no clear correspondence to a column-oriented database
-layout for the graph databases.%
+Typically, an interaction graph database is organized as blocks of temporal
+neighbor lists.  Attributes can be stored in two ways, either separately (e.g., in a relational
+table), or locally with the temporal neighbor lists.  If they are stored
+separately, then the graph database cannot take advantage of locality
+optimizations performed for block organization.  As discussed in Gedik et
+al.~\cite{gedik14}, the database must go back and forth between the disk blocks
+to access the edge attributes.  On the other hand, if attributes are stored
+locally in the disk blocks containing the graph structure, then there can be
+significant overhead due to disk I/O if only a few attributes are needed to
+answer a query.
 
-An important characteristic of interaction graphs are that they are temporal.
-As such, the co-access correlations for the attributes can be different for
-different temporal regions. It might also be unknown at the insertion time and
-may be discovered later depending on the workload. This points to the need for
-adaptively optimizing the layout (somewhat similar to H2O \cite{alagiannis14}
-and HYRISE~\cite{grund10}).%
+This paper presents an algorithm for adaptively optimizing disk block storage
+for interaction graphs. The key idea behind the approach is a novel storage
+scheme, called a \emph{rail layout}.  Initially, attributes are stored with
+their neighbor lists in large blocks. Over time, the blocks are split into
+smaller-blocks that contain the neighbor lists, but only subsets of the
+attributes, depending on the query workload. Intuitively, this is almost like
+having two or more graph databases for certain time regions, each containing a
+different subset of the attributes, but with a link between them in case a query
+needs to access both.
 
-I imagine a solution to this problem, which I name the `rail layout'. The idea
-is to start with large blocks that contain the entire data. As we learn about
-the access properties for different time regions, we might start splitting
-such blocks into smaller blocks that run parallel to each other, almost like
-having two or more graph databases for certain time regions, each containing a
- different subset of the attributes, but with a link between them in case a
-query needs to access both.
+% This points to the need for
+% adaptively optimizing the layout (somewhat similar to H2O \cite{alagiannis14}
+% and HYRISE~\cite{grund10}).%

### 2014-08-15 code/liboptimizer/include/HeuristicNonOverlapping.h
fix namespace

@@ -5,14 +5,14 @@
 
 #include <Solver.h>
 
-using namespace intergdb::common;
+namespace intergdb { namespace optimizer {
 
 class HeuristicNonOverlapping : public Solver
 {
 public:
     HeuristicNonOverlapping() { }
     ~HeuristicNonOverlapping() { }
-    Partitioning * solve(QueryWorkload * workload);
+    intergdb::common::Partitioning * solve(intergdb::common::QueryWorkload * workload);
 };
 
-
+} }

### 2014-08-15 code/liboptimizer/include/OptimalCommon.h

@@ -12,7 +12,7 @@ extern "C" {
 }
 #endif
 
-using namespace intergdb::common;
+namespace intergdb { namespace optimizer {
 
 typedef struct var_env {
     int num_vars;
@@ -43,10 +43,10 @@ class OptimalCommon : public Solver
 public:
     OptimalCommon() {}
     virtual ~OptimalCommon() {}
-    Partitioning * solve(QueryWorkload * workload);
+    intergdb::common::Partitioning * solve(intergdb::common::QueryWorkload * workload);
 
 protected:
-    virtual int constraints(var_env *e, gurobi_ctx *ctx, QueryWorkload * workload) = 0;   
+    virtual int constraints(var_env *e, gurobi_ctx *ctx, intergdb::common::QueryWorkload * workload) = 0;   
 
     int x(var_env *e, int a, int p);
     int y(var_env *e, int p, int q);
@@ -54,18 +54,18 @@ protected:
     int u(var_env *e, int p);
     int c_e();
     int c_n();
-    int s(std::vector<Attribute> const & attributes);
+    int s(std::vector<intergdb::common::Attribute> const & attributes);
     double K();
     double alpha();
-    int accesses(std::vector<Query> const & queries, int q, int a);   
+    int accesses(std::vector<intergdb::common::Query> const & queries, int q, int a);   
     void name_variables(var_env *e, char** vname);
-    void create_env(var_env *e, QueryWorkload * workload);
+    void create_env(var_env *e, intergdb::common::QueryWorkload * workload);
     void init_ctx(var_env *e, gurobi_ctx* ctx);
     void variables(var_env *e, gurobi_ctx *ctx);   
-    void objective(var_env *e, gurobi_ctx *ctx, QueryWorkload * workload);
+    void objective(var_env *e, gurobi_ctx *ctx, intergdb::common::QueryWorkload * workload);
     int solve_model(var_env *e, gurobi_ctx *ctx);
     void cleanup(var_env *e, gurobi_ctx *ctx);
 
 };
 
-
+} }

### 2014-08-15 code/liboptimizer/include/OptimalOverlapping.h

@@ -5,10 +5,10 @@
 #include <Solver.h>
 #include <OptimalCommon.h>
 
-using namespace intergdb::common;
-
 struct var_env;
 struct gurobi_ctx;
+
+namespace intergdb { namespace optimizer {
 
 class OptimalOverlapping : public OptimalCommon
 {
@@ -17,8 +17,8 @@ public:
     ~OptimalOverlapping() {}
 
 private:
-    int constraints(var_env *e, gurobi_ctx *ctx, QueryWorkload * workload) ;
+    int constraints(var_env *e, gurobi_ctx *ctx, intergdb::common::QueryWorkload * workload) ;
 
 };
 
-
+} }

### 2014-08-15 code/liboptimizer/include/SinglePartition.h

@@ -5,14 +5,14 @@
 
 #include <Solver.h>
 
-using namespace intergdb::common;
+namespace intergdb { namespace optimizer {
 
 class SinglePartition : public Solver
 {
 public:
     SinglePartition() { }
-    ~SinglePartition() { }
-    Partitioning * solve(QueryWorkload * workload);
+    ~SinglePartition() { } 
+    intergdb::common::Partitioning * solve(intergdb::common::QueryWorkload * workload);
 };
 
-
+} }

### 2014-08-15 code/liboptimizer/include/solver.h

@@ -1,38 +1,0 @@
-#pragma once
-
-#include <intergdb/common/QueryWorkload.h>
-
-using namespace intergdb::common;
-
-#define EDGE_ID_SIZE 8
-#define TIMESTAMP_SIZE 8
-#define HEAD_VERTEX_SIZE 8
-#define NUM_ENTRIES 4 
-
-struct var_env;
-struct gurobi_ctx;
-
-class Solver
-{
-public:
-    Solver() {}
-    int solve_nov(QueryWorkload * workload);
-    int solve_ov(QueryWorkload * workload);
-
-private:
-    int x(var_env *e, int a, int p);
-    int y(var_env *e, int p, int q);
-    int z(var_env *e, int a, int p, int q);
-    int u(var_env *e, int p);
-    int c_e();
-    int c_n();
-    int s(std::vector<Attribute> const & attributes);
-    double K();
-    double alpha();
-    int accesses(std::vector<Query> const & queries, int q, int a);   
-    void name_variables(var_env *e, char** vname);
-    void create_env(var_env *e, QueryWorkload * workload);
-    void init_ctx(var_env *e, gurobi_ctx* ctx);
-};
-
-

### 2014-08-15 code/liboptimizer/src/OptimalCommon.cpp

@@ -8,6 +8,8 @@
 #include <intergdb/common/SystemConstants.h>
 
 using namespace std;
+using namespace intergdb::common;
+using namespace intergdb::optimizer;
 
 int OptimalCommon::x(var_env *e, int a, int p)
 {

### 2014-08-15 code/liboptimizer/src/OptimalNonOverlapping.cpp

@@ -16,6 +16,8 @@ extern "C" {
 #endif
 
 using namespace std;
+using namespace intergdb::common;
+using namespace intergdb::optimizer;
 
 int OptimalNonOverlapping::constraints(var_env *e, gurobi_ctx *ctx, QueryWorkload * workload)
 {

### 2014-08-15 code/liboptimizer/src/OptimalOverlapping.cpp

@@ -16,6 +16,8 @@ extern "C" {
 #endif
 
 using namespace std;
+using namespace intergdb::common;
+using namespace intergdb::optimizer;
 
 int OptimalOverlapping::constraints(var_env *e, gurobi_ctx *ctx, QueryWorkload * workload)
 {

### 2014-08-15 code/liboptimizer/src/SolverFactory.cpp

@@ -7,14 +7,14 @@
 #include <HeuristicOverlapping.h>
 #include <HeuristicNonOverlapping.h>
 
-SolverFactory* SolverFactory::factory_ = NULL;
+using namespace std;
+using namespace intergdb::optimizer;
 
-SolverFactory * SolverFactory::instance() 
+unique_ptr<SolverFactory> SolverFactory::factory_(new SolverFactory());
+
+SolverFactory & SolverFactory::instance() 
 {
-    if (factory_ == NULL) {
-        factory_ = new SolverFactory();
-    }
-    return factory_;
+    return *factory_;
 }
 
 Solver * SolverFactory::makeOptimalOverlapping()

### 2014-08-15 code/liboptimizer/include/HeuristicOverlapping.h
fix namespace

@@ -5,14 +5,15 @@
 
 #include <Solver.h>
 
-using namespace intergdb::common;
+namespace intergdb { namespace optimizer {
+
 
 class HeuristicOverlapping : public Solver
 {
 public:
     HeuristicOverlapping() { }
     ~HeuristicOverlapping() { }
-    Partitioning * solve(QueryWorkload * workload);
+    intergdb::common::Partitioning * solve(intergdb::common::QueryWorkload * workload);
 };
 
-
+} }

### 2014-08-15 code/liboptimizer/include/OptimalNonOverlapping.h

@@ -5,10 +5,10 @@
 #include <Solver.h>
 #include <OptimalCommon.h>
 
-using namespace intergdb::common;
-
 struct var_env;
 struct gurobi_ctx;
+
+namespace intergdb { namespace optimizer {
 
 class OptimalNonOverlapping : public OptimalCommon
 {
@@ -17,8 +17,10 @@ public:
     ~OptimalNonOverlapping() {}
 
 private:
-    int constraints(var_env *e, gurobi_ctx *ctx, QueryWorkload * workload) ;
+    int constraints(var_env *e, gurobi_ctx *ctx, intergdb::common::QueryWorkload * workload) ;
 
 };
+
+} }
 
 

### 2014-08-15 code/liboptimizer/include/PartitionPerAttribute.h

@@ -5,14 +5,14 @@
 
 #include <Solver.h>
 
-using namespace intergdb::common;
+namespace intergdb { namespace optimizer {
 
 class PartitionPerAttribute : public Solver
 {
 public:
     PartitionPerAttribute() { }
     ~PartitionPerAttribute() { }
-    Partitioning * solve(QueryWorkload * workload);
+    intergdb::common::Partitioning * solve(intergdb::common::QueryWorkload * workload);
 };
 
-
+} }

### 2014-08-15 code/liboptimizer/include/Solver.h

@@ -3,13 +3,16 @@
 #include <intergdb/common/QueryWorkload.h>
 #include <intergdb/common/Partitioning.h>
 
-using namespace intergdb::common;
+
+namespace intergdb { namespace optimizer {
 
 class Solver
 {
 public:
     virtual ~Solver() { }
-    virtual Partitioning * solve(QueryWorkload * workload) = 0;  
+    // TODO: I think this should take a const & and return a non-pointer
+    virtual intergdb::common::Partitioning * solve(intergdb::common::QueryWorkload * workload) = 0;  
 };
 
+} }
 

### 2014-08-15 code/liboptimizer/include/SolverFactory.h

@@ -2,11 +2,15 @@
 
 #include <Solver.h>
 
+#include <memory>
+
+namespace intergdb { namespace optimizer {
+
 class SolverFactory
 {
 public:
-    static SolverFactory * instance();
-    Solver * makeOptimalOverlapping();
+    static SolverFactory & instance(); // TODO: this better returns a reference
+    Solver * makeOptimalOverlapping(); // TODO: These could all return references
     Solver * makeOptimalNonOverlapping();
     Solver * makeHeuristicOverlapping();
     Solver * makeHeuristicNonOverlapping();
@@ -15,6 +19,7 @@ public:
 
 private:
     SolverFactory() {}
-    static SolverFactory* factory_;
+    static std::unique_ptr<SolverFactory> factory_;
 };
 
+} }

### 2014-08-15 code/liboptimizer/src/HeuristicNonOverlapping.cpp

@@ -7,6 +7,8 @@
 #include <intergdb/common/SystemConstants.h>
 
 using namespace std;
+using namespace intergdb::common;
+using namespace intergdb::optimizer;
 
 Partitioning * HeuristicNonOverlapping::solve(QueryWorkload * workload) 
 {

### 2014-08-15 code/liboptimizer/src/HeuristicOverlapping.cpp

@@ -7,6 +7,8 @@
 #include <intergdb/common/SystemConstants.h>
 
 using namespace std;
+using namespace intergdb::common;
+using namespace intergdb::optimizer;
 
 Partitioning * HeuristicOverlapping::solve(QueryWorkload * workload) 
 {

### 2014-08-15 code/liboptimizer/src/PartitionPerAttribute.cpp

@@ -7,6 +7,8 @@
 #include <intergdb/common/SystemConstants.h>
 
 using namespace std;
+using namespace intergdb::common;
+using namespace intergdb::optimizer;
 
 Partitioning * PartitionPerAttribute::solve(QueryWorkload * workload) 
 {

### 2014-08-15 code/liboptimizer/src/SinglePartition.cpp

@@ -7,6 +7,8 @@
 #include <intergdb/common/SystemConstants.h>
 
 using namespace std;
+using namespace intergdb::common;
+using namespace intergdb::optimizer;
 
 Partitioning * SinglePartition::solve(QueryWorkload * workload) 
 {

### 2014-08-15 code/test/OptimizerTest/src/Main.cpp
fix namespaces

@@ -12,7 +12,7 @@
 
 using namespace std;
 using namespace intergdb::common;
-
+using namespace intergdb::optimizer;
 
 // nov-ex1.lp 
 void nov_ex1(QueryWorkload * workload)
@@ -37,15 +37,15 @@ int main()
 
     QueryWorkload workload; 
 
-    Solver * solver = SolverFactory::instance()->makeOptimalOverlapping();    
+    Solver * solver = SolverFactory::instance().makeOptimalOverlapping();    
 
     nov_ex1(&workload);
 
     cerr << workload.toString() << endl;
 
     Partitioning * partitioning = solver->solve(&workload);
-    delete(solver);
-    delete(partitioning);
+    delete solver;
+    delete partitioning;
     return 0;    
 }
  

### 2014-08-15 code/liboptimizer/src/OptimalNonOverlapping.cpp
fixed u constraints

@@ -1,11 +1,13 @@
 #include <stdlib.h>
 #include <stdio.h>
 #include <assert.h>
-
+// <algorithms>
+#include <numeric>
 #include <OptimalNonOverlapping.h>
 #include <iostream>
 
 #include <intergdb/common/SystemConstants.h>
+#include <intergdb/common/Cost.h>
 
 #ifdef __cplusplus
 extern "C" {
@@ -122,14 +124,27 @@ int OptimalNonOverlapping::constraints(var_env *e, gurobi_ctx *ctx, QueryWorkloa
         if (error) return error;    
     }
 
+    
+    auto attributes = workload->getAttributes();
+    Cost cost;
+    Partition part; 
+
+    for_each(attributes.begin(), attributes.end(), [&] (Attribute const & attr) {
+            part.addAttribute(&attr);            
+        });
+    double limit = 1 + (alpha() / 1 - ((s(workload->getAttributes()) * c_e()) / cost.getPartitionSize(part)));
+
     /* Sixth set of constraints */    
     j = 0;
     for (int p = 0; p < e->P; ++p) {
-        ctx->ind[j] = u(e,p);
-        ctx->val[j] = -1.0;
+        int index = u(e,p);
+        cout << "INDEX: " << index << endl;
+        ctx->ind[j] = index;
+        ctx->val[j] = 1;
+        j++;
     }
     error = GRBaddconstr(ctx->model, e->P, ctx->ind, ctx->val, 
-                         GRB_LESS_EQUAL, alpha(), NULL);
+                         GRB_LESS_EQUAL, limit, NULL);
     if (error) return error;
 
     error = GRBupdatemodel(ctx->model);

### 2014-08-15 code/liboptimizer/src/OptimalCommon.cpp
fixed objective function

@@ -148,7 +148,7 @@ void OptimalCommon::variables(var_env *e, gurobi_ctx *ctx)
 /* objective coefficients for each of the variables */
 void OptimalCommon::objective(var_env *e, gurobi_ctx *ctx, QueryWorkload const * workload) 
 {
-    int sa = s(workload->getAttributes());
+    //int sa = s(workload->getAttributes());
 
     for (int a = 0; a < e->A; ++a) {
         for (int p = 0; p < e->P; ++p) {
@@ -167,7 +167,7 @@ void OptimalCommon::objective(var_env *e, gurobi_ctx *ctx, QueryWorkload const *
     for (int a = 0; a < e->A; ++a) {
         for (int p = 0; p < e->P; ++p) {
             for (int q = 0; q < e->Q; ++q) {
-                ctx->obj[z(e,a,p,q)] = sa * c_e() ;
+                ctx->obj[z(e,a,p,q)] = workload->getAttribute(a).getSize() * c_e() ;
             }
         }
     }

### 2014-08-15 code/liboptimizer/src/OptimalNonOverlapping.cpp
parenthesizing bug in limit for u's

@@ -132,13 +132,14 @@ int OptimalNonOverlapping::constraints(var_env *e, gurobi_ctx *ctx, QueryWorkloa
     for_each(attributes.begin(), attributes.end(), [&] (Attribute const & attr) {
             part.addAttribute(&attr);            
         });
-    double limit = 1 + (alpha() / 1 - ((s(workload->getAttributes()) * c_e()) / cost.getPartitionSize(part)));
+    double limit = 1 + (alpha() / (1 - ((s(workload->getAttributes()) * c_e()) / cost.getPartitionSize(part))));
+
+
 
     /* Sixth set of constraints */    
     j = 0;
     for (int p = 0; p < e->P; ++p) {
         int index = u(e,p);
-        cout << "INDEX: " << index << endl;
         ctx->ind[j] = index;
         ctx->val[j] = 1;
         j++;

### 2014-08-15 code/liboptimizer/src/OptimalCommon.cpp
parenthesizing bug in limit for u's

@@ -148,8 +148,6 @@ void OptimalCommon::variables(var_env *e, gurobi_ctx *ctx)
 /* objective coefficients for each of the variables */
 void OptimalCommon::objective(var_env *e, gurobi_ctx *ctx, QueryWorkload const * workload) 
 {
-    //int sa = s(workload->getAttributes());
-
     for (int a = 0; a < e->A; ++a) {
         for (int p = 0; p < e->P; ++p) {
             ctx->obj[x(e,a,p)] = 0;

### 2014-08-15 code/test/OptimizerTest/src/Main.cpp
only the ov_exp2 test

@@ -259,10 +259,10 @@ int main()
     bool success = true;
 
     //success &= test(nov_ex1, nov_ex1_verify, solverNov, "nov_ex1");
-    success &= test(nov_ex2, nov_ex2_verify, solverNov, "nov_ex2");
+    //success &= test(nov_ex2, nov_ex2_verify, solverNov, "nov_ex2");
     //success &= test(nov_ex3, nov_ex3_verify, solverNov, "nov_ex3");
     //success &= test(ov_ex1, ov_ex1_verify, solverOv, "ov_ex1");
-    //success &= test(ov_ex2, ov_ex2_verify, solverOv, "ov_ex2");
+    success &= test(ov_ex2, ov_ex2_verify, solverOv, "ov_ex2");
 
 
     if (success)

### 2014-08-15 code/liboptimizer/src/OptimalNonOverlapping.cpp
Gixed the constraints on the overlapping case

@@ -1,7 +1,6 @@
 #include <stdlib.h>
 #include <stdio.h>
 #include <assert.h>
-// <algorithms>
 #include <numeric>
 #include <OptimalNonOverlapping.h>
 #include <iostream>
@@ -128,12 +127,10 @@ int OptimalNonOverlapping::constraints(var_env *e, gurobi_ctx *ctx, QueryWorkloa
     auto attributes = workload->getAttributes();
     Cost cost;
     Partition part; 
-
     for_each(attributes.begin(), attributes.end(), [&] (Attribute const & attr) {
             part.addAttribute(&attr);            
         });
     double limit = 1 + (alpha() / (1 - ((s(workload->getAttributes()) * c_e()) / cost.getPartitionSize(part))));
-
 
 
     /* Sixth set of constraints */    

### 2014-08-15 code/liboptimizer/src/OptimalOverlapping.cpp

@@ -6,6 +6,7 @@
 #include <iostream>
 
 #include <intergdb/common/SystemConstants.h>
+#include <intergdb/common/Cost.h>
 
 #ifdef __cplusplus
 extern "C" {
@@ -156,14 +157,37 @@ int OptimalOverlapping::constraints(var_env *e, gurobi_ctx *ctx, QueryWorkload c
     }
 
 
+
+    auto attributes = workload->getAttributes();
+    Cost cost;
+    Partition part; 
+    for_each(attributes.begin(), attributes.end(), [&] (Attribute const & attr) {
+            part.addAttribute(&attr);            
+        });
+
+    double limit =  cost.getPartitionSize(part) * (1 + alpha());
+    double u_const =
+        (SystemConstants::edgeIdSize + SystemConstants::timestampSize) * c_e() 
+        + (SystemConstants::headVertexSize + SystemConstants::numEntriesSize) * c_n();
+
     /* 9th set of constraints */    
     j = 0;
     for (int p = 0; p < e->P; ++p) {
         ctx->ind[j] = u(e,p);
-        ctx->val[j] = -1.0;
+        ctx->val[j] = u_const;
+        j++;
     }
-    error = GRBaddconstr(ctx->model, e->P, ctx->ind, ctx->val, 
-                         GRB_LESS_EQUAL, alpha(), NULL);
+
+    for (int p = 0; p < e->P; ++p) {   
+        for (int a = 0; a < e->A; ++a) {
+            ctx->ind[j] = x(e,a,p);
+            ctx->val[j] = workload->getAttribute(a).getSize() * c_n();
+            j++;
+        }
+    }
+
+    error = GRBaddconstr(ctx->model, e->P + e->A, ctx->ind, ctx->val, 
+                         GRB_LESS_EQUAL, limit, NULL);
     if (error) return error;
 
 

### 2014-08-15 code/test/OptimizerTest/src/Main.cpp
run all tests. this fails, probably because of some cleanup issue.

@@ -258,12 +258,11 @@ int main()
 
     bool success = true;
 
-    //success &= test(nov_ex1, nov_ex1_verify, solverNov, "nov_ex1");
-    //success &= test(nov_ex2, nov_ex2_verify, solverNov, "nov_ex2");
-    //success &= test(nov_ex3, nov_ex3_verify, solverNov, "nov_ex3");
-    //success &= test(ov_ex1, ov_ex1_verify, solverOv, "ov_ex1");
+    success &= test(nov_ex1, nov_ex1_verify, solverNov, "nov_ex1");
+    success &= test(nov_ex2, nov_ex2_verify, solverNov, "nov_ex2");
+    success &= test(nov_ex3, nov_ex3_verify, solverNov, "nov_ex3");
+    success &= test(ov_ex1, ov_ex1_verify, solverOv, "ov_ex1");
     success &= test(ov_ex2, ov_ex2_verify, solverOv, "ov_ex2");
-
 
     if (success)
         return EXIT_SUCCESS;    

### 2014-08-15 code/test/OptimizerTest/src/Main.cpp
optimizer tests

@@ -259,11 +259,10 @@ int main()
     bool success = true;
 
     //success &= test(nov_ex1, nov_ex1_verify, solverNov, "nov_ex1");
-    //success &= test(nov_ex2, nov_ex2_verify, solverNov, "nov_ex2");
+    success &= test(nov_ex2, nov_ex2_verify, solverNov, "nov_ex2");
     //success &= test(nov_ex3, nov_ex3_verify, solverNov, "nov_ex3");
-    //success &= test(ov_ex1, ov_ex1_verify, solverOv, "ov_ex1");
-    success &= test(ov_ex2, ov_ex2_verify, solverOv, "ov_ex2");
-
+    // &= test(ov_ex1, ov_ex1_verify, solverOv, "ov_ex1"); 
+    //success &= test(ov_ex2, ov_ex2_verify, solverOv, "ov_ex2");
 
     if (success)
         return EXIT_SUCCESS;    

### 2014-08-15 code/liboptimizer/src/OptimalNonOverlapping.cpp
fix constant miscalculations in hand crafted optimization examples

@@ -130,8 +130,7 @@ int OptimalNonOverlapping::constraints(var_env *e, gurobi_ctx *ctx, QueryWorkloa
     for_each(attributes.begin(), attributes.end(), [&] (Attribute const & attr) {
             part.addAttribute(&attr);            
         });
-    double limit = 1 + (alpha() / (1 - ((s(workload->getAttributes()) * c_e()) / cost.getPartitionSize(part))));
-
+    double limit = 1 + alpha() / (1 - (s(workload->getAttributes()) * c_e()) / cost.getPartitionSize(part));
 
     /* Sixth set of constraints */    
     j = 0;

### 2014-08-15 code/liboptimizer/src/OptimalNonOverlapping.cpp
fix minor comment error

@@ -56,7 +56,7 @@ int OptimalNonOverlapping::constraints(var_env *e, gurobi_ctx *ctx, QueryWorkloa
         }
     }
 
-    /* Second set of constraints */    
+    /* Third set of constraints */    
     for (int p = 0; p < e->P; ++p) {
         for (int q = 0; q < e->Q; ++q) {
             j = 0;
@@ -74,7 +74,7 @@ int OptimalNonOverlapping::constraints(var_env *e, gurobi_ctx *ctx, QueryWorkloa
         }
     }
 
-    /* Third set of constraints */    
+    /* Fourth set of constraints */    
     for (int a = 0; a < e->A; ++a) {
         for (int p = 0; p < e->P; ++p) {
             for (int q = 0; q < e->Q; ++q) {
@@ -91,7 +91,7 @@ int OptimalNonOverlapping::constraints(var_env *e, gurobi_ctx *ctx, QueryWorkloa
         }
     }
     
-    /* Fourth set of constraints */    
+    /* Fifth set of constraints */    
     for (int p = 0; p < e->P; ++p) {
         j = 0;
         for (int a = 0; a < e->A; ++a) {
@@ -107,7 +107,7 @@ int OptimalNonOverlapping::constraints(var_env *e, gurobi_ctx *ctx, QueryWorkloa
 
     }
 
-    /* Fifth set of constraints */    
+    /* Sixth set of constraints */    
     for (int p = 0; p < e->P; ++p) {
         j = 0;
         ctx->ind[j] = u(e,p);
@@ -130,9 +130,10 @@ int OptimalNonOverlapping::constraints(var_env *e, gurobi_ctx *ctx, QueryWorkloa
     for_each(attributes.begin(), attributes.end(), [&] (Attribute const & attr) {
             part.addAttribute(&attr);            
         });
+    cerr << 
     double limit = 1 + alpha() / (1 - (s(workload->getAttributes()) * c_e()) / cost.getPartitionSize(part));
 
-    /* Sixth set of constraints */    
+    /* 6th set of constraints */    
     j = 0;
     for (int p = 0; p < e->P; ++p) {
         int index = u(e,p);

### 2014-08-15 code/test/OptimizerTest/src/Main.cpp
optimizer tests

@@ -257,11 +257,11 @@ int main()
     auto solverNov = SolverFactory::instance().makeOptimalNonOverlapping();    
 
     bool success = true;
-    //success &= test(nov_ex1, nov_ex1_verify, solverNov, "nov_ex1");
+    success &= test(nov_ex1, nov_ex1_verify, solverNov, "nov_ex1");
     success &= test(nov_ex2, nov_ex2_verify, solverNov, "nov_ex2");
-    //success &= test(nov_ex3, nov_ex3_verify, solverNov, "nov_ex3");
-    // &= test(ov_ex1, ov_ex1_verify, solverOv, "ov_ex1"); 
-    //success &= test(ov_ex2, ov_ex2_verify, solverOv, "ov_ex2");
+    success &= test(nov_ex3, nov_ex3_verify, solverNov, "nov_ex3");
+    success &= test(ov_ex1, ov_ex1_verify, solverOv, "ov_ex1"); 
+    success &= test(ov_ex2, ov_ex2_verify, solverOv, "ov_ex2");
 
     if (success)
         return EXIT_SUCCESS;    

### 2014-08-15 code/liboptimizer/src/OptimalCommon.cpp
optimizer free problem

@@ -203,7 +203,8 @@ int OptimalCommon::solve_model(var_env *e, gurobi_ctx *ctx)
     error = GRBgetdblattrarray(ctx->model, GRB_DBL_ATTR_X, 0, e->num_vars, ctx->sol);
     if (error) return error;
 
-    printf("\nOptimization complete\n");
+    // printf("\nOptimization complete\n");
+    
     if (ctx->optimstatus == GRB_OPTIMAL) {
         error = 0;
     } else if (ctx->optimstatus == GRB_INF_OR_UNBD) {

### 2014-08-15 code/test/OptimizerTest/src/Main.cpp

@@ -253,13 +253,13 @@ bool test(TestFunct testFunc, VerifyFunct verifyFunc, Solver solver, string cons
 int main()
 {
     cout << "This is a test program for the solver." << endl;
-    auto solverOv = SolverFactory::instance().makeOptimalOverlapping();    
     auto solverNov = SolverFactory::instance().makeOptimalNonOverlapping();    
-
     bool success = true;
     success &= test(nov_ex1, nov_ex1_verify, solverNov, "nov_ex1");
     success &= test(nov_ex2, nov_ex2_verify, solverNov, "nov_ex2");
     success &= test(nov_ex3, nov_ex3_verify, solverNov, "nov_ex3");
+
+    auto solverOv = SolverFactory::instance().makeOptimalOverlapping();    
     success &= test(ov_ex1, ov_ex1_verify, solverOv, "ov_ex1"); 
     success &= test(ov_ex2, ov_ex2_verify, solverOv, "ov_ex2");
 

### 2014-08-15 code/liboptimizer/include/OptimalCommon.h
optimizer free problem

@@ -14,7 +14,10 @@ extern "C" {
 
 namespace intergdb { namespace optimizer {
 
-typedef struct var_env {
+class OptimalCommon : public Solver
+{
+private:
+  typedef struct var_env {
     int num_vars;
     int x_offset;
     int y_offset;
@@ -23,9 +26,9 @@ typedef struct var_env {
     int A;
     int P;
     int Q;
-} var_env;
+  } var_env;
 
-typedef struct gurobi_ctx {
+  typedef struct gurobi_ctx {
     GRBenv   *env;
     GRBmodel *model;
     double    *sol;
@@ -36,10 +39,8 @@ typedef struct gurobi_ctx {
     char      **vname;
     int       optimstatus;
     double    objval;
-} gurobi_ctx;
-
-class OptimalCommon : public Solver
-{
+  } gurobi_ctx;
+  
 public:
     OptimalCommon() {}
     virtual ~OptimalCommon() {}

### 2014-08-24 code/liboptimizer/src/intergdb/optimizer/OptimalOverlapping.cpp
fix storage overhead bugush

@@ -49,7 +49,7 @@ int OptimalOverlapping::constraints(var_env *e, gurobi_ctx *ctx, QueryWorkload c
                 ctx->val[j] = 1.0;
                 j++;
             }
-           error = GRBaddconstr(ctx->model, e->P, ctx->ind, ctx->val, 
+            error = GRBaddconstr(ctx->model, e->P, ctx->ind, ctx->val, 
                                  GRB_GREATER_EQUAL, accesses(queries,q,a), NULL);
             if (error) return error;
 
@@ -186,7 +186,7 @@ int OptimalOverlapping::constraints(var_env *e, gurobi_ctx *ctx, QueryWorkload c
         }
     }
 
-    error = GRBaddconstr(ctx->model, e->P + e->A, ctx->ind, ctx->val, 
+    error = GRBaddconstr(ctx->model, e->P + e->P*e->A, ctx->ind, ctx->val, 
                          GRB_LESS_EQUAL, limit, NULL);
     if (error) return error;
 

### 2014-08-24 code/liboptimizer/src/intergdb/optimizer/HeuristicNonOverlapping.cpp
minor changes

@@ -20,7 +20,7 @@ Partitioning HeuristicNonOverlapping::solve(QueryWorkload const & workload, doub
     Partitioning partitioning = solve(workload, storageThreshold, numPartitions);
     double overhead = costModel.getStorageOverhead(partitioning, workload);
     if (overhead > storageThreshold)
-      continue;
+      break;
     double cost = costModel.getIOCost(partitioning, workload);
     minCostPart.push(partitioning, cost);
   }

### 2014-09-01 code/simulation/viz/CommonViz.py
new experiment files

@@ -1,0 +1,27 @@
+
+import re
+from collections import OrderedDict
+
+def parseData(dirn, fname):
+  xs = []
+  ysPerSolver = OrderedDict()
+  ydevsPerSolver = OrderedDict()
+  with open(dirn+"/"+fname+".dat") as fin:
+    lines = fin.readlines()
+    for line in lines:
+      if line.startswith("#"):
+        continue
+      (solver, x, y, ydev, nline) = re.split("[\t]", line)
+      x = int(x)
+      y = float(y)
+      ydev = float(ydev)  
+      if solver in ysPerSolver:
+        ysPerSolver[solver].append(y)
+        ydevsPerSolver[solver].append(ydev)
+      else:
+        ysPerSolver[solver] = [y]
+        ydevsPerSolver[solver] = [ydev]
+      if len(xs) == 0 or x > xs[-1]:
+        xs.append(x)
+    return (xs, ysPerSolver, ydevsPerSolver)
+  

### 2014-09-01 code/simulation/viz/QueryIOVsQueryFreqSkew.py

@@ -1,0 +1,36 @@
+import CommonConf
+import CommonViz
+
+import re
+from collections import OrderedDict
+import numpy as np
+import matplotlib.pyplot as pp
+
+EXPERIMENT_NAME = "QueryIOVsQueryFreqSkew"
+X_LABEL         = "Query Freq Skew"
+Y_LABEL         = "I/O Cost (bytes)"
+            
+def main(dirn, fname): 
+  (xs, ysPerSolver, ydevsPerSolver) = CommonViz.parseData(dirn, fname)
+     
+  CommonConf.setupMPPDefaults()
+  fmts = CommonConf.getLineFormats()
+  fig = pp.figure()
+  ax = fig.add_subplot(111)
+  ax.set_xscale("log", basex=2)
+
+  index = 0
+  for (solver, ys), (solver, ydevs) in zip(ysPerSolver.iteritems(),ydevsPerSolver.iteritems()) : 
+    ax.errorbar(xs, ys, yerr=ydevs, label=solver, marker=fmts[index][0], linestyle=fmts[index][1])
+    index = index + 1
+
+  ax.set_xlabel(X_LABEL);
+  ax.set_ylabel(Y_LABEL);
+  ax.legend(loc='best', fancybox=True)
+
+  pp.savefig(dirn+"/"+fname+".pdf")
+  pp.show()
+
+if __name__ == "__main__":
+  main("expData", EXPERIMENT_NAME)
+

### 2014-09-01 code/simulation/viz/RunningTimeVsAttributeSizeSkew.py

@@ -1,0 +1,37 @@
+import CommonConf
+import CommonViz
+
+import re
+from collections import OrderedDict
+import numpy as np
+import matplotlib.pyplot as pp
+
+EXPERIMENT_NAME = "RunningTimeVsAttributeSizeSkew"
+X_LABEL         = "Attribute Size Skew"
+Y_LABEL         = "Time (sec.)"
+
+            
+def main(dirn, fname): 
+  (xs, ysPerSolver, ydevsPerSolver) = CommonViz.parseData(dirn, fname)
+     
+  CommonConf.setupMPPDefaults()
+  fmts = CommonConf.getLineFormats()
+  fig = pp.figure()
+  ax = fig.add_subplot(111)
+  ax.set_xscale("log", basex=2)
+
+  index = 0
+  for (solver, ys), (solver, ydevs) in zip(ysPerSolver.iteritems(),ydevsPerSolver.iteritems()) : 
+    ax.errorbar(xs, ys, yerr=ydevs, label=solver, marker=fmts[index][0], linestyle=fmts[index][1])
+    index = index + 1
+
+  ax.set_xlabel(X_LABEL);
+  ax.set_ylabel(Y_LABEL);
+  ax.legend(loc='best', fancybox=True)
+
+  pp.savefig(dirn+"/"+fname+".pdf")
+  pp.show()
+
+if __name__ == "__main__":
+  main("expData", EXPERIMENT_NAME)
+

### 2014-09-01 code/simulation/viz/RunningTimeVsQueryFreqSkew.py
new experiment files

@@ -1,0 +1,37 @@
+import CommonConf
+import CommonViz
+
+import re
+from collections import OrderedDict
+import numpy as np
+import matplotlib.pyplot as pp
+
+EXPERIMENT_NAME = "RunningTimeVsQueryFreqSkew"
+X_LABEL         = "Query Freq Skew"
+Y_LABEL         = "Time (sec.)"
+
+            
+def main(dirn, fname): 
+  (xs, ysPerSolver, ydevsPerSolver) = CommonViz.parseData(dirn, fname)
+     
+  CommonConf.setupMPPDefaults()
+  fmts = CommonConf.getLineFormats()
+  fig = pp.figure()
+  ax = fig.add_subplot(111)
+  ax.set_xscale("log", basex=2)
+
+  index = 0
+  for (solver, ys), (solver, ydevs) in zip(ysPerSolver.iteritems(),ydevsPerSolver.iteritems()) : 
+    ax.errorbar(xs, ys, yerr=ydevs, label=solver, marker=fmts[index][0], linestyle=fmts[index][1])
+    index = index + 1
+
+  ax.set_xlabel(X_LABEL);
+  ax.set_ylabel(Y_LABEL);
+  ax.legend(loc='best', fancybox=True)
+
+  pp.savefig(dirn+"/"+fname+".pdf")
+  pp.show()
+
+if __name__ == "__main__":
+  main("expData", EXPERIMENT_NAME)
+

### 2014-09-01 code/simulation/viz/StorageOverheadVsAttributeSizeSkew.py

@@ -1,0 +1,37 @@
+import CommonConf
+import CommonViz
+
+import re
+from collections import OrderedDict
+import numpy as np
+import matplotlib.pyplot as pp
+
+EXPERIMENT_NAME = "StorageOverheadVsAttributeSizeSkew"
+X_LABEL         = "Attribute Size Skew"
+Y_LABEL         = "Storage Overhead"
+
+            
+def main(dirn, fname): 
+  (xs, ysPerSolver, ydevsPerSolver) = CommonViz.parseData(dirn, fname)
+     
+  CommonConf.setupMPPDefaults()
+  fmts = CommonConf.getLineFormats()
+  fig = pp.figure()
+  ax = fig.add_subplot(111)
+  ax.set_xscale("log", basex=2)
+
+  index = 0
+  for (solver, ys), (solver, ydevs) in zip(ysPerSolver.iteritems(),ydevsPerSolver.iteritems()) : 
+    ax.errorbar(xs, ys, yerr=ydevs, label=solver, marker=fmts[index][0], linestyle=fmts[index][1])
+    index = index + 1
+
+  ax.set_xlabel(X_LABEL);
+  ax.set_ylabel(Y_LABEL);
+  ax.legend(loc='best', fancybox=True)
+
+  pp.savefig(dirn+"/"+fname+".pdf")
+  pp.show()
+
+if __name__ == "__main__":
+  main("expData", EXPERIMENT_NAME)
+

### 2014-09-01 code/simulation/src/intergdb/simulation/Main.cpp
Refactored older experiments to use the new common code.

@@ -8,12 +8,12 @@ using namespace intergdb::simulation;
 
 int main()
 {
-    //(VsStorageOverheadThreshold()).run();
-    //(VsNumAttributes()).run();
-    //(VsNumQueryKinds()).run();
+    (VsStorageOverheadThreshold()).run();
+    (VsNumAttributes()).run();
+    (VsNumQueryKinds()).run();
     (VsQueryLength()).run();
-    //(VsAttributeSizeSkew()).run();
-    //(VsQueryFreqSkew()).run();
+    (VsAttributeSizeSkew()).run();
+    (VsQueryFreqSkew()).run();
     return EXIT_SUCCESS;
 }
 

### 2014-09-01 code/simulation/viz/QueryIOVsNumQueryKinds.py

@@ -1,54 +1,36 @@
 import CommonConf
+import CommonViz
 
 import re
 from collections import OrderedDict
 import numpy as np
 import matplotlib.pyplot as pp
 
-def main(dirn, fname):
-  attributeSizes = []
-  ioPerSolver = OrderedDict()
-
-  with open(dirn+"/"+fname+".dat") as fin:
-    lines = fin.readlines()
-
-    for line in lines:
-      if line.startswith("#"):
-        continue
-      (solver, attributes, io, deviation, nline) = re.split("[\t]", line)
-      attributes = int(attributes)
-      deviation = int(deviation)
-      io = float(io)
-      if solver in ioPerSolver:
-        ioPerSolver[solver].append((io,deviation))
-      else:
-        ioPerSolver[solver] = [(io,deviation)]
-      if len(attributeSizes) == 0 or attributes > attributeSizes[-1]:
-        attributeSizes.append(attributes)
-
+EXPERIMENT_NAME = "QueryIOVsNumQueryKinds"
+X_LABEL         = "Num Query Kinds"
+Y_LABEL         = "I/O Cost (bytes)"
+            
+def main(dirn, fname): 
+  (xs, ysPerSolver, ydevsPerSolver) = CommonViz.parseData(dirn, fname)
+     
   CommonConf.setupMPPDefaults()
   fmts = CommonConf.getLineFormats()
   fig = pp.figure()
   ax = fig.add_subplot(111)
   ax.set_xscale("log", basex=2)
-    
+
   index = 0
+  for (solver, ys), (solver, ydevs) in zip(ysPerSolver.iteritems(),ydevsPerSolver.iteritems()) : 
+    ax.errorbar(xs, ys, yerr=ydevs, label=solver, marker=fmts[index][0], linestyle=fmts[index][1])
+    index = index + 1
 
-  for solver, ioAndDeviation in ioPerSolver.iteritems():
-    ioAndDeviationLists = map(list, zip(*ioAndDeviation))
-    ax.errorbar(attributeSizes, ioAndDeviationLists[0],
-                yerr=ioAndDeviationLists[1], label=solver,
-                marker=fmts[index][0], linestyle=fmts[index][1])
-    index = index + 1  
-
-  ax.set_xlabel('Number of Query Kinds');
-  ax.set_ylabel('I/O Cost (bytes)');
-  # ax.set_xlim(0, 2100)
[…]

### 2014-09-01 code/simulation/viz/QueryIOVsStorageOverheadThreshold.py

@@ -1,61 +1,36 @@
 import CommonConf
+import CommonViz
 
 import re
 from collections import OrderedDict
 import numpy as np
 import matplotlib.pyplot as pp
 
-def main(dirn, fname):
-  attributeSizes = []
-  ioPerSolver = OrderedDict()
-
-  with open(dirn+"/"+fname+".dat") as fin:
-    lines = fin.readlines()
-
-    for line in lines:
-      if line.startswith("#"):
-        continue
-        
-      (solver, attributes, io, deviation, nline) = re.split("[\t]", line)
-
-      if 'n/a' in io:
-        io = 0
-        deviation = 0
-      else:
-        io = float(io)
-        deviation = float(deviation)
-
-      attributes = float(attributes)
-
-      if solver in ioPerSolver:
-        ioPerSolver[solver].append((io,deviation))
-      else:
-        ioPerSolver[solver] = [(io,deviation)]
-      if len(attributeSizes) == 0 or attributes > attributeSizes[-1]:
-        attributeSizes.append(attributes)
-
+EXPERIMENT_NAME = "QueryIOVsStorageOverheadThreshold"
+X_LABEL         = "Storage Overhead Threshold"
+Y_LABEL         = "I/O Cost (bytes)"
+            
+def main(dirn, fname): 
+  (xs, ysPerSolver, ydevsPerSolver) = CommonViz.parseData(dirn, fname)
+     
   CommonConf.setupMPPDefaults()
   fmts = CommonConf.getLineFormats()
   fig = pp.figure()
   ax = fig.add_subplot(111)
-      
+  ax.set_xscale("log", basex=2)
+
   index = 0
+  for (solver, ys), (solver, ydevs) in zip(ysPerSolver.iteritems(),ydevsPerSolver.iteritems()) : 
+    ax.errorbar(xs, ys, yerr=ydevs, label=solver, marker=fmts[index][0], linestyle=fmts[index][1])
+    index = index + 1
 
-  for solver, ioAndDeviation in ioPerSolver.iteritems():
-    ioAndDeviationLists = map(list, zip(*ioAndDeviation))
-    ax.errorbar(attributeSizes, ioAndDeviationLists[0],
-                yerr=ioAndDeviationLists[1], label=solver,
-                marker=fmts[index][0], linestyle=fmts[index][1])
-    index = index + 1  
-
[…]

### 2014-09-01 code/simulation/viz/RunningTimeVsNumQueryKinds.py

@@ -1,55 +1,37 @@
 import CommonConf
+import CommonViz
 
 import re
 from collections import OrderedDict
 import numpy as np
 import matplotlib.pyplot as pp
 
-def main(dirn, fname):
-  queryKindSizes = []
-  timePerSolver = OrderedDict()
+EXPERIMENT_NAME = "RunningTimeVsNumQueryKinds"
+X_LABEL         = "Num Query Kinds"
+Y_LABEL         = "Time (sec.)"
 
-  with open(dirn+"/"+fname+".dat") as fin:
-    lines = fin.readlines()
-
-    for line in lines:
-      if line.startswith("#"):
-        continue
-      (solver, queryKinds, time, deviation, nline) = re.split("[\t]", line)
-
-      queryKinds = int(queryKinds)
-      time = float(time)
-      deviation = float(deviation)
-
-      if solver in timePerSolver:
-        timePerSolver[solver].append((time, deviation))
-      else:
-        timePerSolver[solver] = [(time, deviation)]
-      if len(queryKindSizes) == 0 or queryKinds > queryKindSizes[-1]:
-        queryKindSizes.append(queryKinds)
-
+            
+def main(dirn, fname): 
+  (xs, ysPerSolver, ydevsPerSolver) = CommonViz.parseData(dirn, fname)
+     
   CommonConf.setupMPPDefaults()
   fmts = CommonConf.getLineFormats()
   fig = pp.figure()
   ax = fig.add_subplot(111)
   ax.set_xscale("log", basex=2)
-    
+
   index = 0
-  for solver, timesAndDeviation in timePerSolver.iteritems():
-    timesAndDeviationLists = map(list, zip(*timesAndDeviation))
-    ax.errorbar(queryKindSizes, timesAndDeviationLists[0],
-                yerr=timesAndDeviationLists[1], label=solver,
-                marker=fmts[index][0], linestyle=fmts[index][1])
+  for (solver, ys), (solver, ydevs) in zip(ysPerSolver.iteritems(),ydevsPerSolver.iteritems()) : 
+    ax.errorbar(xs, ys, yerr=ydevs, label=solver, marker=fmts[index][0], linestyle=fmts[index][1])
     index = index + 1
 
-  ax.set_xlabel('Number of QueryKinds');
-  ax.set_ylabel('Time (sec.)');
-  # ax.set_xlim(0, 2100)
[…]

### 2014-09-01 code/simulation/viz/RunningTimeVsStorageOverheadThreshold.py

@@ -1,59 +1,37 @@
 import CommonConf
+import CommonViz
 
 import re
 from collections import OrderedDict
 import numpy as np
 import matplotlib.pyplot as pp
 
-def main(dirn, fname):
-  storageOverheadThresholdSizes = []
-  timePerSolver = OrderedDict()
+EXPERIMENT_NAME = "RunningTimeVsStorageOverheadThreshold"
+X_LABEL         = "Storage Overhead Threshold"
+Y_LABEL         = "Time (sec.)"
 
-  with open(dirn+"/"+fname+".dat") as fin:
-    lines = fin.readlines()
-
-    for line in lines:
-      if line.startswith("#"):
-        continue
-      (solver, storageOverheadThresholds, time, deviation, nline) = re.split("[\t]", line)
-
-      storageOverheadThresholds = float(storageOverheadThresholds)
-      if 'n/a' in time:
-        time = 0
-        deviation = 0
-      else:
-        time = float(time)
-        deviation = float(deviation)
-
-      if solver in timePerSolver:
-        timePerSolver[solver].append((time, deviation))
-      else:
-        timePerSolver[solver] = [(time, deviation)]
-      if len(storageOverheadThresholdSizes) == 0 or storageOverheadThresholds > storageOverheadThresholdSizes[-1]:
-        storageOverheadThresholdSizes.append(storageOverheadThresholds)
-
+            
+def main(dirn, fname): 
+  (xs, ysPerSolver, ydevsPerSolver) = CommonViz.parseData(dirn, fname)
+     
   CommonConf.setupMPPDefaults()
   fmts = CommonConf.getLineFormats()
   fig = pp.figure()
   ax = fig.add_subplot(111)
-  # ax.set_xscale("log", basex=2)
-    
+  ax.set_xscale("log", basex=2)
+
   index = 0
-  for solver, timesAndDeviation in timePerSolver.iteritems():
-    timesAndDeviationLists = map(list, zip(*timesAndDeviation))
-    ax.errorbar(storageOverheadThresholdSizes, timesAndDeviationLists[0],
-                yerr=timesAndDeviationLists[1], label=solver,
-                marker=fmts[index][0], linestyle=fmts[index][1])
+  for (solver, ys), (solver, ydevs) in zip(ysPerSolver.iteritems(),ydevsPerSolver.iteritems()) : 
[…]

### 2014-09-01 code/simulation/viz/StorageOverheadVsNumAttributes.py

@@ -1,54 +1,37 @@
 import CommonConf
+import CommonViz
 
 import re
 from collections import OrderedDict
 import numpy as np
 import matplotlib.pyplot as pp
 
-def main(dirn, fname):
-  attributeSizes = []
-  storagePerSolver = OrderedDict()
+EXPERIMENT_NAME = "StorageOverheadVsNumAttributes"
+X_LABEL         = "Num Attributes"
+Y_LABEL         = "Storage Overhead"
 
-  with open(dirn+"/"+fname+".dat") as fin:
-    lines = fin.readlines()
-
-    for line in lines:
-      if line.startswith("#"):
-        continue
-      (solver, attributes, storage, deviation, nline) = re.split("[\t]", line)
-       
-      attributes = int(attributes)
-      storage = float(storage)
-      deviation = float(deviation)
-      if solver in storagePerSolver:
-        storagePerSolver[solver].append((storage,deviation))
-      else:
-        storagePerSolver[solver] = [(storage,deviation)]
-      if len(attributeSizes) == 0 or attributes > attributeSizes[-1]:
-        attributeSizes.append(attributes)
-
+            
+def main(dirn, fname): 
+  (xs, ysPerSolver, ydevsPerSolver) = CommonViz.parseData(dirn, fname)
+     
   CommonConf.setupMPPDefaults()
   fmts = CommonConf.getLineFormats()
   fig = pp.figure()
   ax = fig.add_subplot(111)
   ax.set_xscale("log", basex=2)
-    
+
   index = 0
-  for solver, storageAndDeviation in storagePerSolver.iteritems():
-    storageAndDeviationLists = map(list, zip(*storageAndDeviation))
-    ax.errorbar(attributeSizes, storageAndDeviationLists[0],
-                yerr=storageAndDeviationLists[1], label=solver,
-                marker=fmts[index][0], linestyle=fmts[index][1])
+  for (solver, ys), (solver, ydevs) in zip(ysPerSolver.iteritems(),ydevsPerSolver.iteritems()) : 
+    ax.errorbar(xs, ys, yerr=ydevs, label=solver, marker=fmts[index][0], linestyle=fmts[index][1])
     index = index + 1
 
-  ax.set_xlabel('Number of Attributes');
-  ax.set_ylabel('Storage Overhead ($\alpha$)');
-  # ax.set_xlim(0, 2100)
[…]

### 2014-09-01 code/simulation/viz/QueryIOVsNumAttributes.py
Refactored older experiments to use the new common code.

@@ -1,54 +1,36 @@
 import CommonConf
+import CommonViz
 
 import re
 from collections import OrderedDict
 import numpy as np
 import matplotlib.pyplot as pp
 
-def main(dirn, fname):
-  attributeSizes = []
-  ioPerSolver = OrderedDict()
-
-  with open(dirn+"/"+fname+".dat") as fin:
-    lines = fin.readlines()
-
-    for line in lines:
-      if line.startswith("#"):
-        continue
-      (solver, attributes, io, deviation, nline) = re.split("[\t]", line)
-      attributes = int(attributes)
-      deviation = int(deviation)
-      io = float(io)
-      if solver in ioPerSolver:
-        ioPerSolver[solver].append((io,deviation))
-      else:
-        ioPerSolver[solver] = [(io,deviation)]
-      if len(attributeSizes) == 0 or attributes > attributeSizes[-1]:
-        attributeSizes.append(attributes)
-
+EXPERIMENT_NAME = "QueryIOVsNumAttributes"
+X_LABEL         = "Num Attributes"
+Y_LABEL         = "I/O Cost (bytes)"
+            
+def main(dirn, fname): 
+  (xs, ysPerSolver, ydevsPerSolver) = CommonViz.parseData(dirn, fname)
+     
   CommonConf.setupMPPDefaults()
   fmts = CommonConf.getLineFormats()
   fig = pp.figure()
   ax = fig.add_subplot(111)
   ax.set_xscale("log", basex=2)
-    
+
   index = 0
+  for (solver, ys), (solver, ydevs) in zip(ysPerSolver.iteritems(),ydevsPerSolver.iteritems()) : 
+    ax.errorbar(xs, ys, yerr=ydevs, label=solver, marker=fmts[index][0], linestyle=fmts[index][1])
+    index = index + 1
 
-  for solver, ioAndDeviation in ioPerSolver.iteritems():
-    ioAndDeviationLists = map(list, zip(*ioAndDeviation))
-    ax.errorbar(attributeSizes, ioAndDeviationLists[0],
-                yerr=ioAndDeviationLists[1], label=solver,
-                marker=fmts[index][0], linestyle=fmts[index][1])
-    index = index + 1  
-
-  ax.set_xlabel('Number of Attributes');
-  ax.set_ylabel('I/O Cost (bytes)');
-  # ax.set_xlim(0, 2100)
[…]

### 2014-09-01 code/simulation/viz/RunningTimeVsNumAttributes.py

@@ -1,55 +1,37 @@
 import CommonConf
+import CommonViz
 
 import re
 from collections import OrderedDict
 import numpy as np
 import matplotlib.pyplot as pp
 
-def main(dirn, fname):
-  attributeSizes = []
-  timePerSolver = OrderedDict()
+EXPERIMENT_NAME = "RunningTimeVsNumAttributes.py"
+X_LABEL         = "Num Attributes"
+Y_LABEL         = "Time (sec.)"
 
-  with open(dirn+"/"+fname+".dat") as fin:
-    lines = fin.readlines()
-
-    for line in lines:
-      if line.startswith("#"):
-        continue
-      (solver, attributes, time, deviation, nline) = re.split("[\t]", line)
-
-      attributes = int(attributes)
-      time = float(time)
-      deviation = float(deviation)
-
-      if solver in timePerSolver:
-        timePerSolver[solver].append((time, deviation))
-      else:
-        timePerSolver[solver] = [(time, deviation)]
-      if len(attributeSizes) == 0 or attributes > attributeSizes[-1]:
-        attributeSizes.append(attributes)
-
+            
+def main(dirn, fname): 
+  (xs, ysPerSolver, ydevsPerSolver) = CommonViz.parseData(dirn, fname)
+     
   CommonConf.setupMPPDefaults()
   fmts = CommonConf.getLineFormats()
   fig = pp.figure()
   ax = fig.add_subplot(111)
   ax.set_xscale("log", basex=2)
-    
+
   index = 0
-  for solver, timesAndDeviation in timePerSolver.iteritems():
-    timesAndDeviationLists = map(list, zip(*timesAndDeviation))
-    ax.errorbar(attributeSizes, timesAndDeviationLists[0],
-                yerr=timesAndDeviationLists[1], label=solver,
-                marker=fmts[index][0], linestyle=fmts[index][1])
+  for (solver, ys), (solver, ydevs) in zip(ysPerSolver.iteritems(),ydevsPerSolver.iteritems()) : 
+    ax.errorbar(xs, ys, yerr=ydevs, label=solver, marker=fmts[index][0], linestyle=fmts[index][1])
     index = index + 1
 
-  ax.set_xlabel('Number of Attributes');
-  ax.set_ylabel('Time (sec.)');
-  # ax.set_xlim(0, 2100)
+  ax.set_xlabel(X_LABEL);
[…]

### 2014-09-01 code/simulation/viz/StorageOverheadVsNumQueryKinds.py

@@ -1,54 +1,37 @@
 import CommonConf
+import CommonViz
 
 import re
 from collections import OrderedDict
 import numpy as np
 import matplotlib.pyplot as pp
 
-def main(dirn, fname):
-  attributeSizes = []
-  storagePerSolver = OrderedDict()
+EXPERIMENT_NAME = "StorageOverheadVsNumQueryKinds"
+X_LABEL         = "Num Query Kinds"
+Y_LABEL         = "Storage Overhead"
 
-  with open(dirn+"/"+fname+".dat") as fin:
-    lines = fin.readlines()
-
-    for line in lines:
-      if line.startswith("#"):
-        continue
-      (solver, attributes, storage, deviation, nline) = re.split("[\t]", line)
-       
-      attributes = float(attributes)
-      storage = float(storage)
-      deviation = float(deviation)
-      if solver in storagePerSolver:
-        storagePerSolver[solver].append((storage,deviation))
-      else:
-        storagePerSolver[solver] = [(storage,deviation)]
-      if len(attributeSizes) == 0 or attributes > attributeSizes[-1]:
-        attributeSizes.append(attributes)
-
+            
+def main(dirn, fname): 
+  (xs, ysPerSolver, ydevsPerSolver) = CommonViz.parseData(dirn, fname)
+     
   CommonConf.setupMPPDefaults()
   fmts = CommonConf.getLineFormats()
   fig = pp.figure()
   ax = fig.add_subplot(111)
   ax.set_xscale("log", basex=2)
-    
+
   index = 0
-  for solver, storageAndDeviation in storagePerSolver.iteritems():
-    storageAndDeviationLists = map(list, zip(*storageAndDeviation))
-    ax.errorbar(attributeSizes, storageAndDeviationLists[0],
-                yerr=storageAndDeviationLists[1], label=solver,
-                marker=fmts[index][0], linestyle=fmts[index][1])
+  for (solver, ys), (solver, ydevs) in zip(ysPerSolver.iteritems(),ydevsPerSolver.iteritems()) : 
+    ax.errorbar(xs, ys, yerr=ydevs, label=solver, marker=fmts[index][0], linestyle=fmts[index][1])
     index = index + 1
 
-  ax.set_xlabel('Number of Query Kinds');
-  ax.set_ylabel(r'Storage Overhead ($\alpha$)');
[…]

### 2014-09-01 code/simulation/viz/StorageOverheadVsStorageOverheadThreshold.py

@@ -1,58 +1,37 @@
 import CommonConf
+import CommonViz
 
 import re
 from collections import OrderedDict
 import numpy as np
 import matplotlib.pyplot as pp
 
-def main(dirn, fname):
-  storageOverheadThresholdSizes = []
-  storagePerSolver = OrderedDict()
+EXPERIMENT_NAME = "StorageOverheadVsStorageOverheadThreshold.py"
+X_LABEL         = "Storage Overhead Threshold"
+Y_LABEL         = "Storage Overhead"
 
-  with open(dirn+"/"+fname+".dat") as fin:
-    lines = fin.readlines()
-
-    for line in lines:
-      if line.startswith("#"):
-        continue
-      (solver, storageOverheadThresholds, storage, deviation, nline) = re.split("[\t]", line)
-      if 'n/a' in storage:
-        storage = 0
-        deviation = 0
-      else:
-        storage = float(storage)
-        deviation = float(deviation)
-       
-      storageOverheadThresholds = float(storageOverheadThresholds)
-      if solver in storagePerSolver:
-        storagePerSolver[solver].append((storage,deviation))
-      else:
-        storagePerSolver[solver] = [(storage,deviation)]
-      if len(storageOverheadThresholdSizes) == 0 or storageOverheadThresholds > storageOverheadThresholdSizes[-1]:
-        storageOverheadThresholdSizes.append(storageOverheadThresholds)
-
+            
+def main(dirn, fname): 
+  (xs, ysPerSolver, ydevsPerSolver) = CommonViz.parseData(dirn, fname)
+     
   CommonConf.setupMPPDefaults()
   fmts = CommonConf.getLineFormats()
   fig = pp.figure()
   ax = fig.add_subplot(111)
-  #ax.set_xscale("log", basex=2)
-    
+  ax.set_xscale("log", basex=2)
+
   index = 0
-  for solver, storageAndDeviation in storagePerSolver.iteritems():
-    storageAndDeviationLists = map(list, zip(*storageAndDeviation))
-    ax.errorbar(storageOverheadThresholdSizes, storageAndDeviationLists[0],
-                yerr=storageAndDeviationLists[1], label=solver,
-                marker=fmts[index][0], linestyle=fmts[index][1])
[…]

### 2014-09-01 docs/edbt2015/text/paper.tex
edbt submission skeleton

--- 2014-07-29 docs/paper/text/solution.tex
@@
-\clearpage
-\newpage
-\section{Partitioning}
-This section is still a work-in-progress. We are trying to find the algorithms.
+\input{config}
 
-\subsection{Non-Overlapping Attributes}
+\begin{document}
+%
+% --- Author Metadata here ---
+\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
+%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
+%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
+% --- End of Author Metadata ---
 
-\paragraph*{Problem.$\,$} \emph{Find a true partitioning of attributes that
-minimizes the query I/O and bounds the storage cost by some upper limit.}
+\title{Adaptive Disk Storage for Interaction Graphs}
 
-\subsubsection{Integer Linear Program Formulation}\label{subsubsec:nov-ilp}
-We present an ILP formuation of the problem. For this purpose, we define a 
-number of binary ($0$ or $1$) variables: 
-\begin{itemize}
-\item $x_{a,p}$: $1$ if attribute $a$ is in partition $p$, $0$ otherwise.
-\item $y_{p,q}$: $1$ if partition $p$ is used by query $q$, $0$ otherwise.
-\item $z_{a,p,q}$: $1$ if partition $p$ is used by query $q$ and attribute $a$
-is in partition $p$, $0$ otherwise.
-\item $u_{p}$: $1$ if partition $p$ is assigned at least $1$ attribute, $0$ otherwise.
-\end{itemize}
-
-Each of these variables serve a purpose:
-\begin{itemize}
-\item $x$s define the attribute-to-partition assignment.
-\item $y$s help formulate the query I/O contribution
-of partitions, excluding their assigned attributes.
-\item $z$s help formulate the query I/O contribution
-of partitions, only considering their assigned attributes.
-\item $u$s help formulate the storage overhead requirement.
-\end{itemize}
-
-We also define a helper notation for representing whether a variable is
-accessed by a query or not: $q(a)\equiv \mathbf{1}(a \in q.A)$, where
-$\mathbf{1}$ is the indicator function. We assume that there are a maximum of
-$k$ partitions. $k$ can be taken as $|A|$, as there cannot be more partitions
-than there are attributes.
-
-We are now ready to state the ILP formulation. We start with the objective
-function, that is the total query I/O:
-\begin{eqnarray}
-\sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot
-y_{p,q}\nonumber\\ 
-&+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q}\label{eq:no-obj}
-\end{eqnarray}
-
-In Eq.~\ref{eq:no-obj}, we simply sum for each query and each partition, and
-add the I/O cost of reading in the structural information found in a
-sub-block, if the partition is used by the query. We then sum over each
-attribute as well, and add the I/O cost of reading in the attributes. Note
-that $z_{a,p,q}$ could have been  replaced with $x_{a,p}\cdot y_{p,q}$, but
-that would make the objective function non-linear.
-
-We are now ready to state our constraints. Our first constraint is that, each
-attribute must be assigned to a single partition. Formally:
-\begin{eqnarray}
-\forall_{a\in A}, \sum_{p=1}^{k} x_{a,p} = 1
-\end{eqnarray}
-
-Our second constraint is that, if a query $q$ contains an attribute $a$
-assigned to a partition $p$, then partition $p$ is used by the query, i.e.,
-$y_{p,q}=1$. In essence, we want to state: $\forall_{\{p,q\}\in [1..k]\times
-Q}, y_{p,q} = \mathbf{1}(\sum_{a\in A} q(a)\cdot x_{a,p}>0)$.
-
-In order to formulate this constraint, we use the following ILP  construction:
-Assume we have two variables, $\beta_1$ and $\beta_2$, where $\beta_2\in[0,1]$
-and $\beta_1\geq 0$. We want to implement the following constraint: $\beta_2 =
-\mathbf{1}(\beta_1 > 0)$. This could be expressed as a linear constraint as
-follows, where $K$ is a large constant guaranteed to be larger than $\beta_1$
-for practical purposes:
-\begin{eqnarray}
-&& \beta_1 - \beta_2 \geq 0\nonumber\\
-&& K\cdot\beta_2 - \beta_1 \geq 0\label{eq:beta-ilp}
-\end{eqnarray}
-
-We now apply this construction to our second constraint, where
-$\beta_1=\sum_{a\in A} q(a)\cdot x_{a,p}$ and $\beta_2=y_{p,q}$. This results
-in the following linear constraints:
-\begin{eqnarray}
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  \sum_{a\in A} q(a)\cdot x_{a,p} - y_{p,q} \geq 0 \nonumber\\
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  K\cdot y_{p,q} - \sum_{a\in A} q(a)\cdot x_{a,p}  \geq 0 
-\end{eqnarray}
-
-Our third constraint is that, if an attribute $a$ is assigned to a partition
-$p$, and partition $p$ is used by a query $q$, then the corresponding $z$
-variable must be set to $1$. That is, we want: $\forall_{\{a,p,q\}\in A\times
-[1..k]\times Q}, z_{a,p,q}=\mathbf{1}(x_{a,p} = y_{p,q} = 1)$. We express this
-as a linear  constraint, as follows:
-\begin{eqnarray}
-\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
-    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\label{eq:no-z}
-\end{eqnarray}
-
-In Eq.~\ref{eq:no-z}, when the $x$ and $y$ variables are both $1$, the  $z$
-variable is simply forced to be $1$. Otherwise, the $z$ variable can be either
-$0$ or $1$, but since the $z$ variables appear in the objective fucntion as
-positive terms, the solver will set them to $0$, which is what we want.
-
-Our fourth constraint is that, if a partition is non-empty, then its
-corresponding $u$ variable must be set to $0$. In other words,  we want
-$\forall_{p\in[1..k]}, u_p = \mathbf{1}(\sum_{a\in A} x_{a,p}>0)$. This is
-expressed as linear constraints, as follows:
-\begin{eqnarray}
-\forall_{p\in[1..k]},
-    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
-\forall_{p\in[1..k]},
-    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \label{eq:no-u}
-\end{eqnarray}
-
-Eq.~\ref{eq:no-u} uses the same construction as the second constraint, where
-$\beta_1=\sum_{a\in A} x_{a,p}$ and $\beta_2=u_p$.
-
-Our fifth, and the last, constraint deals with the storage overhead. We want to
- make sure that the storage overhead does not go over $\alpha$. The storage
-overhead depends on the number of partitions used. That means that the only 
-ILP variables it depends on is the $u$s. In particular, the number of
-partitions used is given by $\sum_{p=1}^{k} u_p$. This results in the
-following linear constraint:
-\begin{equation}
-\sum_{p=1}^{k} u_p \leq 1 + \frac{\alpha}
-  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}}
-\end{equation}
-
-% NOTE: In the simulation we will have query weights that will factor into the objective function
-\begin{figure}[!t]
-\begin{mdframed}
-\begin{eqnarray}
-\text{minimize}  
-    \sum_{q\in Q} \sum_{p=1}^{k} \!\!&&\!\! (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
-    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
-\text{subject to}&&\nonumber\\
-\forall_{a\in A}, 
-    && \sum_{p=1}^{k} x_{a,p} = 1\nonumber\\
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  \sum_{a\in A} q(a)\cdot x_{a,p} - y_{p,q} \geq 0 \nonumber\\
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  K\cdot y_{p,q} - \sum_{a\in A} q(a)\cdot x_{a,p}  \geq 0 \nonumber\\
-\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
-    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\nonumber\\
-\forall_{p\in[1..k]},
-    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
-\forall_{p\in[1..k]},
-    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
-&& \sum_{p=1}^{k} u_p \leq 1 + \frac{\alpha}
-  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}} \nonumber
-\end{eqnarray}
-\end{mdframed}
-\caption{ILP formulation for the non-overlapping partitioning}
-\label{fig:nov-ilp}
-\end{figure}
-
-The final ILP formulation for the non-overlapping partitioning is given in
-Figure~\ref{fig:nov-ilp}.
-
-\subsubsection{Heuristic Solution}
-
-\paragraph*{Approach.$\,$}
-TODO
-
-\begin{algorithm}[ht]
-\scriptsize
-\caption{Algorithm for partitioning blocks into sub-blocks with non-overlapping attributes.}
-\label{alg:overlappingP}
-\KwData{$B$: block, $Q$: set of queries}
-$c^*\leftarrow \infty$ \tcp*{Lowest cost over all \# of partitions}
-\For(\tcp*[f]{For each possible \# of partitions}){$k=1$ to $|A|$}{
-   $R[i]\leftarrow \emptyset, \forall i\in [1..k]$ \tcp*{Initialize partitions}
-   \For(\tcp*[f]{For each attribute}){$a \in A$\textnormal{, in decr.\/ order of }$f(a)$}{
-      $c\leftarrow \infty$ \tcp*{Lowest cost over all assignments}  
-      $j\leftarrow -1$ \tcp*{Best partition assignment}
-      \For(\tcp*[f]{For each partition assignment}){$i\in [1..k]$} {
-         $R[i]\leftarrow R[i] \cup \{a\}$\tcp*{Assign attribute}
-         \If(\tcp*[f]{If query cost is lower}){$L(R, B, Q)<c$}{
-            $c\leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
-            $j\leftarrow i$\tcp*{Update the best partition}
-         }
-         $R[i]\leftarrow R[i] \setminus \{a\}$\tcp*{Un-assign attribute}
-      }
-      $R[j]\leftarrow R[j] \cup \{a\}$\tcp*{Assign to best partition}
-   }
-   \tcp{If solution is feasable and has lower cost}
-   \If{$H(R, B, Q)\geq\alpha \wedge L(R, B, Q)<c^*$}{
-      $c^* \leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
-      $\mathcal{P}(B)\leftarrow R$\tcp*{Update the best partitioning}
-   }
-}
-\Return $\mathcal{P}(B)$ \tcp*{Final set of sub-blocks}
-\end{algorithm}
-
-\clearpage
-\newpage
-\subsection{Overlapping Attributes}
-
-\paragraph*{Problem.$\,$} \emph{Find an overlapping partitioning of attributes
-that minimizes the query I/O and bounds the storage cost by some upper
-limit.}
-
-\subsubsection{Integer Linear Program Formulation}
-We present an ILP formuation of the problem, as we did for the case of
-non-overlapping partitions in Section~\ref{subsubsec:nov-ilp}. We use the same
-set of variables and the same objective function. However, the formulation of
-the constraints differ.
-
-Our first constraint is that, each attribute must be assigned to at leat one
-partition. Formally:
-\begin{eqnarray}
-\forall_{a\in A}, \sum_{p=1}^{k} x_{a,p} \geq 1
-\end{eqnarray}
-
-As our second constraint, we require that for each attribute used by a query,
-there needs to be a partition that is used by that query and that contains the
-attribute in question. Formally:
-\begin{eqnarray}
-\forall_{\{a,q\}\in A\times Q}, \sum_{p=1}^{k} z_{a,p,q} \geq q(a) 
-\end{eqnarray}
-
-\begin{sloppypar}
-As our third constraint, we require that if a query is using an attribute from
-a partition, then that partition must contain the attribute. I.e., we need
-to link the $z$ variables with the $x$ variables as
-$\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, (z_{a,p,q} = 1) \implies 
-(x_{a,p} = 1)$. This can be state as linear constraints:
-\begin{eqnarray}
-\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, x_{a,p} - z_{a,p,q} \geq 0 
-\end{eqnarray}
-\end{sloppypar}
-
-As our fourth constraint, we require that if a query is using at least one
-attribute from a partition, then that partition must be used by the query.
-I.e., we need to link the $z$ variables with the $y$ variables as
-$\forall_{\{p,q\}\in [1..k]\times Q}, y_{p,q} = \mathbf{1}(\sum_{a\in A}
-z_{a,p,q}>0)$. As before, we use the ILP construction from
-Eq.~\ref{eq:beta-ilp} for this, where $\beta_2=y_{p,q}$ and $\beta_1 =
-\sum_{a\in A} z_{a,p,q}$. We get:
-\begin{eqnarray}
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  \sum_{a\in A} z_{a,p,q} - y_{p,q} \geq 0 \nonumber\\
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  K\cdot y_{p,q} - \sum_{a\in A} z_{a,p,q} \geq 0 
-\end{eqnarray}
-
-Our fifth constraint is that, if an attribute $a$ is assigned to a partition
-$p$, and partition $p$ is used by a query $q$, then the corresponding $z$
-variable must be set to $1$. This is same as the formulation for the
-non-overlapping case from Eq.~\ref{eq:no-z}.
-
-Our sixth constraint is that, if a partition is non-empty, then its
-corresponding $u$ variable must be set to $0$. Again, this is same as the
-formulation for the non-overlapping case from Eq.~\ref{eq:no-u}.
-
-Our seventh, and the last, constraint deals with the storage overhead. However,
-the storage overhead formulation for the overlapping case is different from 
-the one for the non-overlapping. This is because the overhead does not merely
-depend on the number of partitions, as attributes might have to be loaded
-multiple times from different partitions. As a result, we express the overhead
-using base variables as in the objective function. Formally:
-\begin{eqnarray}
-\sum_{p=1}^{k} (16\cdot c_e(B) &+& 12 \cdot c_n(B)) \cdot u_p  \nonumber \\ 
-+ \sum_{a\in A} s(a) \!\!&\cdot&\!\! c_n(B)\cdot x_{a,p} \leq s(B)\cdot (1+\alpha)
-\end{eqnarray}
-
-% NOTE: In the simulation we will have query weights that will factor into the objective function
-\begin{figure}[!t]
-\begin{mdframed}
-\begin{eqnarray}
-\text{minimize}  
-    \sum_{q\in Q} \sum_{p=1}^{k} \!\!&&\!\! (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
-    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
-\text{subject to}&&\nonumber\\
-\forall_{a\in A}, 
-    && \sum_{p=1}^{k} x_{a,p} \geq 1\nonumber\\
-\forall_{\{a,q\}\in A\times Q},
-    &&  \sum_{p=1}^{k} z_{a,p,q} \geq q(a) \nonumber\\
-\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, 
-    && x_{a,p} - z_{a,p,q} \geq 0 \nonumber\\
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  \sum_{a\in A} z_{a,p,q} - y_{p,q} \geq 0 \nonumber\\
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  K\cdot y_{p,q} - \sum_{a\in A} z_{a,p,q}  \geq 0 \nonumber\\
-\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, 
-    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1 \nonumber\\
-\forall_{p\in[1..k]},
-    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
-\forall_{p\in[1..k]},
-    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
-\sum_{p=1}^{k} (16\cdot c_e(B) &+& 12 \cdot c_n(B)) \cdot u_p  \nonumber \\ 
-+ \sum_{a\in A} s(a) \!\!&\cdot&\!\! c_n(B)\cdot x_{a,p} \leq s(B)\cdot (1+\alpha)\nonumber
-\end{eqnarray}
-\end{mdframed}
-\caption{ILP formulation for the overlapping partitioning}
-\label{fig:ov-ilp}
-\end{figure}
-
-The final ILP formulation for the overlapping partitioning is given in
-Figure~\ref{fig:ov-ilp}.
-
-\subsubsection{Heuristic Solution}
-
-\paragraph*{Approach.$\,$} We start the algorithm with a 
- partitioning based on what queries we have seen.
-Every query gets its own sub-block. This is the``ideal'' partitioning,
-because the I/O cost would be minimized for every query that we would
-have seen. As input to the algorithm, we provide a function $\dist$, which
-is used to compute the distance between two blocks.
-
-$$
-\dist: block \rightarrow block \rightarrow int
-$$
-
-Note that the $\dist$ function can be chosen arbitrarily.  Examples of useful
-$\dist$ functions are the frequencies with which the queries are accessed, or the Jaccard index, which measures the set-similarity of
-the attributes.
-
-Using this function, the algorithm iteratively combines the two partitions that
-are closest together. Algorithm~\ref{alg:closestBF} shows the pseudocode for a brute-force approach
-  to finding the closest paris that runs in $O(n^2)$ time. However, there is
-  also a divide-and-conquer algorithm~\cite{cormen01} for solving closest pair
-  of points problem in $O(n \log n)$.
-After each combination of partitions, the algorithm calculate the storage cost
-for the partitioning. The algorithm stops when the  storage cost is below some
-specified threshold.  The result is the block partitioning.
-
-\begin{algorithm}[ht]
-\SetKwFunction{closestPair}{closestPair}
-\scriptsize
-\caption{Algorithm for partitioning blocks into sub-blocks with overlapping attributes.}
-\label{alg:closestPair}
-\KwData{$B$: block, $Q$: set of queries}
-$\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}
-
-\For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
- $\mathcal{P}(B) = \mathcal{P}(B) \cup q.A$ 
+\numberofauthors{2} 
+\author{
+\alignauthor
+Robert Soul\'{e}\\
+       \affaddr{University of Lugano}\\
+       \email{robert.soule@usi.ch}  
+\alignauthor
+Bu\u{g}ra Gedik\\
+       \affaddr{Bilkent University}\\
+       \email{bgedik@cs.bilkent.edu.tr}
 }
 
-\While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
-  \text{max}$}{
-$\text{x}, \text{y}  \leftarrow  \closestPair{}  $\;
- $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - \{x, y\}$ \;
-  $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  \cup \{ x \cup y \}$ \;
-}
-\Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
-\end{algorithm}
+\maketitle
+\begin{abstract}
+  \input{abstract}            
+\end{abstract}
 
- \begin{algorithm}[ht]
-\SetKwFunction{length}{length}
-\SetKwFunction{dist}{dist}
- \scriptsize
- \caption{Brute-force algorithm for closest points in $O(n^2)$.}
- \label{alg:closestBF}
-$minDist  \gets \infty $ \;
-\For{$i \gets 1$ \textbf{to} $\length(P)-1$} {
-\For{$j \gets i+ 1$ \textbf{to} $\length(P)$} {
+% A category with the (minimum) three required fields
+\category{H.4}{Information Systems Applications}{Miscellaneous}
+%A category including the fourth, optional field follows...
+\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
 
-  $p \gets P[i]$ \;
-  $q \gets P[j]$ \;
-  \If{$\dist(p,q) < minDist$} {
-    $minDist \gets \dist(p,q)$\;
-    $closestPair \gets (p,q)$\;
-  }
-}
-}
-\Return $closestPair$
-% \KwData{$B$: block, $Q$: set of queries}
-% $\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}
-% \For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
-%  $\mathcal{P}(B) = \mathcal{P}(B) \cup q.A$ 
-% }
-% \While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
-%   \text{max}$}{
-% $\text{x}, \text{y}  \leftarrow  \closestPair{}  $\;
-%  $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - x - y$ \;
-%   $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  + x \cup y$ \;
-% \Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
- \end{algorithm}
+\terms{Theory}
+
+\keywords{ACM proceedings, \LaTeX, text tagging}
+
+\section{Introduction}
+
+\section{Related Work}
+
+\begin{alltt}\scriptsize
+Gedik et al.~\cite{gedik14}
+H2O \cite{alagiannis14}
+HYRISE~\cite{grund10}
+Bornea et al.~\cite{bornea13}
+\end{alltt}
+
+\bibliographystyle{abbrv}
+\bibliography{paper}  
+\end{document}

### 2014-09-01 docs/vldbj/text/solution.tex
Renamed paper to vldbj

--- 2014-07-29 docs/paper/text/solution.tex
@@
 \begin{eqnarray}
-\sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot
+\sum_{q\in Q} w(q) \cdot \Big(\sum_{p=1}^{k} \!\!&&\!\! (16\cdot c_e(B) + 12\cdot c_n(B))\cdot
 y_{p,q}\nonumber\\ 
-&+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q}\label{eq:no-obj}
+&+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q}\Big)\label{eq:no-obj}
 \end{eqnarray}
@@
 
-% NOTE: In the simulation we will have query weights that will factor into the objective function
 \begin{figure}[!t]
@@
 \text{minimize}  
-    \sum_{q\in Q} \sum_{p=1}^{k} \!\!&&\!\! (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
-    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
+    \sum_{q\in Q} w(q)\cdot \Big(\sum_{p=1}^{k} \!\!&&\!\! (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
+    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \Big) \nonumber\\
 \text{subject to}&&\nonumber\\
@@
 \caption{Algorithm for partitioning blocks into sub-blocks with non-overlapping attributes.}
-\label{alg:overlappingP}
+\label{alg:non-overlappingP}
 \KwData{$B$: block, $Q$: set of queries}
@@
    }
-   \tcp{If solution is feasable and has lower cost}
-   \If{$H(R, B, Q)\geq\alpha \wedge L(R, B, Q)<c^*$}{
+   \lIf(\tcp*[f]{If solution infeasible}){$H(R, B, Q)>\alpha$}{\textbf{break}}
+   \If(\tcp*[f]{If solution has lower cost}){$L(R, B, Q)<c^*$}{
      $c^* \leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
@@
 \begin{eqnarray}
-\sum_{p=1}^{k} (16\cdot c_e(B) &+& 12 \cdot c_n(B)) \cdot u_p  \nonumber \\ 
-+ \sum_{a\in A} s(a) \!\!&\cdot&\!\! c_n(B)\cdot x_{a,p} \leq s(B)\cdot (1+\alpha)
+\sum_{p=1}^{k} \Big((16\cdot c_e(B) &+& 12 \cdot c_n(B)) \cdot u_p  \nonumber \\ 
++ \sum_{a\in A} s(a) \!\!&\cdot&\!\! c_n(B)\cdot x_{a,p} \Big) \leq s(B)\cdot (1+\alpha)
 \end{eqnarray}
 
-% NOTE: In the simulation we will have query weights that will factor into the objective function
 \begin{figure}[!t]
@@
     && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
-\sum_{p=1}^{k} (16\cdot c_e(B) &+& 12 \cdot c_n(B)) \cdot u_p  \nonumber \\ 
-+ \sum_{a\in A} s(a) \!\!&\cdot&\!\! c_n(B)\cdot x_{a,p} \leq s(B)\cdot (1+\alpha)\nonumber
+\sum_{p=1}^{k} \Big((16\cdot c_e(B) &+& 12 \cdot c_n(B)) \cdot u_p  \nonumber \\ 
++ \sum_{a\in A} s(a) \!\!&\cdot&\!\! c_n(B)\cdot x_{a,p}  \Big)\leq s(B)\cdot (1+\alpha)\nonumber
 \end{eqnarray}
@@
 
-\paragraph*{Approach.$\,$} We start the algorithm with a 
- partitioning based on what queries we have seen.
-Every query gets its own sub-block. This is the``ideal'' partitioning,
-because the I/O cost would be minimized for every query that we would
-have seen. As input to the algorithm, we provide a function $\dist$, which
-is used to compute the distance between two blocks.
-
-$$
-\dist: block \rightarrow block \rightarrow int
-$$
-
-Note that the $\dist$ function can be chosen arbitrarily.  Examples of useful
-$\dist$ functions are the frequencies with which the queries are accessed, or the Jaccard index, which measures the set-similarity of
-the attributes.
-
-Using this function, the algorithm iteratively combines the two partitions that
-are closest together. Algorithm~\ref{alg:closestBF} shows the pseudocode for a brute-force approach
-  to finding the closest paris that runs in $O(n^2)$ time. However, there is
-  also a divide-and-conquer algorithm~\cite{cormen01} for solving closest pair
-  of points problem in $O(n \log n)$.
-After each combination of partitions, the algorithm calculate the storage cost
-for the partitioning. The algorithm stops when the  storage cost is below some
-specified threshold.  The result is the block partitioning.
+\paragraph*{Approach.$\,$} We start the algorithm with a   partitioning based
+on what queries we have seen. Every query gets its own sub-block. This is the
+``ideal'' partitioning, because the I/O cost would be minimized for every
+query that we would have seen. The algorithm iteratively combines the two
+partitions that are closest together.  After each combination of partitions,
+the algorithm calculate the storage overhead for the partitioning. The
+algorithm stops when the  storage cost is below some specified threshold.  The
+result is the block partitioning.
 
 \begin{algorithm}[ht]
-\SetKwFunction{closestPair}{closestPair}
 \scriptsize
 \caption{Algorithm for partitioning blocks into sub-blocks with overlapping attributes.}
-\label{alg:closestPair}
+\label{alg:overlappingP}
 \KwData{$B$: block, $Q$: set of queries}
-$\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}
-
-\For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
- $\mathcal{P}(B) = \mathcal{P}(B) \cup q.A$ 
+$\mathcal{P}(B) \leftarrow \{q.A: q\in Q\}$ \tcp*{Each query gets its own sub-block}
+\While(\tcp*[f]{Until storage overhead is below $\alpha$}){$H(\mathcal{P},B) > \alpha$}{
+  $c^{*}\gets \infty $ \tcp*{Lowest cost over all sub-block pairs}
+  $(b_x,b_y)\gets (\emptyset,\emptyset)$ \tcp*{Sub-block pair with the lowest cost}
+  \For(\tcp*[f]{For each pair of blocks}){$\{b_i,b_j\}\in\mathcal{P}(B)$}{
+    $\mathcal{P'}(B) \leftarrow \mathcal{P}(B) \setminus \{b_i, b_j\} \cup \{b_i \cup b_j\}$\\
+    $c\gets \frac{L(\mathcal{P}',B,Q)-L(\mathcal{P},B,Q)}{H(\mathcal{P},B)-H(\mathcal{P}',B)}$\tcp*{Cost of merge}
+    \If(\tcp*[f]{Cost is lower}){$c<c^{*}$}{
+        $c^{*}\gets c$\tcp*{Update the lowest cost}
+        $(b_x,b_y)\gets (b_i,b_j)$\tcp*{Update the best pair}
     }
-
-\While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
-  \text{max}$}{
-$\text{x}, \text{y}  \leftarrow  \closestPair{}  $\;
- $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - \{x, y\}$ \;
-  $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  \cup \{ x \cup y \}$ \;
+  }
+  $\mathcal{P}(B) \leftarrow \mathcal{P}(B) \setminus \{b_x, b_y\} \cup \{b_x \cup b_y\}$ \\
 }
@@
 \end{algorithm}
-
- \begin{algorithm}[ht]
-\SetKwFunction{length}{length}
-\SetKwFunction{dist}{dist}
- \scriptsize
- \caption{Brute-force algorithm for closest points in $O(n^2)$.}
- \label{alg:closestBF}
-$minDist  \gets \infty $ \;
-\For{$i \gets 1$ \textbf{to} $\length(P)-1$} {
-\For{$j \gets i+ 1$ \textbf{to} $\length(P)$} {
-
-  $p \gets P[i]$ \;
-  $q \gets P[j]$ \;
-  \If{$\dist(p,q) < minDist$} {
-    $minDist \gets \dist(p,q)$\;
-    $closestPair \gets (p,q)$\;
-  }
-}
-}
-\Return $closestPair$
-% \KwData{$B$: block, $Q$: set of queries}
-% $\mathcal{P}(B) \leftarrow \emptyset;$  \tcp*{Sub-blocks;}
-% \For(\tcp*[f]{Every query gets its own sub-block}){$q \in Q$}{ 
-%  $\mathcal{P}(B) = \mathcal{P}(B) \cup q.A$ 
-% }
-% \While(\tcp*[f]{While storage exceeds threshold}){$H(\mathcal{P}, B) \geq
-%   \text{max}$}{
-% $\text{x}, \text{y}  \leftarrow  \closestPair{}  $\;
-%  $ \mathcal{P}(B)  \leftarrow \mathcal{P}(B) - x - y$ \;
-%   $\mathcal{P}(B)  \leftarrow \mathcal{P}(B)  + x \cup y$ \;
-% \Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
- \end{algorithm}

### 2014-09-01 code/simulation/viz/QueryIOVsAttributeSizeSkew.py
bugs in the visualization scripts

@@ -17,7 +17,7 @@ def main(dirn, fname):
   fmts = CommonConf.getLineFormats()
   fig = pp.figure()
   ax = fig.add_subplot(111)
-  ax.set_xscale("log", basex=2)
+  # ax.set_xscale("log", basex=2)
 
   index = 0
   for (solver, ys), (solver, ydevs) in zip(ysPerSolver.iteritems(),ydevsPerSolver.iteritems()) : 

### 2014-09-01 code/simulation/viz/RunningTimeVsAttributeSizeSkew.py

@@ -18,7 +18,7 @@ def main(dirn, fname):
   fmts = CommonConf.getLineFormats()
   fig = pp.figure()
   ax = fig.add_subplot(111)
-  ax.set_xscale("log", basex=2)
+  #ax.set_xscale("log", basex=2)
 
   index = 0
   for (solver, ys), (solver, ydevs) in zip(ysPerSolver.iteritems(),ydevsPerSolver.iteritems()) : 

### 2014-09-01 code/simulation/viz/RunningTimeVsNumAttributes.py

@@ -6,7 +6,7 @@ from collections import OrderedDict
 import numpy as np
 import matplotlib.pyplot as pp
 
-EXPERIMENT_NAME = "RunningTimeVsNumAttributes.py"
+EXPERIMENT_NAME = "RunningTimeVsNumAttributes"
 X_LABEL         = "Num Attributes"
 Y_LABEL         = "Time (sec.)"
 

### 2014-09-01 code/simulation/viz/StorageOverheadVsStorageOverheadThreshold.py

@@ -6,7 +6,7 @@ from collections import OrderedDict
 import numpy as np
 import matplotlib.pyplot as pp
 
-EXPERIMENT_NAME = "StorageOverheadVsStorageOverheadThreshold.py"
+EXPERIMENT_NAME = "StorageOverheadVsStorageOverheadThreshold"
 X_LABEL         = "Storage Overhead Threshold"
 Y_LABEL         = "Storage Overhead"
 
@@ -18,7 +18,7 @@ def main(dirn, fname):
   fmts = CommonConf.getLineFormats()
   fig = pp.figure()
   ax = fig.add_subplot(111)
-  ax.set_xscale("log", basex=2)
+  #ax.set_xscale("log", basex=2)
 
   index = 0
   for (solver, ys), (solver, ydevs) in zip(ysPerSolver.iteritems(),ydevsPerSolver.iteritems()) : 

### 2014-09-01 code/simulation/viz/CommonViz.py
bugs in the visualization scripts

@@ -12,9 +12,13 @@ def parseData(dirn, fname):
       if line.startswith("#"):
         continue
       (solver, x, y, ydev, nline) = re.split("[\t]", line)
-      x = int(x)
-      y = float(y)
-      ydev = float(ydev)  
+      x = float(x)
+      try:
+        y = float(y)
+        ydev = float(ydev)  
+      except ValueError:
+        y = 0.0
+        ydev = 0.0  
       if solver in ysPerSolver:
         ysPerSolver[solver].append(y)
         ydevsPerSolver[solver].append(ydev)

### 2014-09-01 code/simulation/viz/StorageOverheadVsQueryFreqSkew.py

@@ -6,7 +6,7 @@ from collections import OrderedDict
 import numpy as np
 import matplotlib.pyplot as pp
 
-EXPERIMENT_NAME = "StorageOverheadVsQueryFreqSkew.py"
+EXPERIMENT_NAME = "StorageOverheadVsQueryFreqSkew"
 X_LABEL         = "Query Freq Skew"
 Y_LABEL         = "Storage Overhead"
 
@@ -18,7 +18,7 @@ def main(dirn, fname):
   fmts = CommonConf.getLineFormats()
   fig = pp.figure()
   ax = fig.add_subplot(111)
-  ax.set_xscale("log", basex=2)
+  #ax.set_xscale("log", basex=2)
 
   index = 0
   for (solver, ys), (solver, ydevs) in zip(ysPerSolver.iteritems(),ydevsPerSolver.iteritems()) : 

### 2014-09-02 code/simulation/include/intergdb/simulation/Experiments.h
removed unused Sample experiment

@@ -7,11 +7,6 @@ namespace intergdb { namespace simulation
 
   class ExperimentalData;
 
-  class SampleExperiment : public ExperimentalRun 
-  {
-    void process() override;
-  };
-
   class VsNumAttributes : public ExperimentalRun 
   {
       void process() override;

### 2014-09-06 docs/edbt2015/text/paper.tex
added figures to edbt paper

--- 2014-08-07 docs/cidr2015/text/main.tex
@@
 \input{config}
-
-%\documentclass{sig-alternate}
 
@@
 \numberofauthors{2} 
-
 \author{
-% 1st. author
 \alignauthor
@@
        \affaddr{University of Lugano}\\
-       \affaddr{Faculty of Informatics}\\
-       \affaddr{Lugano 6904, Switzerland}\\
        \email{robert.soule@usi.ch}  
-% 2nd. author
 \alignauthor
@@
        \affaddr{Bilkent University}\\
-       \affaddr{Department of Computer Engineering}\\
-       \affaddr{Bilkent, Ankara 06800, Turkey}
        \email{bgedik@cs.bilkent.edu.tr}
@@
 \begin{abstract}
-TODO
+  \input{abstract}            
 \end{abstract}
@@
 needs to access both.
-
-% This points to the need for
-% adaptively optimizing the layout (somewhat similar to H2O \cite{alagiannis14}
-% and HYRISE~\cite{grund10}).%
 
@@
 \begin{equation}
-L(\mathcal{P}, B) = \sum_{q\in Q} \mathbf{1}(q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} \!\!s(B')
+L(\mathcal{P}, B) = \sum_{q\in Q} w(q)\cdot\mathbf{1}(q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} \!\!s(B')
 \end{equation}
+
+Here, $w(q)$ represents the query weight.
 
@@
 
+\subsubsection{Heuristic Solution}
+
+\paragraph*{Approach.$\,$}
+TODO
+
+\begin{algorithm}[ht]
+\scriptsize
+\caption{Algorithm for partitioning blocks into sub-blocks with non-overlapping attributes.}
+\label{alg:non-overlappingP}
+\KwData{$B$: block, $Q$: set of queries}
+$c^*\leftarrow \infty$ \tcp*{Lowest cost over all \# of partitions}
+\For(\tcp*[f]{For each possible \# of partitions}){$k=1$ to $|A|$}{
+   $R[i]\leftarrow \emptyset, \forall i\in [1..k]$ \tcp*{Initialize partitions}
+   \For(\tcp*[f]{For each attribute}){$a \in A$\textnormal{, in decr.\/ order of }$f(a)$}{
+      $c\leftarrow \infty$ \tcp*{Lowest cost over all assignments}  
+      $j\leftarrow -1$ \tcp*{Best partition assignment}
+      \For(\tcp*[f]{For each partition assignment}){$i\in [1..k]$} {
+         $R[i]\leftarrow R[i] \cup \{a\}$\tcp*{Assign attribute}
+         \If(\tcp*[f]{If query cost is lower}){$L(R, B, Q)<c$}{
+            $c\leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
+            $j\leftarrow i$\tcp*{Update the best partition}
+         }
+         $R[i]\leftarrow R[i] \setminus \{a\}$\tcp*{Un-assign attribute}
+      }
+      $R[j]\leftarrow R[j] \cup \{a\}$\tcp*{Assign to best partition}
+   }
+   \lIf(\tcp*[f]{If solution infeasible}){$H(R, B, Q)>\alpha$}{\textbf{break}}
+   \If(\tcp*[f]{If solution has lower cost}){$L(R, B, Q)<c^*$}{
+     $c^* \leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
+     $\mathcal{P}(B)\leftarrow R$\tcp*{Update the best partitioning}
+   } 
+}
+\Return $\mathcal{P}(B)$ \tcp*{Final set of sub-blocks}
+\end{algorithm}
+
+\clearpage
+\newpage
 \subsection{Overlapping Attributes}
@@
 
+\subsubsection{Heuristic Solution}
+
+\paragraph*{Approach.$\,$} We start the algorithm with a   partitioning based
+on what queries we have seen. Every query gets its own sub-block. This is the
+``ideal'' partitioning, because the I/O cost would be minimized for every
+query that we would have seen. The algorithm iteratively combines the two
+partitions that are closest together.  After each combination of partitions,
+the algorithm calculate the storage overhead for the partitioning. The
+algorithm stops when the  storage cost is below some specified threshold.  The
+result is the block partitioning.
+
+\begin{algorithm}[ht]
+\scriptsize
+\caption{Algorithm for partitioning blocks into sub-blocks with overlapping attributes.}
+\label{alg:overlappingP}
+\KwData{$B$: block, $Q$: set of queries}
+$\mathcal{P}(B) \leftarrow \{q.A: q\in Q\}$ \tcp*{Each query gets its own sub-block}
+\While(\tcp*[f]{Until storage overhead is below $\alpha$}){$H(\mathcal{P},B) > \alpha$}{
+  $c^{*}\gets \infty $ \tcp*{Lowest cost over all sub-block pairs}
+  $(b_x,b_y)\gets (\emptyset,\emptyset)$ \tcp*{Sub-block pair with the lowest cost}
+  \For(\tcp*[f]{For each pair of blocks}){$\{b_i,b_j\}\in\mathcal{P}(B)$}{
+    $\mathcal{P'}(B) \leftarrow \mathcal{P}(B) \setminus \{b_i, b_j\} \cup \{b_i \cup b_j\}$\\
+    $c\gets \frac{L(\mathcal{P}',B,Q)-L(\mathcal{P},B,Q)}{H(\mathcal{P},B)-H(\mathcal{P}',B)}$\tcp*{Cost of merge}
+    \If(\tcp*[f]{Cost is lower}){$c<c^{*}$}{
+        $c^{*}\gets c$\tcp*{Update the lowest cost}
+        $(b_x,b_y)\gets (b_i,b_j)$\tcp*{Update the best pair}
+    }
+  }
+  $\mathcal{P}(B) \leftarrow \mathcal{P}(B) \setminus \{b_x, b_y\} \cup \{b_x \cup b_y\}$ \\
+}
+\Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
+\end{algorithm}
+
 \section{Evaluation}
+
+ \begin{figure*}[ht!]
+ \centerline{\begin{tabular}{c@{ }c@{ }c}
+ \includegraphics[width=0.33\textwidth]{figures/RunningTimeVsNumAttributes.pdf} &
+ \includegraphics[width=0.33\textwidth]{figures/RunningTimeVsNumQueryKinds.pdf} &
+ \includegraphics[width=0.33\textwidth]{figures/RunningTimeVsStorageOverheadThreshold.pdf}\\
+ \includegraphics[width=0.33\textwidth]{figures/QueryIOVsNumAttributes.pdf} &
+ \includegraphics[width=0.33\textwidth]{figures/QueryIOVsNumQueryKinds.pdf} &
+ \includegraphics[width=0.33\textwidth]{figures/QueryIOVsStorageOverheadThreshold.pdf}\\
+ \includegraphics[width=0.33\textwidth]{figures/StorageOverheadVsNumAttributes.pdf} &
+ \includegraphics[width=0.33\textwidth]{figures/StorageOverheadVsNumQueryKinds.pdf} &
+ \includegraphics[width=0.33\textwidth]{figures/StorageOverheadVsStorageOverheadThreshold.pdf}\\
+ (a)Num Attributes & (b) NumQueryKind &
+ (c) StorageOverheadThreshold \\
+ \end{tabular}}
+\vspace*{1mm}
+ \caption{Running time, QueryIO, and StorageOverhead.}
+ \label{fig:results}
+ \end{figure*}
+
 \section{Related Work}
@@
 \begin{alltt}\scriptsize
-- H2O \cite{alagiannis14}
-- HYRISE~\cite{grund10}).%
+Gedik et al.~\cite{gedik14}
+H2O \cite{alagiannis14}
+HYRISE~\cite{grund10}
+Bornea et al.~\cite{bornea13}
 \end{alltt}
-\section{Conclusion}
 
-%
-% The following two commands are all you need in the
-% initial runs of your .tex file to
-% produce the bibliography for the citations in your paper.
 \bibliographystyle{abbrv}
-\bibliography{main}  % sigproc.bib is the name of the Bibliography in this case
-
+\bibliography{paper}  
 \end{document}

### 2014-09-06 docs/vldbj/text/problem.tex

--- 2014-09-01 docs/edbt2015/text/paper.tex
@@
-\input{config}
+\section{Problem Description}
 
-\begin{document}
+The problem is to determine a partitioning of the disk blocks into sub-blocks
+that would minimize query I/O. Of course, partitioning into sub-blocks results
+in an increase in storage cost. The storage cost must stay below some
+specified limit.
+
+There are two possibilities: \emph{overlapping} and \emph{non-overlapping}
+partitioning. In the overlapping case, attributes can be replicated in the
+sub-blocks. The non-overlapping case is a true partition.
+
+We start by defining the cost model, which formalizes the query I/O and storage
+costs.
+
+\subsection{Cost Model}
+Let $Q$ be the query workload, where each query $q\in Q$ accesses a set of
+attributes $q.A$ and traverses parts of the graph for the time range
+$q.T=[q.t_s,q.t_e]$. We denote the set of all attributes as $A$. Given a block
+$B$, we denote its time range as $B.T$, which is the union of the time ranges
+of its temporal neighborlists. Let $s(a)$ denote the size of an attribute $a$.
+We use $c_n(B)$ to denote the number of temporal neighborlists within block
+$B$ and $c_e(B)$ to denote the total number of edges in the temporal
+neighborlists within the block. We overload the notation for block size and
+use $s(B)$ to denote the size of a block $B$. We have: 
+\begin{equation}
+s(B) = c_e(B) \cdot \Big(16 + \sum_{a\in B.A} s(a)\Big) + c_n(B) \cdot 12  
+\end{equation}
+Here, $16$ corresponds to the cost of storing the edge id and the timestamp,
+and $12$ corresponds to the cost of storing the head vertex ($8$ bytes) plus
+the number of entries ($4$ bytes) for a temporal neighborlist.
+
+Our goal is to create a potentially overlapping partitioning of a block,
+denoted by $\mathcal{P}(B)$. We call the partitions \emph{sub-blocks}. We have
+$\bigcup_{B'\in \mathcal{P}(B)} B'.A = A$. We aim to find the function
+$\mathcal{P}$ that minimizes the query I/O over $B$, while keeping the
+relative storage overhead beyond a limit, say $1+\alpha$ times the original.
+If we represent the query I/O as $L(\mathcal{P}, B)$ and the relative storage
+overhead as $H(\mathcal{P}, B)$, our goal is to find:
+\begin{equation}
+\mathcal{P} \leftarrow \mbox{argmin}_{\{\mathcal{P}: H(\mathcal{P}(B)) < \alpha\}} L(\mathcal{P},B)
+\end{equation}
+
+The storage overhead can be formalized as follows for the non-overlapping case:
+\begin{equation}
+H(\mathcal{P}, B) = (|\mathcal{P}(B)|-1)\cdot\Big(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\Big) 
+\end{equation}
+
+For the general case (including overlapping), we can formulate the storage
+overhead as follows:
+\begin{equation}
+H(\mathcal{P}, B) = \frac{\sum_{B'\in \mathcal{P}(B)} s(B')}{s(B)} - 1 
+\end{equation}
+
+%TODO: note about why we have two different cases
+
+\subsubsection{Query I/O}
+
+Let $m$ be a function that maps a query $q$ to the set of sub-blocks that are
+accessed to satisfy it for a relevant block $B$ under a given partitioning
+$\mathcal{P}$. For the case of non-overlapping attributes, we have:
+\begin{equation}
+m(\mathcal{P}, B, q) = \{B': B'\in \mathcal{P}(B) \wedge q.A \cap B'.A \ne \emptyset\}  
+\end{equation}
+
+For the case of overlapping attributes, we use a simple heuristic to define the
+set of sub-blocks to be used for answering the query.
+Algorithm~\ref{alg:greedyM} captures it. The idea is to select sub-blocks that
+bring the highest relative marginal gain, which is defined as the size of
+attribute data that contributes to the query result, relative to the sub-block
+size. While computing the marginal gain, attributes that are covered by
+sub-blocks that are already selected are not considered.
 %
-% --- Author Metadata here ---
-\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
-%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
-%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
-% --- End of Author Metadata ---
+\begin{algorithm}[ht]
+\scriptsize
+\caption{m-overlapping($\mathcal{P}, B, q$)}
+\label{alg:greedyM}
+\KwData{$\mathcal{P}$: partitioning function, $B$: block, $q$: query}
+$S\leftarrow \emptyset; R\leftarrow \emptyset$ \tcp*{Selected attributes; Resulting sub-blocks}
+\While(\tcp*[f]{While unselected attributes remain}){$S \subset q.A$}{
+  $B' \leftarrow \mbox{argmax}_{B'\in\mathcal{P}(B)\setminus R} \sum_{a\in B'.A \cap q.A \setminus S} \frac{c_e(B') \cdot s(a)}{s(B')}$
+  $S \leftarrow S \cup B'.A$\tcp*{Extend the selected attributes}
+  $R\leftarrow R \cup B'$\tcp*{Extend the selected sub-blocks}
+}
+\Return R \tcp*{Final set of sub-blocks covering the query attributes}
+\end{algorithm} 
+%
+The query I/O cost for a block is then given by:
+\begin{equation}
+L(\mathcal{P}, B) = \sum_{q\in Q} w(q)\cdot\mathbf{1}(q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} \!\!s(B')
+\end{equation}
 
-\title{Adaptive Disk Storage for Interaction Graphs}
+Here, $w(q)$ represents the query weight.
 
-\numberofauthors{2} 
-\author{
-\alignauthor
-Robert Soul\'{e}\\
-       \affaddr{University of Lugano}\\
-       \email{robert.soule@usi.ch}  
-\alignauthor
-Bu\u{g}ra Gedik\\
-       \affaddr{Bilkent University}\\
-       \email{bgedik@cs.bilkent.edu.tr}
-}
-
-\maketitle
-\begin{abstract}
-  \input{abstract}            
-\end{abstract}
-
-% A category with the (minimum) three required fields
-\category{H.4}{Information Systems Applications}{Miscellaneous}
-%A category including the fourth, optional field follows...
-\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
-
-\terms{Theory}
-
-\keywords{ACM proceedings, \LaTeX, text tagging}
-
-\section{Introduction}
-
-\section{Related Work}
-
-\begin{alltt}\scriptsize
-Gedik et al.~\cite{gedik14}
-H2O \cite{alagiannis14}
-HYRISE~\cite{grund10}
-Bornea et al.~\cite{bornea13}
-\end{alltt}
-
-\bibliographystyle{abbrv}
-\bibliography{paper}  
-\end{document}
+%\paragraph{\textbf{Other useful equations}}
+%\noindent The following are unused, but may be needed for the partitioning
+%algorithm.
+%
+%Let $f: \mathbb{R}\times A \rightarrow \mathbb{N}$ be a function that maps a
+%attribute-timestamp pair $(a, t)$ to the frequency of queries that access
+%attribute $a$ for time $t$. More formally:
+%\begin{equation}
+%f(a,t) = \sum_{q\in Q} \textbf{1}(a\in q.A \wedge t\in q.T)
+%\end{equation}
+%
+%Let $T$ represent a time range $[T.s, T.e]$. We denote the frequency of
+%attribute access for the time interval $T$ for attribute $a$ as follows:
+%\begin{equation}
+%f(a, T) = \int_{T.s}^{T.e} f(a, t)\cdot dt =  \sum_{q\in Q} |q.T \cap T| 
+%\end{equation}

### 2014-09-06 code/libcommon/include/intergdb/common/SystemConstants.h
fix the system constants

@@ -9,8 +9,8 @@ namespace intergdb { namespace common
       static const int timestampSize = 8;      
       static const int headVertexSize = 8;    
       static const int numEntriesSize = 4;
-      static const int numberOfEdgesInABlock = 48;  // TODO: correct 
-      static const int numberOfNeighborListsInABlock = 8; // TODO: correct
+      static const int numberOfEdgesInABlock = 24;
+      static const int numberOfNeighborListsInABlock = 8;
       SystemConstants() = delete;
   };
 } }

### 2014-09-06 code/libintergdb/include/intergdb/core/EndianHelper.h
fix typo

@@ -3,8 +3,8 @@
 
 #ifdef __MACH__
 #include <machine/endian.h>
-elif defined __linux__
-#include <mendian.h>
+#elif defined __linux__
+#include <endian.h>
 #endif
 #include <inttypes.h>
 #include <netinet/in.h>

### 2014-09-09 docs/edbt2015/text/paper.tex
removed text
outline

--- 2014-09-01 docs/edbt2015/text/paper.tex
@@
 
+\section{System Overview}
+
+\begin{alltt}\scriptsize
+- Intro
+    - What is an interaction graphs
+    - Relationship graphs vs interaction graphs 
+        - interactions get big
+- System overview
+    - How interaction graph are currently stored
+    - Introduce the rail layout
+- Layout or partitioning problem 
+    - Cost model
+- Optimal Solutions
+- Heuristic solutions
+\end{alltt}
+
 \section{Related Work}
@@
 
+\section{Conclusion}
+
 \bibliographystyle{abbrv}

### 2014-09-09 docs/edbt2015/text/paper.tex
more

--- 2014-09-01 docs/edbt2015/text/paper.tex
@@
 
-\title{Adaptive Disk Storage for Interaction Graphs}
+\title{Railway: Adaptive Disk Storage for Interaction Graphs}
 
@@
 
+\section{System Overview}
+
+\begin{alltt}\scriptsize
+- Intro
+    - What is an interaction graphs
+    - Relationship graphs vs interaction graphs 
+        - interactions get big
+- System overview
+    - How interaction graph are currently stored
+    - Introduce the rail layout
+- Layout or partitioning problem 
+    - Cost model
+- Optimal Solutions
+- Heuristic solutions
+- Evaluation
+\end{alltt}
+
 \section{Related Work}

### 2014-09-10 docs/edbt2015/text/introduction.tex
added old intro

--- 2014-08-07 docs/cidr2015/text/main.tex
@@
-\input{config}
+\section{Introduction}\label{sec:introduction}
 
-%\documentclass{sig-alternate}
-
-\begin{document}
-%
-% --- Author Metadata here ---
-\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
-%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
-%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
-% --- End of Author Metadata ---
-
-\title{Adaptive Disk Storage for Interaction Graphs}
-
-\numberofauthors{2}
-
-\author{
-% 1st. author
-\alignauthor
-Robert Soul\'{e}\\
-       \affaddr{University of Lugano}\\
-       \affaddr{Faculty of Informatics}\\
-       \affaddr{Lugano 6904, Switzerland}\\
-       \email{robert.soule@usi.ch}
-% 2nd. author
-\alignauthor
-Bu\u{g}ra Gedik\\
-       \affaddr{Bilkent University}\\
-       \affaddr{Department of Computer Engineering}\\
-       \affaddr{Bilkent, Ankara 06800, Turkey}
-       \email{bgedik@cs.bilkent.edu.tr}
-}
-
-\maketitle
-\begin{abstract}
-TODO
-\end{abstract}
-
-% A category with the (minimum) three required fields
-\category{H.4}{Information Systems Applications}{Miscellaneous}
-%A category including the fourth, optional field follows...
-\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
-
-\terms{Theory}
-
-\keywords{ACM proceedings, \LaTeX, text tagging}
-
-\section{Introduction}
-
-Graph databases, and \emph{interaction graph} databases in
-particular, are a critical component for many of today's most popular
-applications.  Graph databases store entities as vertices, relationships as
-edges, and have data associated with either vertices or edges, called
-\emph{attributes} or \emph{properties}.  Interaction graphs differ from general
-graphs in that they are \emph{append-only}, and edges are assigned a
-\emph{temporal value}.  Thus, interaction graphs encode the relationships over
-time between entities in their structure, making them ideal for use in social
-networking, telecommunications, or modeling the world-wide web.
+Graph databases, and \emph{interaction graph} databases in particular, are a
+critical component for many of today's most popular applications.  Graph
+databases store entities as vertices, relationships as edges, and have data
+associated with either vertices or edges, called \emph{attributes} or
+\emph{properties}.  Interaction graphs differ from general graphs in that they
+are \emph{append-only}, and edges are assigned a \emph{temporal value}.  Thus,
+interaction graphs encode the relationships over time between entities in
+their structure, making them ideal for use in social networking,
+telecommunications, or modeling the world-wide web.
 
@@
 might access attributes $a_1$ and $a_2$, while queries $q_2$, $q_3$ and $q_4$
-access attributes $a_3$ and $a_4$. Because interaction graphs are temporal, the
-co-access correlations for the attributes can vary for different
-temporal regions.  Moreover, the co-access correlations might be unknown at the
+access attributes $a_3$ and $a_4$. Because interaction graphs are temporal,
+the co-access correlations for the attributes can vary for different temporal
+regions.  Moreover, the co-access correlations might be unknown at the
 insertion time, but be discovered later, when the workload is known.
@@
 table-based, relational databases, this fact has led database designers to
-develop alternative approaches for storage layout: row-oriented storage is more
-efficient when queries access many attributes from a small number of records,
-and column-oriented storage is more efficient when queries access a small number
-of attributes from many records~\cite{stonebraker05}.  Unfortunately, although
-interaction graph databases, like relational databases, are the target of
-diverse query workloads, there is no clear correspondence to a row-oriented or
-column-oriented storage layout.
+develop alternative approaches for storage layout: row-oriented storage is
+more efficient when queries access many attributes from a small number of
+records, and column-oriented storage is more efficient when queries access a
+small number of attributes from many records~\cite{stonebraker05}. 
+Unfortunately, although interaction graph databases, like relational
+databases, are the target of diverse query workloads, there is no clear
+correspondence to a row-oriented or column-oriented storage layout.
 
 Typically, an interaction graph database is organized as blocks of temporal
-neighbor lists.  Attributes can be stored in two ways, either separately (e.g., in a relational
-table), or locally with the temporal neighbor lists.  If they are stored
-separately, then the graph database cannot take advantage of locality
-optimizations performed for block organization.  As discussed in Gedik et
-al.~\cite{gedik14}, the database must go back and forth between the disk blocks
-to access the edge attributes.  On the other hand, if attributes are stored
-locally in the disk blocks containing the graph structure, then there can be
-significant overhead due to disk I/O if only a few attributes are needed to
-answer a query.
+neighbor lists.  Attributes can be stored in two ways, either separately
+(e.g., in a relational table), or locally with the temporal neighbor lists. 
+If they are stored separately, then the graph database cannot take advantage
+of locality optimizations performed for block organization.  As discussed in
+Gedik et al.~\cite{gedik14}, the database must go back and forth between the
+disk blocks to access the edge attributes.  On the other hand, if attributes
+are stored locally in the disk blocks containing the graph structure, then
+there can be significant overhead due to disk I/O if only a few attributes are
+needed to answer a query.
 
@@
 having two or more graph databases for certain time regions, each containing a
-different subset of the attributes, but with a link between them in case a query
-needs to access both.
+different subset of the attributes, but with a link between them in case a
+query needs to access both.
 
@@
 % and HYRISE~\cite{grund10}).%
-
-\section{Problem Description}
-
-The problem is to determine a partitioning of the disk blocks into sub-blocks
-that would minimize query I/O. Of course, partitioning into sub-blocks results
-in an increase in storage cost. The storage cost must stay below some
-specified limit.
-
-There are two possibilities: \emph{overlapping} and \emph{non-overlapping}
-partitioning. In the overlapping case, attributes can be replicated in the
-sub-blocks. The non-overlapping case is a true partition.
-
-We start by defining the cost model, which formalizes the query I/O and storage
-costs.
-
-\subsection{Cost Model}
-Let $Q$ be the query workload, where each query $q\in Q$ accesses a set of
-attributes $q.A$ and traverses parts of the graph for the time range
-$q.T=[q.t_s,q.t_e]$. We denote the set of all attributes as $A$. Given a block
-$B$, we denote its time range as $B.T$, which is the union of the time ranges
-of its temporal neighborlists. Let $s(a)$ denote the size of an attribute $a$.
-We use $c_n(B)$ to denote the number of temporal neighborlists within block
-$B$ and $c_e(B)$ to denote the total number of edges in the temporal
-neighborlists within the block. We overload the notation for block size and
-use $s(B)$ to denote the size of a block $B$. We have: 
-\begin{equation}
-s(B) = c_e(B) \cdot \Big(16 + \sum_{a\in B.A} s(a)\Big) + c_n(B) \cdot 12  
-\end{equation}
-Here, $16$ corresponds to the cost of storing the edge id and the timestamp,
-and $12$ corresponds to the cost of storing the head vertex ($8$ bytes) plus
-the number of entries ($4$ bytes) for a temporal neighborlist.
-
-Our goal is to create a potentially overlapping partitioning of a block,
-denoted by $\mathcal{P}(B)$. We call the partitions \emph{sub-blocks}. We have
-$\bigcup_{B'\in \mathcal{P}(B)} B'.A = A$. We aim to find the function
-$\mathcal{P}$ that minimizes the query I/O over $B$, while keeping the
-relative storage overhead beyond a limit, say $1+\alpha$ times the original.
-If we represent the query I/O as $L(\mathcal{P}, B)$ and the relative storage
-overhead as $H(\mathcal{P}, B)$, our goal is to find:
-\begin{equation}
-\mathcal{P} \leftarrow \mbox{argmin}_{\{\mathcal{P}: H(\mathcal{P}(B)) < \alpha\}} L(\mathcal{P},B)
-\end{equation}
-
-The storage overhead can be formalized as follows for the non-overlapping case:
-\begin{equation}
-H(\mathcal{P}, B) = (|\mathcal{P}(B)|-1)\cdot\Big(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\Big) 
-\end{equation}
-
-For the general case (including overlapping), we can formulate the storage
-overhead as follows:
-\begin{equation}
-H(\mathcal{P}, B) = \frac{\sum_{B'\in \mathcal{P}(B)} s(B')}{s(B)} - 1 
-\end{equation}
-
-%TODO: note about why we have two different cases
-
-\subsubsection{Query I/O}
-
-Let $m$ be a function that maps a query $q$ to the set of sub-blocks that are
-accessed to satisfy it for a relevant block $B$ under a given partitioning
-$\mathcal{P}$. For the case of non-overlapping attributes, we have:
-\begin{equation}
-m(\mathcal{P}, B, q) = \{B': B'\in \mathcal{P}(B) \wedge q.A \cap B'.A \ne \emptyset\}  
-\end{equation}
-
-For the case of overlapping attributes, we use a simple heuristic to define the
-set of sub-blocks to be used for answering the query.
-Algorithm~\ref{alg:greedyM} captures it. The idea is to select sub-blocks that
-bring the highest relative marginal gain, which is defined as the size of
-attribute data that contributes to the query result, relative to the sub-block
-size. While computing the marginal gain, attributes that are covered by
-sub-blocks that are already selected are not considered.
-%
-\begin{algorithm}[ht]
-\scriptsize
-\caption{m-overlapping($\mathcal{P}, B, q$)}
-\label{alg:greedyM}
-\KwData{$\mathcal{P}$: partitioning function, $B$: block, $q$: query}
-$S\leftarrow \emptyset; R\leftarrow \emptyset$ \tcp*{Selected attributes; Resulting sub-blocks}
-\While(\tcp*[f]{While unselected attributes remain}){$S \subset q.A$}{
-  $B' \leftarrow \mbox{argmax}_{B'\in\mathcal{P}(B)\setminus R} \sum_{a\in B'.A \cap q.A \setminus S} \frac{c_e(B') \cdot s(a)}{s(B')}$
-  $S \leftarrow S \cup B'.A$\tcp*{Extend the selected attributes}
-  $R\leftarrow R \cup B'$\tcp*{Extend the selected sub-blocks}
-}
-\Return R \tcp*{Final set of sub-blocks covering the query attributes}
-\end{algorithm} 
-%
-The query I/O cost for a block is then given by:
-\begin{equation}
-L(\mathcal{P}, B) = \sum_{q\in Q} \mathbf{1}(q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} \!\!s(B')
-\end{equation}
-
-%\paragraph{\textbf{Other useful equations}}
-%\noindent The following are unused, but may be needed for the partitioning
-%algorithm.
-%
-%Let $f: \mathbb{R}\times A \rightarrow \mathbb{N}$ be a function that maps a
-%attribute-timestamp pair $(a, t)$ to the frequency of queries that access
-%attribute $a$ for time $t$. More formally:
-%\begin{equation}
-%f(a,t) = \sum_{q\in Q} \textbf{1}(a\in q.A \wedge t\in q.T)
-%\end{equation}
-%
-%Let $T$ represent a time range $[T.s, T.e]$. We denote the frequency of
-%attribute access for the time interval $T$ for attribute $a$ as follows:
-%\begin{equation}
-%f(a, T) = \int_{T.s}^{T.e} f(a, t)\cdot dt =  \sum_{q\in Q} |q.T \cap T| 
-%\end{equation}
-
-\section{Partitioning}
-This section is still a work-in-progress. We are trying to find the algorithms.
-
-\subsection{Non-Overlapping Attributes}
-
-\paragraph*{Problem.$\,$} \emph{Find a true partitioning of attributes that
-minimizes the query I/O and bounds the storage cost by some upper limit.}
-
-\subsubsection{Integer Linear Program Formulation}\label{subsubsec:nov-ilp}
-We present an ILP formuation of the problem. For this purpose, we define a 
-number of binary ($0$ or $1$) variables: 
-\begin{itemize}
-\item $x_{a,p}$: $1$ if attribute $a$ is in partition $p$, $0$ otherwise.
-\item $y_{p,q}$: $1$ if partition $p$ is used by query $q$, $0$ otherwise.
-\item $z_{a,p,q}$: $1$ if partition $p$ is used by query $q$ and attribute $a$
-is in partition $p$, $0$ otherwise.
-\item $u_{p}$: $1$ if partition $p$ is assigned at least $1$ attribute, $0$ otherwise.
-\end{itemize}
-
-Each of these variables serve a purpose:
-\begin{itemize}
-\item $x$s define the attribute-to-partition assignment.
-\item $y$s help formulate the query I/O contribution
-of partitions, excluding their assigned attributes.
-\item $z$s help formulate the query I/O contribution
-of partitions, only considering their assigned attributes.
-\item $u$s help formulate the storage overhead requirement.
-\end{itemize}
-
-We also define a helper notation for representing whether a variable is
-accessed by a query or not: $q(a)\equiv \mathbf{1}(a \in q.A)$, where
-$\mathbf{1}$ is the indicator function. We assume that there are a maximum of
-$k$ partitions. $k$ can be taken as $|A|$, as there cannot be more partitions
-than there are attributes.
-
-We are now ready to state the ILP formulation. We start with the objective
-function, that is the total query I/O:
-\begin{eqnarray}
-\sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot
-y_{p,q}\nonumber\\ 
-&+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q}\label{eq:no-obj}
-\end{eqnarray}
-
-In Eq.~\ref{eq:no-obj}, we simply sum for each query and each partition, and
-add the I/O cost of reading in the structural information found in a
-sub-block, if the partition is used by the query. We then sum over each
-attribute as well, and add the I/O cost of reading in the attributes. Note
-that $z_{a,p,q}$ could have been  replaced with $x_{a,p}\cdot y_{p,q}$, but
-that would make the objective function non-linear.
-
-We are now ready to state our constraints. Our first constraint is that, each
-attribute must be assigned to a single partition. Formally:
-\begin{eqnarray}
-\forall_{a\in A}, \sum_{p=1}^{k} x_{a,p} = 1
-\end{eqnarray}
-
-Our second constraint is that, if a query $q$ contains an attribute $a$
-assigned to a partition $p$, then partition $p$ is used by the query, i.e.,
-$y_{p,q}=1$. In essence, we want to state: $\forall_{\{p,q\}\in [1..k]\times
-Q}, y_{p,q} = \mathbf{1}(\sum_{a\in A} q(a)\cdot x_{a,p}>0)$.
-
-In order to formulate this constraint, we use the following ILP  construction:
-Assume we have two variables, $\beta_1$ and $\beta_2$, where $\beta_2\in[0,1]$
-and $\beta_1\geq 0$. We want to implement the following constraint: $\beta_2 =
-\mathbf{1}(\beta_1 > 0)$. This could be expressed as a linear constraint as
-follows, where $K$ is a large constant guaranteed to be larger than $\beta_1$
-for practical purposes:
-\begin{eqnarray}
-&& \beta_1 - \beta_2 \geq 0\nonumber\\
-&& K\cdot\beta_2 - \beta_1 \geq 0\label{eq:beta-ilp}
-\end{eqnarray}
-
-We now apply this construction to our second constraint, where
-$\beta_1=\sum_{a\in A} q(a)\cdot x_{a,p}$ and $\beta_2=y_{p,q}$. This results
-in the following linear constraints:
-\begin{eqnarray}
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  \sum_{a\in A} q(a)\cdot x_{a,p} - y_{p,q} \geq 0 \nonumber\\
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  K\cdot y_{p,q} - \sum_{a\in A} q(a)\cdot x_{a,p}  \geq 0 
-\end{eqnarray}
-
-Our third constraint is that, if an attribute $a$ is assigned to a partition
-$p$, and partition $p$ is used by a query $q$, then the corresponding $z$
-variable must be set to $1$. That is, we want: $\forall_{\{a,p,q\}\in A\times
-[1..k]\times Q}, z_{a,p,q}=\mathbf{1}(x_{a,p} = y_{p,q} = 1)$. We express this
-as a linear  constraint, as follows:
-\begin{eqnarray}
-\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
-    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\label{eq:no-z}
-\end{eqnarray}
-
-In Eq.~\ref{eq:no-z}, when the $x$ and $y$ variables are both $1$, the  $z$
-variable is simply forced to be $1$. Otherwise, the $z$ variable can be either
-$0$ or $1$, but since the $z$ variables appear in the objective fucntion as
-positive terms, the solver will set them to $0$, which is what we want.
-
-Our fourth constraint is that, if a partition is non-empty, then its
-corresponding $u$ variable must be set to $0$. In other words,  we want
-$\forall_{p\in[1..k]}, u_p = \mathbf{1}(\sum_{a\in A} x_{a,p}>0)$. This is
-expressed as linear constraints, as follows:
-\begin{eqnarray}
-\forall_{p\in[1..k]},
-    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
-\forall_{p\in[1..k]},
-    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \label{eq:no-u}
-\end{eqnarray}
-
-Eq.~\ref{eq:no-u} uses the same construction as the second constraint, where
-$\beta_1=\sum_{a\in A} x_{a,p}$ and $\beta_2=u_p$.
-
-Our fifth, and the last, constraint deals with the storage overhead. We want to
- make sure that the storage overhead does not go over $\alpha$. The storage
-overhead depends on the number of partitions used. That means that the only 
-ILP variables it depends on is the $u$s. In particular, the number of
-partitions used is given by $\sum_{p=1}^{k} u_p$. This results in the
-following linear constraint:
-\begin{equation}
-\sum_{p=1}^{k} u_p \leq 1 + \frac{\alpha}
-  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}}
-\end{equation}
-
-% NOTE: In the simulation we will have query weights that will factor into the objective function
-\begin{figure}[!t]
-\begin{mdframed}
-\begin{eqnarray}
-\text{minimize}  
-    \sum_{q\in Q} \sum_{p=1}^{k} \!\!&&\!\! (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
-    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
-\text{subject to}&&\nonumber\\
-\forall_{a\in A}, 
-    && \sum_{p=1}^{k} x_{a,p} = 1\nonumber\\
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  \sum_{a\in A} q(a)\cdot x_{a,p} - y_{p,q} \geq 0 \nonumber\\
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  K\cdot y_{p,q} - \sum_{a\in A} q(a)\cdot x_{a,p}  \geq 0 \nonumber\\
-\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
-    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\nonumber\\
-\forall_{p\in[1..k]},
-    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
-\forall_{p\in[1..k]},
-    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
-&& \sum_{p=1}^{k} u_p \leq 1 + \frac{\alpha}
-  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}} \nonumber
-\end{eqnarray}
-\end{mdframed}
-\caption{ILP formulation for the non-overlapping partitioning}
-\label{fig:nov-ilp}
-\end{figure}
-
-The final ILP formulation for the non-overlapping partitioning is given in
-Figure~\ref{fig:nov-ilp}.
-
-\subsection{Overlapping Attributes}
-
-\paragraph*{Problem.$\,$} \emph{Find an overlapping partitioning of attributes
-that minimizes the query I/O and bounds the storage cost by some upper
-limit.}
-
-\subsubsection{Integer Linear Program Formulation}
-We present an ILP formuation of the problem, as we did for the case of
-non-overlapping partitions in Section~\ref{subsubsec:nov-ilp}. We use the same
-set of variables and the same objective function. However, the formulation of
-the constraints differ.
-
-Our first constraint is that, each attribute must be assigned to at leat one
-partition. Formally:
-\begin{eqnarray}
-\forall_{a\in A}, \sum_{p=1}^{k} x_{a,p} \geq 1
-\end{eqnarray}
-
-As our second constraint, we require that for each attribute used by a query,
-there needs to be a partition that is used by that query and that contains the
-attribute in question. Formally:
-\begin{eqnarray}
-\forall_{\{a,q\}\in A\times Q}, \sum_{p=1}^{k} z_{a,p,q} \geq q(a) 
-\end{eqnarray}
-
-\begin{sloppypar}
-As our third constraint, we require that if a query is using an attribute from
-a partition, then that partition must contain the attribute. I.e., we need
-to link the $z$ variables with the $x$ variables as
-$\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, (z_{a,p,q} = 1) \implies 
-(x_{a,p} = 1)$. This can be state as linear constraints:
-\begin{eqnarray}
-\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, x_{a,p} - z_{a,p,q} \geq 0 
-\end{eqnarray}
-\end{sloppypar}
-
-As our fourth constraint, we require that if a query is using at least one
-attribute from a partition, then that partition must be used by the query.
-I.e., we need to link the $z$ variables with the $y$ variables as
-$\forall_{\{p,q\}\in [1..k]\times Q}, y_{p,q} = \mathbf{1}(\sum_{a\in A}
-z_{a,p,q}>0)$. As before, we use the ILP construction from
-Eq.~\ref{eq:beta-ilp} for this, where $\beta_2=y_{p,q}$ and $\beta_1 =
-\sum_{a\in A} z_{a,p,q}$. We get:
-\begin{eqnarray}
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  \sum_{a\in A} z_{a,p,q} - y_{p,q} \geq 0 \nonumber\\
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  K\cdot y_{p,q} - \sum_{a\in A} z_{a,p,q} \geq 0 
-\end{eqnarray}
-
-Our fifth constraint is that, if an attribute $a$ is assigned to a partition
-$p$, and partition $p$ is used by a query $q$, then the corresponding $z$
-variable must be set to $1$. This is same as the formulation for the
-non-overlapping case from Eq.~\ref{eq:no-z}.
-
-Our sixth constraint is that, if a partition is non-empty, then its
-corresponding $u$ variable must be set to $0$. Again, this is same as the
-formulation for the non-overlapping case from Eq.~\ref{eq:no-u}.
-
-Our seventh, and the last, constraint deals with the storage overhead. However,
-the storage overhead formulation for the overlapping case is different from 
-the one for the non-overlapping. This is because the overhead does not merely
-depend on the number of partitions, as attributes might have to be loaded
-multiple times from different partitions. As a result, we express the overhead
-using base variables as in the objective function. Formally:
-\begin{eqnarray}
-\sum_{p=1}^{k} (16\cdot c_e(B) &+& 12 \cdot c_n(B)) \cdot u_p  \nonumber \\ 
-+ \sum_{a\in A} s(a) \!\!&\cdot&\!\! c_n(B)\cdot x_{a,p} \leq s(B)\cdot (1+\alpha)
-\end{eqnarray}
-
-% NOTE: In the simulation we will have query weights that will factor into the objective function
-\begin{figure}[!t]
-\begin{mdframed}
-\begin{eqnarray}
-\text{minimize}  
-    \sum_{q\in Q} \sum_{p=1}^{k} \!\!&&\!\! (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
-    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
-\text{subject to}&&\nonumber\\
-\forall_{a\in A}, 
-    && \sum_{p=1}^{k} x_{a,p} \geq 1\nonumber\\
-\forall_{\{a,q\}\in A\times Q},
-    &&  \sum_{p=1}^{k} z_{a,p,q} \geq q(a) \nonumber\\
-\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, 
-    && x_{a,p} - z_{a,p,q} \geq 0 \nonumber\\
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  \sum_{a\in A} z_{a,p,q} - y_{p,q} \geq 0 \nonumber\\
-\forall_{\{p,q\}\in [1..k]\times Q}, 
-    &&  K\cdot y_{p,q} - \sum_{a\in A} z_{a,p,q}  \geq 0 \nonumber\\
-\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, 
-    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1 \nonumber\\
-\forall_{p\in[1..k]},
-    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
-\forall_{p\in[1..k]},
-    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
-\sum_{p=1}^{k} (16\cdot c_e(B) &+& 12 \cdot c_n(B)) \cdot u_p  \nonumber \\ 
-+ \sum_{a\in A} s(a) \!\!&\cdot&\!\! c_n(B)\cdot x_{a,p} \leq s(B)\cdot (1+\alpha)\nonumber
-\end{eqnarray}
-\end{mdframed}
-\caption{ILP formulation for the overlapping partitioning}
-\label{fig:ov-ilp}
-\end{figure}
-
-The final ILP formulation for the overlapping partitioning is given in
-Figure~\ref{fig:ov-ilp}.
-
-\section{Evaluation}
-\section{Related Work}
-
-\begin{alltt}\scriptsize
-- H2O \cite{alagiannis14}
-- HYRISE~\cite{grund10}).%
-\end{alltt}
-\section{Conclusion}
-
-%
-% The following two commands are all you need in the
-% initial runs of your .tex file to
-% produce the bibliography for the citations in your paper.
-\bibliographystyle{abbrv}
-\bibliography{main}  % sigproc.bib is the name of the Bibliography in this case
-
-\end{document}

### 2014-09-11 docs/edbt2015/text/introduction.tex
minor progress over intro

@@
 \section{Introduction}\label{sec:introduction}
+
+We are living in an ever more connected world, where the data generated by
+people, software systems, and the physical world is more accessible than
+before and is much larger in volume, variety, and velocity. In many
+application domains, such as telecommunications and social media, live data
+recording the interactions between people, systems, and the environment is
+available for analysis. This data often takes the form of a temporally
+evolving graph, where entities are the vertices and the interactions between
+them are the edges. We call such graphs \emph{interaction graphs}.
+
+Data analytics performed on interaction graphs can bring new business insights
+and improve decision making. For instance, the graph structure may represent
+the interactions in a social network, where finding communities in the graph
+can facilitate targeted advertising. In the telecommunications (telco) domain,
+call details records (CDRs) can be used to capture the call interactions
+between people, and locating closely connected groups of people can be used
+for generating promotions.
+
+Interaction graphs are temporal in nature, and more importantly, they are
+append-only. This is in contrast to relationship graphs, which are updated via
+insertion and deletion operations. An example of a relationship graph is a
+social network capturing the follower-followee relationship among users.
+Examples of interactions graphs include CDR graphs capturing calls between
+telco customers or mention graphs capturing interactions between users of a
+micro-blogging service. The append-only nature of the interaction graphs make
+storing them on disk a necessity. Furthermore, the analysis of this historical
+interaction data form an important aspect of the analytical landscape.
+
+The ability to efficiently support historical analysis over interaction graphs
+require effective solutions for the problem of \emph{data layout} on the disk.
+
+% Importance of the data layout on disk and locality 
+% Earlier work on blocks formed of temporal neighborlists
+
+% Attributes associated with interactions
+% I/O inefficiency due to attributes that are not accessed
+% Similar to the problem of row vs column stores
+
+% Challenges: finding the right partitioning
+% what if the access pattern changes over time
+% query access patterns are different over different time ranges
+
+% Solution: railway layout 
+% adaptive layout optimization
+% algorithms for optimal partitioning
+% heuristics & evaluation
 

### 2014-09-11 docs/edbt2015/text/introduction.tex
some more progress

@@
 The ability to efficiently support historical analysis over interaction graphs
-require effective solutions for the problem of \emph{data layout} on the disk.
+require effective solutions for the problem of \emph{data layout} on disk.
+Most graph algorithms are characterized by locality of
+access~\cite{localityLayout}, which has been taken advantage of co-locating
+edges in close proximity within the same disk blocks~\cite{gstoreThesis}. This
+ensures that once a disk block is loaded from disk into the main memory
+buffers, several edges can be used. The locality of reference is a result of
+the traversal-based nature of most of the graph algorithms.
 
-% Importance of the data layout on disk and locality 
-% Earlier work on blocks formed of temporal neighborlists
+In interaction graphs, the locality of access is even more pronounced. First, 
+the analysis to be performed on the interactions can be restricted to a
+temporal view of the graph, such as finding the influential users over a given
+week of interactions. This means that edges that are temporally close are to
+be accessed together. Second, traversals are again key to many algorithms,
+such as connected components, clustering coefficient, PageRank, etc. This
+means that edges that are close by in terms of the path between their incident
+vertices as well as their timestamps are to be located together with the same
+blocks.
+
+In our earlier work, we introduced an interaction graph database~\cite{gedik14}
+that works on this principle of access locality. It proposed a disk
+organization that consists of as set of blocks, each containing a list of
+\emph{temporal neighbor lists}. A temporal neighborlist contains a head vertex
+and a set of incident edges within a given time range. The layout optimizer
+aims at bringing together temporal neighborlists that are (i) close in terms
+of their temporal ranges, (ii) have many edges between them, and (iii) have
+few edges going into temporal neighborlists outside the block.
 
@@
 % Similar to the problem of row vs column stores
+
+Attributes can be stored in two ways, either separately
+(e.g., in a relational table), or locally with the temporal neighbor lists. 
+If they are stored separately, then the graph database cannot take advantage
+of locality optimizations performed for block organization.  As discussed in
+Gedik et al.~\cite{gedik14}, the database must go back and forth between the
+disk blocks to access the edge attributes.  On the other hand, if attributes
+are stored locally in the disk blocks containing the graph structure, then
+there can be significant overhead due to disk I/O if only a few attributes are
+needed to answer a query.
 
@@
 
-Typically, an interaction graph database is organized as blocks of temporal
-neighbor lists.  Attributes can be stored in two ways, either separately
-(e.g., in a relational table), or locally with the temporal neighbor lists. 
-If they are stored separately, then the graph database cannot take advantage
-of locality optimizations performed for block organization.  As discussed in
-Gedik et al.~\cite{gedik14}, the database must go back and forth between the
-disk blocks to access the edge attributes.  On the other hand, if attributes
-are stored locally in the disk blocks containing the graph structure, then
-there can be significant overhead due to disk I/O if only a few attributes are
-needed to answer a query.
-
 This paper presents an algorithm for adaptively optimizing disk block storage

### 2014-09-11 docs/edbt2015/text/heuristic.tex
paper skeleton

\section{Heuristic Solution}\label{sec:heuristic}

### 2014-09-11 docs/edbt2015/text/paper.tex
started intro, 'outlined' remaining parts
paper skeleton

@@
 
-\section{Introduction}
-
-\section{System Overview}
+\input{introduction}
 
@@
         - interactions get big
-- System overview
-    - How interaction graph are currently stored
-    - Introduce the rail layout
-- Layout or partitioning problem 
-    - Cost model
-- Optimal Solutions
-- Heuristic solutions
-- Evaluation
 \end{alltt}
 
-\section{Related Work}
-
-\begin{alltt}\scriptsize
-Gedik et al.~\cite{gedik14}
-H2O \cite{alagiannis14}
-HYRISE~\cite{grund10}
-Bornea et al.~\cite{bornea13}
-\end{alltt}
-
-\section{Conclusion}
+\input{system.tex}
+\input{problem.tex}
+\input{optimal.tex}
+\input{heuristic.tex}
+\input{evaluation.tex}
+\input{related.tex}
+\input{conclusion.tex}
 

### 2014-09-11 docs/edbt2015/text/system.tex
paper skeleton

\section{System Overview}\label{sec:system}
\begin{alltt}\scriptsize
- System overview
    - How interaction graph are currently stored
    - Introduce the rail layout
\end{alltt}

### 2014-09-11 docs/edbt2015/text/introduction.tex
some more progress

@@
 
-In our earlier work, we introduced an interaction graph database~\cite{gedik14}
+In our earlier work~\cite{gedik14}, we introduced an interaction graph database
 that works on this principle of access locality. It proposed a disk
@@
 
-% Attributes associated with interactions
-% I/O inefficiency due to attributes that are not accessed
-% Similar to the problem of row vs column stores
-
-Attributes can be stored in two ways, either separately
-(e.g., in a relational table), or locally with the temporal neighbor lists. 
-If they are stored separately, then the graph database cannot take advantage
-of locality optimizations performed for block organization.  As discussed in
-Gedik et al.~\cite{gedik14}, the database must go back and forth between the
-disk blocks to access the edge attributes.  On the other hand, if attributes
-are stored locally in the disk blocks containing the graph structure, then
-there can be significant overhead due to disk I/O if only a few attributes are
-needed to answer a query.
-
-% Challenges: finding the right partitioning
-% what if the access pattern changes over time
-% query access patterns are different over different time ranges
-
-% Solution: railway layout 
-% adaptive layout optimization
-% algorithms for optimal partitioning
-% heuristics & evaluation
-
-Graph databases, and \emph{interaction graph} databases in particular, are a
-critical component for many of today's most popular applications.  Graph
-databases store entities as vertices, relationships as edges, and have data
-associated with either vertices or edges, called \emph{attributes} or
-\emph{properties}.  Interaction graphs differ from general graphs in that they
-are \emph{append-only}, and edges are assigned a \emph{temporal value}.  Thus,
-interaction graphs encode the relationships over time between entities in
-their structure, making them ideal for use in social networking,
-telecommunications, or modeling the world-wide web.
+Many real world graph databases contain attributes. In the case of interaction
+graphs, the attributes can be considered as properties associated with the
+edges. Attributes can be stored in two ways, either separately (e.g., in a
+relational table), or locally with the temporal neighbor lists.  If they are
+stored separately, then the graph database cannot take advantage of locality
+optimizations performed for block organization. The database must go back and
+forth between the disk blocks to access the edge attributes.  On the other
+hand, if attributes are stored locally in the disk blocks containing the graph
+structure, then there can be significant overhead due to disk I/O if only a
+few attributes are needed to answer a query.
 
@@
 table-based, relational databases, this fact has led database designers to
-develop alternative approaches for storage layout: row-oriented storage is
-more efficient when queries access many attributes from a small number of
-records, and column-oriented storage is more efficient when queries access a
-small number of attributes from many records~\cite{stonebraker05}. 
-Unfortunately, although interaction graph databases, like relational
-databases, are the target of diverse query workloads, there is no clear
-correspondence to a row-oriented or column-oriented storage layout.
+develop alternative approaches for storage layout: row-oriented
+storage~\cite{rowOrg} is more efficient when queries access many attributes
+from a small number of records, and column-oriented storage~\cite{colOrg} is
+more efficient when queries access a small number of attributes from many
+records~\cite{stonebraker05}.  Unfortunately, although interaction graph
+databases, like relational databases, are the target of diverse query
+workloads, there is no clear correspondence to a row-oriented or
+column-oriented storage layout.
 
-This paper presents an algorithm for adaptively optimizing disk block storage
-for interaction graphs. The key idea behind the approach is a novel storage
-scheme, called a \emph{rail layout}.  Initially, attributes are stored with
-their neighbor lists in large blocks. Over time, the blocks are split into
-smaller-blocks that contain the neighbor lists, but only subsets of the
-attributes, depending on the query workload. Intuitively, this is almost like
-having two or more graph databases for certain time regions, each containing a
-different subset of the attributes, but with a link between them in case a
-query needs to access both.
+This paper presents an adaptive disk layout called the \emph{railway layout}
+and associated algorithms for optimizing disk block storage for interaction
+graphs. The key idea is to divide blocks into one or more sub-blocks, where
+each sub-block contains a subset of the attributes (potentially overlapping),
+but the entire graph structure is replicated within each sub-block. This way,
+a query can be answered completely by only reading the sub-blocks that contain
+the attributes of interest, reducing the overall I/O.
+
+There a number of challenges in achieving an effective adaptive layout. First,
+we need to find the partitioning of attributes that optimizes the query I/O.
+To address this, we model the problems of overlapping and non-overlapping
+attribute partitioning as mixed-integer linear programs (ILPs), and provide
+optimal solutions that minimize the I/O cost. Second, the query workload, and
+thus the attribute access pattern can change over time. For this purpose, our
+railway layout supports customization of the attribute partitioning of
+sub-blocks on a per-block basis. Third, such flexibility necessitates online
+configuration of attribute partitioning as the query workload evolves, which
+in turn requires fast algorithms for performing the attribute partitioning.
+For this purpose, we develop greedy heuristic algorithms for both overlapping
+and non-overlapping partitioning scenarios. These algorithms can scale to
+larger number of attributes, yet provide close to optimal query I/O
+performance. Finally, the railroad layout tradeoffs storage space to gain
+improved query I/O performance. The storage overhead is more pronounced for
+the case of overlapping partitioning. To address this, we limit the amount of
+storage  overhead that can be tolerated with an upper limit, and integrate
+this limit to both our ILP formulations, as well as our greedy heuristics.
+
+In summary, this paper makes the following contributions:
+\begin{itemize}
+\item We introduce the railroad layout for adaptive organization of interaction
+graphs on disk. 
+\item We introduce optimal ILP formulations for partitioning disk blocks into 
+sub-blocks with overlapping and non-overlapping attributes, given a query
+workload. Our formulation also support upper bounding the amount of storage
+overhead introduced as a result of the railroad layout.
+\item To support online adaptation, we develop greedy heuristics that can scale
+better compared to the ILP alternatives, yet achieve close to optimal query
+I/O.
+\item We provide an extensive experimental study comparing our approach to a
+few baseline alternatives.
+\end{itemize}
+
+The rest of the paper is organized as follows. Section...
 

### 2014-09-11 docs/edbt2015/text/related.tex

\section{Related Work}\label{sec:related}

\begin{alltt}\scriptsize
Gedik et al.~\cite{gedik14}
H2O \cite{alagiannis14}
HYRISE~\cite{grund10}
Bornea et al.~\cite{bornea13}
\end{alltt}

### 2014-09-11 docs/edbt2015/text/introduction.tex
first cut of the introduction

@@
 storing them on disk a necessity. Furthermore, the analysis of this historical
-interaction data form an important aspect of the analytical landscape.
+interaction data form an important part of the analytical landscape.
 
@@
 Most graph algorithms are characterized by locality of
-access~\cite{localityLayout}, which has been taken advantage of co-locating
-edges in close proximity within the same disk blocks~\cite{gstoreThesis}. This
-ensures that once a disk block is loaded from disk into the main memory
-buffers, several edges can be used. The locality of reference is a result of
-the traversal-based nature of most of the graph algorithms.
+access~\cite{localityLayout}, which is a direct result of the traversal-based
+nature of most of the graph algorithms. This is often taken advantage of by
+co-locating edges in close proximity within the same disk
+blocks~\cite{gstoreThesis}. This way, once a disk block is loaded into main
+memory buffers, several edges from it can be used for processing, reducing the
+disk I/O.
 
@@
 week of interactions. This means that edges that are temporally close are to
-be accessed together. Second, traversals are again key to many algorithms,
-such as connected components, clustering coefficient, PageRank, etc. This
-means that edges that are close by in terms of the path between their incident
-vertices as well as their timestamps are to be located together with the same
-blocks.
-
-In our earlier work~\cite{gedik14}, we introduced an interaction graph database
-that works on this principle of access locality. It proposed a disk
-organization that consists of as set of blocks, each containing a list of
-\emph{temporal neighbor lists}. A temporal neighborlist contains a head vertex
-and a set of incident edges within a given time range. The layout optimizer
-aims at bringing together temporal neighborlists that are (i) close in terms
-of their temporal ranges, (ii) have many edges between them, and (iii) have
-few edges going into temporal neighborlists outside the block.
+be accessed together. Second, traversals are again key to many graph
+algorithms, such as connected components, clustering coefficient, PageRank,
+etc. This means that edges that are close by in terms of the path between
+their incident vertices as well as their timestamps should be located together
+with the same blocks.  In our earlier work~\cite{gedik14}, we introduced an
+interaction graph database that works on this principle of access locality. It
+uses a disk organization that consists of as set of blocks, each containing a
+list of \emph{temporal neighbor lists}. A temporal neighbor list contains a
+head vertex and a set of incident edges within a time range. The layout
+optimizer aims at bringing together, into the same disk block, temporal
+neighbor lists that are ($i$) close in terms of their temporal ranges, ($ii$)
+have many edges between them, and ($iii$) have few edges going into temporal
+neighbor lists outside the block.
 
@@
 graphs, the attributes can be considered as properties associated with the
-edges. Attributes can be stored in two ways, either separately (e.g., in a
-relational table), or locally with the temporal neighbor lists.  If they are
-stored separately, then the graph database cannot take advantage of locality
-optimizations performed for block organization. The database must go back and
-forth between the disk blocks to access the edge attributes.  On the other
-hand, if attributes are stored locally in the disk blocks containing the graph
-structure, then there can be significant overhead due to disk I/O if only a
-few attributes are needed to answer a query.
+edges representing the interactions. Attributes can be stored in two ways,
+either separately (e.g., in a relational table), or locally with the temporal
+neighbor lists.  If they are stored separately, then the graph database cannot
+take full advantage of locality optimizations performed for block
+organization. The database must go back and forth between the disk blocks to
+access the edge attributes.  On the other hand, if attributes are stored
+locally in the disk blocks containing the graph structure, then there can be
+significant overhead due to disk I/O if only a few attributes are needed to
+answer a query.
 
@@
 impact on database performance~\cite{alagiannis14,grund10,stonebraker05}.  For
-table-based, relational databases, this fact has led database designers to
+table-based relational databases, this fact has led database designers to
 develop alternative approaches for storage layout: row-oriented
@@
 There a number of challenges in achieving an effective adaptive layout. First,
-we need to find the partitioning of attributes that optimizes the query I/O.
+we need to find the partitioning of attributes that minimized the query I/O.
 To address this, we model the problems of overlapping and non-overlapping
 attribute partitioning as mixed-integer linear programs (ILPs), and provide
-optimal solutions that minimize the I/O cost. Second, the query workload, and
-thus the attribute access pattern can change over time. For this purpose, our
-railway layout supports customization of the attribute partitioning of
-sub-blocks on a per-block basis. Third, such flexibility necessitates online
-configuration of attribute partitioning as the query workload evolves, which
-in turn requires fast algorithms for performing the attribute partitioning.
-For this purpose, we develop greedy heuristic algorithms for both overlapping
-and non-overlapping partitioning scenarios. These algorithms can scale to
-larger number of attributes, yet provide close to optimal query I/O
-performance. Finally, the railroad layout tradeoffs storage space to gain
-improved query I/O performance. The storage overhead is more pronounced for
-the case of overlapping partitioning. To address this, we limit the amount of
-storage  overhead that can be tolerated with an upper limit, and integrate
+optimal solutions that minimize the query I/O cost. Second, the query
+workload, and thus the attribute access pattern can change over time. For this
+purpose, our railway layout supports customization of the attribute
+partitioning of sub-blocks on a per-block basis. Third, such flexibility
+necessitates online configuration of attribute partitioning as the query
+workload evolves, which in turn requires fast algorithms for performing the
+attribute partitioning. For this purpose, we develop greedy heuristic
+algorithms for both overlapping and non-overlapping partitioning scenarios.
+These algorithms can scale to larger number of attributes, yet provide close
+to optimal query I/O performance. Finally, the railroad layout trades off
+storage space to gain improved query I/O performance. The storage overhead is
+more pronounced for the case of overlapping partitioning. To address this, we
+limit the amount of storage  overhead that can be tolerated, and integrate
 this limit to both our ILP formulations, as well as our greedy heuristics.

### 2014-09-11 docs/edbt2015/text/introduction.tex
railroad -> railway

@@
 These algorithms can scale to larger number of attributes, yet provide close
-to optimal query I/O performance. Finally, the railroad layout trades off
+to optimal query I/O performance. Finally, the railway layout trades off
 storage space to gain improved query I/O performance. The storage overhead is
@@
 \begin{itemize}
-\item We introduce the railroad layout for adaptive organization of interaction
+\item We introduce the railway layout for adaptive organization of interaction
 graphs on disk. 
@@
 workload. Our formulation also support upper bounding the amount of storage
-overhead introduced as a result of the railroad layout.
+overhead introduced as a result of the railway layout.
 \item To support online adaptation, we develop greedy heuristics that can scale

### 2014-09-16 docs/edbt2015/text/optimal.tex
renamed sections

\section{Optimal Railway Design}\label{sec:optimal}

\begin{alltt}\scriptsize
- Layout or partitioning problem
    - Cost model
\end{alltt}

### 2014-09-16 docs/edbt2015/text/system.tex
outline of the systems section
describe new figure

@@
     - How interaction graph are currently stored
+        - Most of the disk space work is on regular graphs, 
+           not interaction graphs. 
+        - Approach this as this is the railway, and we built
+           on this prior work.
+        - Builds on existing work that does not support 
+          adaptivity.
     - Introduce the rail layout
+    - Add the figure
+    - Query model
+   - Note that we need to talk here:
+    - For the system implementation, we need to add an additional
+       index to have a list of partitions for a block
+   - We also need to know how attributes are partitioned across sub-blocks 
+      for a given time interval.
+
+    - Current layout:
+    head vertex (id), num items in neighbor list, 
+    for each entry:
+        timestamp of edge
+        id of other edge of vertex
+        any edge properties
+
 \end{alltt}

### 2014-09-16 docs/edbt2015/text/system.tex
outline of the systems section

@@
     - Introduce the rail layout
-    - Add the figure
+    - Add the Figure~\ref{fig:}
     - Query model

### 2014-09-17 docs/edbt2015/text/optimal.tex
started the optimal railway design section

@@
 \section{Optimal Railway Design}\label{sec:optimal}  
+The optimal railway design concerns the partitioning of the disk blocks into
+sub-blocks such that the query I/O is minimized, while the storage overhead
+induced is kept below a specified threshold. This optimization is guided by
+the query workload, as observed by a particular disk block at hand.
 
-\begin{alltt}\scriptsize
-- Layout or partitioning problem
-    - Cost model
-\end{alltt}
+The partitioning of disk blocks into sub-blocks can be performed in two ways:
+\emph{non-overlapping} and \emph{overlapping}. In the non-overlapping case,
+each sub-block contains a non-overlapping subset of the edge attributes. In
+contrast, in the overlapping case, the attributes contained within sub-blocks
+can overlap. In both cases, the complete graph structure for the block is
+replicated in the sub-blocks, which is the main cause of increase in the query
+I/O.
+
+In both overlapping and non-overlapping partitioning, we trade increased
+storage  overhead for reduced space. In the overlapping case, the increase in
+the storage overhead is higher, as the space used for some of the attributes
+are replicated  as well, in addition to the graph structure. On the other
+hand, enabling overlapping attributes is expected to improve the query I/O.
+While the non-overlapping partitioning scenario is a special case of the more
+general case of overlapping partitioning, more specialized algorithms can be
+used to solve the non-overlapping partitioning problem.
+
+In the rest of this section, we first introduce some basic notation, and then
+formulate the overall optimization problem. The formulation of the query I/O
+cost and  storage overhead are presented next to complete the definition of
+the optimal railway design problem.
+
+HERE
+
+\subsection{Basic Notation}
+Let $Q$ be the query workload, where each query $q\in Q$ accesses a set of
+attributes $q.A$ and traverses parts of the graph for the time range
+$q.T=[q.t_s,q.t_e]$. We denote the set of all attributes as $A$. Given a block
+$B$, we denote its time range as $B.T$, which is the union of the time ranges
+of its temporal neighborlists. Let $s(a)$ denote the size of an attribute $a$.
+We use $c_n(B)$ to denote the number of temporal neighborlists within block
+$B$ and $c_e(B)$ to denote the total number of edges in the temporal
+neighborlists within the block. We overload the notation for block size and
+use $s(B)$ to denote the size of a block $B$. We have: 
+\begin{equation}
+s(B) = c_e(B) \cdot \Big(16 + \sum_{a\in B.A} s(a)\Big) + c_n(B) \cdot 12  
+\end{equation}
+Here, $16$ corresponds to the cost of storing the edge id and the timestamp,
+and $12$ corresponds to the cost of storing the head vertex ($8$ bytes) plus
+the number of entries ($4$ bytes) for a temporal neighborlist.
+
+\subsection{Optimization Problem}
+Our goal is to create a potentially overlapping partitioning of a block,
+denoted by $\mathcal{P}(B)$. We call the partitions \emph{sub-blocks}. We have
+$\bigcup_{B'\in \mathcal{P}(B)} B'.A = A$. We aim to find the function
+$\mathcal{P}$ that minimizes the query I/O over $B$, while keeping the
+relative storage overhead beyond a limit, say $1+\alpha$ times the original.
+If we represent the query I/O as $L(\mathcal{P}, B)$ and the relative storage
+overhead as $H(\mathcal{P}, B)$, our goal is to find:
+\begin{equation}
+\mathcal{P} \leftarrow \mbox{argmin}_{\{\mathcal{P}: H(\mathcal{P}(B)) < \alpha\}} L(\mathcal{P},B)
+\end{equation}
+
+\subsection{Storage Overhead Formulation}
+The storage overhead can be formalized as follows for the non-overlapping case:
+\begin{equation}
+H(\mathcal{P}, B) = (|\mathcal{P}(B)|-1)\cdot\Big(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\Big) 
+\end{equation}
+
+For the general case (including overlapping), we can formulate the storage
+overhead as follows:
+\begin{equation}
+H(\mathcal{P}, B) = \frac{\sum_{B'\in \mathcal{P}(B)} s(B')}{s(B)} - 1 
+\end{equation}
+
+\subsection{Query I/O Formulation}
+Let $m$ be a function that maps a query $q$ to the set of sub-blocks that are
+accessed to satisfy it for a relevant block $B$ under a given partitioning
+$\mathcal{P}$. For the case of non-overlapping attributes, we have:
+\begin{equation}
+m(\mathcal{P}, B, q) = \{B': B'\in \mathcal{P}(B) \wedge q.A \cap B'.A \ne \emptyset\}  
+\end{equation}
+
+For the case of overlapping attributes, we use a simple heuristic to define the
+set of sub-blocks to be used for answering the query.
+Algorithm~\ref{alg:greedyM} captures it. The idea is to select sub-blocks that
+bring the highest relative marginal gain, which is defined as the size of
+attribute data that contributes to the query result, relative to the sub-block
+size. While computing the marginal gain, attributes that are covered by
+sub-blocks that are already selected are not considered.
+%
+\begin{algorithm}[ht]
+\scriptsize
+\caption{m-overlapping($\mathcal{P}, B, q$)}
+\label{alg:greedyM}
+\KwData{$\mathcal{P}$: partitioning function, $B$: block, $q$: query}
+$S\leftarrow \emptyset; R\leftarrow \emptyset$ \tcp*{Selected attributes; Resulting sub-blocks}
+\While(\tcp*[f]{While unselected attributes remain}){$S \subset q.A$}{
+  $B' \leftarrow \mbox{argmax}_{B'\in\mathcal{P}(B)\setminus R} \sum_{a\in B'.A \cap q.A \setminus S} \frac{c_e(B') \cdot s(a)}{s(B')}$
+  $S \leftarrow S \cup B'.A$\tcp*{Extend the selected attributes}
+  $R\leftarrow R \cup B'$\tcp*{Extend the selected sub-blocks}
+}
+\Return R \tcp*{Final set of sub-blocks covering the query attributes}
+\end{algorithm} 
+%
+The query I/O cost for a block is then given by:
+\begin{equation}
+L(\mathcal{P}, B) = \sum_{q\in Q} w(q)\cdot\mathbf{1}(q.T \cap B.T \neq \emptyset) \cdot \sum_{B'\in m(\mathcal{P}, B, q)} \!\!s(B')
+\end{equation}
+
+Here, $w(q)$ represents the query weight.

### 2014-09-17 docs/edbt2015/text/optimal.tex
started the optimal railway design section

@@
 sub-blocks such that the query I/O is minimized, while the storage overhead
-induced is kept below a specified threshold. This optimization is guided by
-the query workload, as observed by a particular disk block at hand.
+induced is kept below a desired threshold. This optimization is guided by the
+query workload, as observed by a disk block within a particular time range.
 
-The partitioning of disk blocks into sub-blocks can be performed in two ways:
-\emph{non-overlapping} and \emph{overlapping}. In the non-overlapping case,
-each sub-block contains a non-overlapping subset of the edge attributes. In
-contrast, in the overlapping case, the attributes contained within sub-blocks
-can overlap. In both cases, the complete graph structure for the block is
-replicated in the sub-blocks, which is the main cause of increase in the query
-I/O.
+The partitioning of disk blocks into sub-blocks can be \emph{non-overlapping}
+or \emph{overlapping}. In the non-overlapping case, each sub-block contains a
+non-overlapping subset of the edge attributes. In the overlapping case, the
+subset of attributes contained within sub-blocks can overlap. In both cases,
+the complete graph structure for the block is replicated in the sub-blocks,
+which is the main cause of the increased storage overhead.
 
 In both overlapping and non-overlapping partitioning, we trade increased
-storage  overhead for reduced space. In the overlapping case, the increase in
-the storage overhead is higher, as the space used for some of the attributes
-are replicated  as well, in addition to the graph structure. On the other
-hand, enabling overlapping attributes is expected to improve the query I/O.
-While the non-overlapping partitioning scenario is a special case of the more
-general case of overlapping partitioning, more specialized algorithms can be
-used to solve the non-overlapping partitioning problem.
+storage overhead for reduced query I/O cost. In the overlapping case, the
+increase in the storage overhead is higher, as some of the attributes are
+replicated, in addition to the graph structure. On the other hand, enabling
+overlapping attributes is expected to reduce the query I/O (in the extreme
+case, there could be one sub-block per query). While the non-overlapping
+partitioning scenario is a special case of the  overlapping one, specialized
+algorithms can be used to solve the former problem.
 
-In the rest of this section, we first introduce some basic notation, and then
-formulate the overall optimization problem. The formulation of the query I/O
-cost and  storage overhead are presented next to complete the definition of
-the optimal railway design problem.
-
-HERE
+In the rest of this section, we first introduce basic notation and then
+formulate the overall optimization problem. The modeling of the query I/O and
+storage overhead are presented next, which complete the formalization of the
+optimal railway design problem.
 

### 2014-09-17 docs/edbt2015/text/introduction.tex
fixed reference
outline of the systems section
started the optimal railway design section

@@
 \section{Introduction}\label{sec:introduction}
-
+\noindent
 We are living in an ever more connected world, where the data generated by
@@
 co-locating edges in close proximity within the same disk
-blocks~\cite{gstoreThesis}. This way, once a disk block is loaded into main
+blocks~\cite{g-store}. This way, once a disk block is loaded into main
 memory buffers, several edges from it can be used for processing, reducing the
@@
 
+% TODO: Feels a little out of place, rephrase to make attribute issue more pronounced
 To query an interaction graph, most algorithms traverse the graph structure to

### 2014-09-17 docs/edbt2015/text/optimal.tex
started the optimal railway design section

@@
 \section{Optimal Railway Design}\label{sec:optimal}  
-The optimal railway design concerns the partitioning of the disk blocks into
+\noindent
+The optimal railway design concerns the partitioning of disk blocks into
 sub-blocks such that the query I/O is minimized, while the storage overhead
 induced is kept below a desired threshold. This optimization is guided by the
-query workload, as observed by a disk block within a particular time range.
+query workload observed by a disk block within a particular time range.
 
 The partitioning of disk blocks into sub-blocks can be \emph{non-overlapping}
-or \emph{overlapping}. In the non-overlapping case, each sub-block contains a
-non-overlapping subset of the edge attributes. In the overlapping case, the
+or \emph{overlapping}. In the non-overlapping case, the attributes are
+partitioned among the sub-blocks with no overlap. In the overlapping case, the
 subset of attributes contained within sub-blocks can overlap. In both cases,
-the complete graph structure for the block is replicated in the sub-blocks,
-which is the main cause of the increased storage overhead.
+the complete graph structure for the block is replicated within the
+sub-blocks, which results in a storage overhead.
 
@@
 \subsection{Basic Notation}
+\noindent
 Let $Q$ be the query workload, where each query $q\in Q$ accesses a set of
@@
 
+Our goal is to create a potentially overlapping partitioning of attributes for
+block $B$, resulting in a set of sub-blocks denoted by $\mathcal{P}(B)$. In
+other words, we have $\bigcup_{B'\in \mathcal{P}(B)} B'.A = A$. Here,
+$\mathcal{P}$ is the partitioning function.
+
 \subsection{Optimization Problem}
-Our goal is to create a potentially overlapping partitioning of a block,
-denoted by $\mathcal{P}(B)$. We call the partitions \emph{sub-blocks}. We have
-$\bigcup_{B'\in \mathcal{P}(B)} B'.A = A$. We aim to find the function
-$\mathcal{P}$ that minimizes the query I/O over $B$, while keeping the
-relative storage overhead beyond a limit, say $1+\alpha$ times the original.
-If we represent the query I/O as $L(\mathcal{P}, B)$ and the relative storage
-overhead as $H(\mathcal{P}, B)$, our goal is to find:
+\noindent
+We aim to find the partitioning function $\mathcal{P}$ that minimizes the query
+I/O over $B$, while keeping the storage overhead below a limit, say
+$1+\alpha$ times the original. The original corresponds to the case of a single
+block that contains all the attributes. Let us denote the query I/O as
+$L(\mathcal{P}, B)$ and the storage overhead as $H(\mathcal{P}, B)$,
+our goal is to find: 
 \begin{equation} 
@@
 \subsection{Storage Overhead Formulation}
-The storage overhead can be formalized as follows for the non-overlapping case:
+\noindent
+The storage overhead is defined as the additional amount of disk space used to
+store the sub-blocks, normalized by the original space needed by a single
+block (no partitioning). The storage overhead can be formalized as follows,
+for the non-overlapping case:
 \begin{equation}
 H(\mathcal{P}, B) = (|\mathcal{P}(B)|-1)\cdot\Big(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\Big)
+\label{eq:nov}
 \end{equation}
 
-For the general case (including overlapping), we can formulate the storage
-overhead as follows:
+Basically, for the non-overlapping case, there is no overhead due to the
+attributes, as they are not repeated. However, there is overhead for the block
+structure that is repeated for each sub-block. There are $|\mathcal{P}(B)|-1$
+such extra sub-blocks, and for each, the contribution to the overhead due to
+storing the graph structure is given by $s(B)-c_e(B)\cdot \sum_{a\in A} s(a)$.
+Equation~\ref{eq:nov} has one nice feature, that is, it does not depend on the
+details of the attribute partitioning, other than the number of partitions. We
+make use of this feature, later for the ILP formulation of the problem.
+
+For the general case of a potentially overlapping partitioning, we can
+formulate the storage overhead as follows:
 \begin{equation}
@@
 
+This formulation follows directly from the definition of storage overhead.
+While simple, it depends on the details of the partitioning, as $s(B')$ is the
+size of a sub-block $B'$, which depends on the list of attributes within the
+sub-block.
+
 \subsection{Query I/O Formulation}
+\noindent
 Let $m$ be a function that maps a query $q$ to the set of sub-blocks that are
 accessed to satisfy it for a relevant block $B$ under a given partitioning
-$\mathcal{P}$. For the case of non-overlapping attributes, we have:
+$\mathcal{P}$.
+
+For the case of non-overlapping attributes, the $m$ function lists all the 
+sub-blocks whose attributes intersect with those from the query. Formally:
 \begin{equation}
@@
 For the case of overlapping attributes, we use a simple heuristic to define the
-set of sub-blocks to be used for answering the query.
-Algorithm~\ref{alg:greedyM} captures it. The idea is to select sub-blocks that
-bring the highest relative marginal gain, which is defined as the size of
-attribute data that contributes to the query result, relative to the sub-block
-size. While computing the marginal gain, attributes that are covered by
-sub-blocks that are already selected are not considered.
+set of sub-blocks used for answering the query. Algorithm~\ref{alg:greedyM}
+captures it. The basic idea is to start with an empty list of sub-blocks and
+greedily add new sub-blocks to it, until all query attributes are covered.  At
+each iteration, the sub-block that brings the highest relative marginal gain
+is  picked. The relative marginal gain is defined as  the total size of the
+attributes from the sub-block that contribute to the query result, relative to
+the sub-block size. While computing the marginal gain, attributes that are
+already covered by sub-blocks that are selected earlier are not considered.
+
 %
@@
 \end{algorithm}
-%
+
+HERE
 The query I/O cost for a block is then given by:

### 2014-09-18 docs/edbt2015/text/system.tex
started example interaction graph figure
working on system section

--- 2014-09-16 docs/edbt2015/text/system.tex
@@
 \section{System Overview}\label{sec:system}
+
+\begin{figure}[t]
+\includegraphics[width=0.9\columnwidth]{figures/example_interaction.pdf} 
+ \caption{An example interaction graph for call data records, capturing the
+   telephone call interactions among four people. Each edge in the graph is
+   associated with attributes for the interaction, including the time the call
+   was placed, the duration of the call, the cell phone tower, and the IMEI
+   number identifying the device used.}
+ \label{fig:example}
+ \end{figure}
+
+The design of the railway disk layout builds on our prior work~\cite{gedik14}, 
+which organized the disk layout for an interaction graph databases to improve
+access locality. The railway layout extends the previous design, by enabling the
+system to adapt the layout for changing workloads.
+
+\paragraph*{Motivating Example}
+%
+To explain the design of the railway layout, we first introduce a small,
+motivating example, in Figure~\ref{fig:example}. This figure illustrates an
+interaction graph for call data records, capturing the telephone call
+interactions among four people. Each edge in the graph is associated with
+attributes for the interaction, including the time the call was placed, the
+duration of the call, the cell phone tower, and the IMEI number identifying the
+device used to place the call. In this example, there were four calls
+placed. One of them is a call from Alice to Bob, starting at 13:46. They spoke
+for 600 seconds. The call was received by cell phone tower 1, and the Alice's
+phone had an IMEI number of 100.
+
+A telecommunications company typically performs various analytics on this
+data. For example, for billing purposes, they might want to capture the
+duration of all calls. To plan for infrastructure provisioning, they might want
+to record a count of the number of calls that each cell phone tower handled.
+
+\begin{figure*}[ht]
+\includegraphics[width=0.9\textwidth]{figures/before_after.pdf} 
+ \caption{The standard disk block strorage for an interaction graph, and a
+   partitioning into sub-blocks for the railway layout. Each sub-block maintains
+ its own copy of the neighbor list, but a subset of the attributes.}
+ \label{fig:before_after}
+ \end{figure*}
+
+\paragraph*{Storage for Locality}
+%
+The lefthand side of Figure~\ref{fig:before_after} shows the typical storage for interaction graphs.
+Each disk block contains an adjacency list representation of the interaction
+graph. The block contains a sequence of vertices, identified by a
+\emph{head-node id}, followed by a \emph{count} of the number of neighbors, and
+then the neighbor list itself. Each entry in the neighbor list is composed of a
+\emph{timestamp}, an \emph{id} for the destination vertex, and the properties
+for that edge.
+
+\paragraph*{Rail Layout}
+%
+
 \begin{alltt}\scriptsize
@@
     - Introduce the rail layout
-    - Add the figure
+    - Add the Figure~\ref{fig:rail_layout}
     - Query model
@@
 \end{alltt}
+
+\begin{figure}[ht]
+\includegraphics[width=0.9\columnwidth]{figures/rail_layout.pdf} 
+ \caption{todo}
+ \label{fig:rail_layout}
+ \end{figure}

### 2014-09-18 docs/edbt2015/text/optimal.tex
first cut of the optimal railway design section is complete

@@
 
-HERE
-The query I/O cost for a block is then given by:
+Given that we have defined the function $m$ that maps a query to the set of 
+sub-blocks used to answer it, we can now formalize the total query I/O cost
+for a block under a given workload:
 \begin{equation}
@@
 
-Here, $w(q)$ represents the query weight.
+We simply sum the I/O cost contributions of the queries to compute the total
+I/O cost. A query contributes to the total I/O cost if and only if its time
+range intersects with that of the block ($\mathbf{1}(q.T \cap B.T \neq
+\emptyset)$). If it does, then we add the sizes of all the sub-blocks used to
+answer the query to the total I/O cost. Furthermore, we multiply the I/O cost
+contribution of a query with its frequency, denoted by $w(q)$ in the formula.

### 2014-09-19 docs/edbt2015/text/ilp.tex
ILP formulation is started

--- 2014-09-01 docs/vldbj/text/solution.tex
@@
-\clearpage
-\newpage
-\section{Partitioning}
-This section is still a work-in-progress. We are trying to find the algorithms.
+\section{ILP Solution}\label{sec:ilp}
+\noindent
+In this section, we formulate the optimal railway design problem as a mixed
+Integer Linear Program (ILP). The main challenge is to represent the objective
+function and the constraint as a linear combination of potentially integer
+variables. We separate the cases of non-overlapping and overlapping
+partitioning, as the former case can be formulated using smaller number of 
+constraints.
 
-\subsection{Non-Overlapping Attributes}
-
-\paragraph*{Problem.$\,$} \emph{Find a true partitioning of attributes that
-minimizes the query I/O and bounds the storage cost by some upper limit.}
-
-\subsubsection{Integer Linear Program Formulation}\label{subsubsec:nov-ilp}
-We present an ILP formuation of the problem. For this purpose, we define a 
-number of binary ($0$ or $1$) variables: 
+For the ILP formulation, we define a number of binary ($0$ or $1$) variables: 
 \begin{itemize}
@@
 \begin{itemize}
-\item $x$s define the attribute-to-partition assignment.
-\item $y$s help formulate the query I/O contribution
-of partitions, excluding their assigned attributes.
-\item $z$s help formulate the query I/O contribution
-of partitions, only considering their assigned attributes.
+\item $x$s define the attribute-to-partition assignments.
+\item $y$s help formulate the query I/O contribution of each partition due to
+the graph structure they contain (excluding their assigned attributes).
+\item $z$s help formulate the query I/O contribution of each partition, only
+considering the attributes they are assigned.
 \item $u$s help formulate the storage overhead requirement.
 \end{itemize}
+
+Here, we assume that the maximum number of partitions is fixed. In fact, we
+cannot have more partitions than attributes, so the number of partitions is
+upper bounded by $|A|$, and thus $0\leq p<|A|$. However, some of these
+partitions can be empty in the optimal solution, which means that the number
+of partitions found by the ILP solution is typically lower than the maximum
+possible. A simple post-processing step removes empty partitions and creates
+the final partitioning to be used.
 
@@
 than there are attributes.
+
+\subsection{Non-Overlapping Attributes}\label{subsubsec:nov-ilp}
 
@@
 \begin{figure}[!t]
-\begin{mdframed}
+%\begin{mdframed}
 \begin{eqnarray}
@@
 \end{eqnarray}
-\end{mdframed}
+%\end{mdframed}
 \caption{ILP formulation for the non-overlapping partitioning}
@@
 
-\subsubsection{Heuristic Solution}
-
-\paragraph*{Approach.$\,$}
-TODO
-
-\begin{algorithm}[ht]
-\scriptsize
-\caption{Algorithm for partitioning blocks into sub-blocks with non-overlapping attributes.}
-\label{alg:non-overlappingP}
-\KwData{$B$: block, $Q$: set of queries}
-$c^*\leftarrow \infty$ \tcp*{Lowest cost over all \# of partitions}
-\For(\tcp*[f]{For each possible \# of partitions}){$k=1$ to $|A|$}{
-   $R[i]\leftarrow \emptyset, \forall i\in [1..k]$ \tcp*{Initialize partitions}
-   \For(\tcp*[f]{For each attribute}){$a \in A$\textnormal{, in decr.\/ order of }$f(a)$}{
-      $c\leftarrow \infty$ \tcp*{Lowest cost over all assignments}  
-      $j\leftarrow -1$ \tcp*{Best partition assignment}
-      \For(\tcp*[f]{For each partition assignment}){$i\in [1..k]$} {
-         $R[i]\leftarrow R[i] \cup \{a\}$\tcp*{Assign attribute}
-         \If(\tcp*[f]{If query cost is lower}){$L(R, B, Q)<c$}{
-            $c\leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
-            $j\leftarrow i$\tcp*{Update the best partition}
-         }
-         $R[i]\leftarrow R[i] \setminus \{a\}$\tcp*{Un-assign attribute}
-      }
-      $R[j]\leftarrow R[j] \cup \{a\}$\tcp*{Assign to best partition}
-   }
-   \lIf(\tcp*[f]{If solution infeasible}){$H(R, B, Q)>\alpha$}{\textbf{break}}
-   \If(\tcp*[f]{If solution has lower cost}){$L(R, B, Q)<c^*$}{
-     $c^* \leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
-     $\mathcal{P}(B)\leftarrow R$\tcp*{Update the best partitioning}
-   } 
-}
-\Return $\mathcal{P}(B)$ \tcp*{Final set of sub-blocks}
-\end{algorithm}
-
-\clearpage
-\newpage
-\subsection{Overlapping Attributes}
+\subsection{Overlapping Attributes}\label{subsubsec:ov-ilp}
 
@@
 
-\subsubsection{Integer Linear Program Formulation}
 We present an ILP formuation of the problem, as we did for the case of
@@
 \sum_{p=1}^{k} \Big((16\cdot c_e(B) &+& 12 \cdot c_n(B)) \cdot u_p  \nonumber \\ 
-+ \sum_{a\in A} s(a) \!\!&\cdot&\!\! c_n(B)\cdot x_{a,p} \Big) \leq s(B)\cdot (1+\alpha)
++ \sum_{a\in A} s(a) \!\!&\cdot&\!\! c_e(B)\cdot x_{a,p} \Big) \leq s(B)\cdot (1+\alpha)
 \end{eqnarray}
@@
 \sum_{p=1}^{k} \Big((16\cdot c_e(B) &+& 12 \cdot c_n(B)) \cdot u_p  \nonumber \\ 
-+ \sum_{a\in A} s(a) \!\!&\cdot&\!\! c_n(B)\cdot x_{a,p}  \Big)\leq s(B)\cdot (1+\alpha)\nonumber
++ \sum_{a\in A} s(a) \!\!&\cdot&\!\! c_e(B)\cdot x_{a,p}  \Big)\leq s(B)\cdot (1+\alpha)\nonumber
 \end{eqnarray}
-\end{mdframed}
+%\end{mdframed}
 \caption{ILP formulation for the overlapping partitioning}
@@
 Figure~\ref{fig:ov-ilp}.
-
-\subsubsection{Heuristic Solution}
-
-\paragraph*{Approach.$\,$} We start the algorithm with a   partitioning based
-on what queries we have seen. Every query gets its own sub-block. This is the
-``ideal'' partitioning, because the I/O cost would be minimized for every
-query that we would have seen. The algorithm iteratively combines the two
-partitions that are closest together.  After each combination of partitions,
-the algorithm calculate the storage overhead for the partitioning. The
-algorithm stops when the  storage cost is below some specified threshold.  The
-result is the block partitioning.
-
-\begin{algorithm}[ht]
-\scriptsize
-\caption{Algorithm for partitioning blocks into sub-blocks with overlapping attributes.}
-\label{alg:overlappingP}
-\KwData{$B$: block, $Q$: set of queries}
-$\mathcal{P}(B) \leftarrow \{q.A: q\in Q\}$ \tcp*{Each query gets its own sub-block}
-\While(\tcp*[f]{Until storage overhead is below $\alpha$}){$H(\mathcal{P},B) > \alpha$}{
-  $c^{*}\gets \infty $ \tcp*{Lowest cost over all sub-block pairs}
-  $(b_x,b_y)\gets (\emptyset,\emptyset)$ \tcp*{Sub-block pair with the lowest cost}
-  \For(\tcp*[f]{For each pair of blocks}){$\{b_i,b_j\}\in\mathcal{P}(B)$}{
-    $\mathcal{P'}(B) \leftarrow \mathcal{P}(B) \setminus \{b_i, b_j\} \cup \{b_i \cup b_j\}$\\
-    $c\gets \frac{L(\mathcal{P}',B,Q)-L(\mathcal{P},B,Q)}{H(\mathcal{P},B)-H(\mathcal{P}',B)}$\tcp*{Cost of merge}
-    \If(\tcp*[f]{Cost is lower}){$c<c^{*}$}{
-        $c^{*}\gets c$\tcp*{Update the lowest cost}
-        $(b_x,b_y)\gets (b_i,b_j)$\tcp*{Update the best pair}
-    }
-  }
-  $\mathcal{P}(B) \leftarrow \mathcal{P}(B) \setminus \{b_x, b_y\} \cup \{b_x \cup b_y\}$ \\
-}
-\Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
-\end{algorithm}

### 2014-09-19 docs/edbt2015/text/optimal.tex
the first cut of the ILP formulation is completed

@@
 H(\mathcal{P}, B) = (|\mathcal{P}(B)|-1)\cdot\Big(1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}\Big)
-\label{eq:nov}
+\label{eq:nov-ohead}
 \end{equation}
@@
 storing the graph structure is given by $s(B)-c_e(B)\cdot \sum_{a\in A} s(a)$.
-Equation~\ref{eq:nov} has one nice feature, that is, it does not depend on the
+Eq.~\ref{eq:nov-ohead} has one nice feature, that is, it does not depend on the
 details of the attribute partitioning, other than the number of partitions. We

### 2014-09-19 docs/edbt2015/text/ilp.tex
the first cut of the ILP formulation is completed

@@
 function and the constraint as a linear combination of potentially integer
-variables. We separate the cases of non-overlapping and overlapping
-partitioning, as the former case can be formulated using smaller number of 
-constraints.
+variables.
 
@@
 
-Here, we assume that the maximum number of partitions is fixed. In fact, we
-cannot have more partitions than attributes, so the number of partitions is
-upper bounded by $|A|$, and thus $0\leq p<|A|$. However, some of these
-partitions can be empty in the optimal solution, which means that the number
-of partitions found by the ILP solution is typically lower than the maximum
-possible. A simple post-processing step removes empty partitions and creates
-the final partitioning to be used.
+In total, we have $|A|\cdot(|A|+1)\cdot(|Q|+1)$ variables. Here, we assume that
+the maximum number of partitions is fixed. In fact, we cannot have more
+partitions than attributes, so the number of partitions is upper bounded by
+$k=|A|$, and thus $0\leq p<k$. However, some of these partitions can be empty
+in the optimal solution, which means that the number of partitions found by
+the ILP solution is typically lower than the maximum possible. A simple
+post-processing step removes empty partitions and creates the final
+partitioning to be used.
 
-We also define a helper notation for representing whether a variable is
-accessed by a query or not: $q(a)\equiv \mathbf{1}(a \in q.A)$, where
-$\mathbf{1}$ is the indicator function. We assume that there are a maximum of
-$k$ partitions. $k$ can be taken as $|A|$, as there cannot be more partitions
-than there are attributes.
+Finally, we define a helper notation for representing whether a variable is
+accessed by a query or not: $q(a)\equiv \mathbf{1}(a \in q.A)$.
 
-\subsection{Non-Overlapping Attributes}\label{subsubsec:nov-ilp}
+We are now ready to state the ILP formulation. We separate the cases of
+non-overlapping and overlapping partitioning, as the former case can be
+formulated using smaller number of constraints.
 
-We are now ready to state the ILP formulation. We start with the objective
-function, that is the total query I/O:
+\subsection{Non-Overlapping Partitions}\label{subsubsec:nov-ilp}
+\noindent
+We start with the objective function, that is the total query I/O, which is to
+be minimized:
 \begin{eqnarray}
@@
 that $z_{a,p,q}$ could have been replaced with $x_{a,p}\cdot y_{p,q}$, but
-that would make the objective function non-linear.
+that would have made the objective function non-linear.
 
@@
 follows, where $K$ is a large constant guaranteed to be larger than $\beta_1$
-for practical purposes:
+for all practical purposes:
 \begin{eqnarray}
@@
 variable is simply forced to be $1$. Otherwise, the $z$ variable can be either
-$0$ or $1$, but since the $z$ variables appear in the objective fucntion as
-positive terms, the solver will set them to $0$, which is what we want.
+$0$ or $1$, but since the $z$ variables appear in the objective function as
+positive terms, the solver will set them to $0$ to minimize the I/O cost (Note
+that the $z$ variables do not appear in any other constraint).
 
@@
 Our fifth, and the last, constraint deals with the storage overhead. We want to
- make sure that the storage overhead does not go over $\alpha$. The storage
-overhead depends on the number of partitions used. That means that the only 
-ILP variables it depends on is the $u$s. In particular, the number of
+ make sure that the storage overhead does not go over $\alpha$. Recall that
+for  the non-overlapping attributes case, the storage overhead depends on the
+number of partitions used (Eq.~\ref{eq:nov-ohead}). That means that the only 
+ILP variables it depends on are the $u$s. In particular, the number of
 partitions used is given by $\sum_{p=1}^{k} u_p$. This results in the
@@
 %\end{mdframed}
-\caption{ILP formulation for the non-overlapping partitioning}
+\caption{ILP formulation for the non-overlapping optimal railway design.}
 \label{fig:nov-ilp}
@@
 The final ILP formulation for the non-overlapping partitioning is given in
-Figure~\ref{fig:nov-ilp}.
+Figure~\ref{fig:nov-ilp}. We have a total of $|A|^2\cdot|Q| +
+2\cdot|A|\cdot|Q| + 3\cdot|A| + 1$ constraints and the objective function
+contains $|A|\cdot|Q|\cdot(1+|A|)$ variables.
 
-\subsection{Overlapping Attributes}\label{subsubsec:ov-ilp}
-
-\paragraph*{Problem.$\,$} \emph{Find an overlapping partitioning of attributes
-that minimizes the query I/O and bounds the storage cost by some upper
-limit.}
-
-We present an ILP formuation of the problem, as we did for the case of
+\subsection{Overlapping Partitions}\label{subsubsec:ov-ilp}
+\noindent
+We present an ILP formulation of the problem, as we did for the case of
 non-overlapping partitions in Section~\ref{subsubsec:nov-ilp}. We use the same
@@
 
-Our first constraint is that, each attribute must be assigned to at leat one
+Our first constraint is that, each attribute must be assigned to at least one
 partition. Formally:
@@
 
-As our second constraint, we require that for each attribute used by a query,
-there needs to be a partition that is used by that query and that contains the
-attribute in question. Formally:
+As our second constraint, we require that for each attribute contained in a
+query, there needs to be a partition that is used by that query and that
+contains the attribute in question. Formally:
 \begin{eqnarray}
@@
 $\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, (z_{a,p,q} = 1) \implies 
-(x_{a,p} = 1)$. This can be state as linear constraints:
+(x_{a,p} = 1)$. This can be stated as linear constraints:
 \begin{eqnarray}
@@
 Our fifth constraint is that, if an attribute $a$ is assigned to a partition
-$p$, and partition $p$ is used by a query $q$, then the corresponding $z$
-variable must be set to $1$. This is same as the formulation for the
-non-overlapping case from Eq.~\ref{eq:no-z}.
+$p$, and partition $p$ is used by a query $q$, then the corresponding
+$z_{a,p,q}$ variable must be set to $1$. This is same as the formulation for
+the non-overlapping case from Eq.~\ref{eq:no-z}.
 
@@
 the storage overhead formulation for the overlapping case is different from 
-the one for the non-overlapping. This is because the overhead does not merely
-depend on the number of partitions, as attributes might have to be loaded
-multiple times from different partitions. As a result, we express the overhead
-using base variables as in the objective function. Formally:
+the one for the non-overlapping case. This is because the overhead does not
+merely depend on the number of partitions, as attributes might have to be read
+multiple times from different partitions (due to the overlaps). As a result,
+we express the overhead using base variables as in the objective function.
+Formally:
 \begin{eqnarray}
@@
 The final ILP formulation for the overlapping partitioning is given in
-Figure~\ref{fig:ov-ilp}.
+Figure~\ref{fig:ov-ilp}. We have a total of $2\cdot|A|^2\cdot|Q| +
+3\cdot|A|\cdot|Q| + 3\cdot|A| + 1$ constraints and the objective function
+contains $|A|\cdot|Q|\cdot(1+|A|)$ variables.

### 2014-09-20 docs/edbt2015/text/heuristic.tex
started on heuristic solution

@@
 \section{Heuristic Solution}\label{sec:heuristic}
+
+\begin{alltt}
+- Talk about the scalability of the optimal solution
+- Do we have a running time?
+- Introduce a heuristic approximation approach
+\end{alltt}
+
+\subsection{Non-Overlapping Partitions}\label{subsec:nov-heuristic}
+\noindent
+
+\paragraph*{Approach$\,$}
+TODO
+
+\begin{algorithm}[ht]
+\scriptsize
+\caption{Algorithm for partitioning blocks into sub-blocks with non-overlapping attributes.}
+\label{alg:non-overlappingP}
+\KwData{$B$: block, $Q$: set of queries}
+$c^*\leftarrow \infty$ \tcp*{Lowest cost over all \# of partitions}
+\For(\tcp*[f]{For each possible \# of partitions}){$k=1$ to $|A|$}{
+   $R[i]\leftarrow \emptyset, \forall i\in [1..k]$ \tcp*{Initialize partitions}
+   \For(\tcp*[f]{For each attribute}){$a \in A$\textnormal{, in decr.\/ order of }$f(a)$}{
+      $c\leftarrow \infty$ \tcp*{Lowest cost over all assignments}  
+      $j\leftarrow -1$ \tcp*{Best partition assignment}
+      \For(\tcp*[f]{For each partition assignment}){$i\in [1..k]$} {
+         $R[i]\leftarrow R[i] \cup \{a\}$\tcp*{Assign attribute}
+         \If(\tcp*[f]{If query cost is lower}){$L(R, B, Q)<c$}{
+            $c\leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
+            $j\leftarrow i$\tcp*{Update the best partition}
+         }
+         $R[i]\leftarrow R[i] \setminus \{a\}$\tcp*{Un-assign attribute}
+      }
+      $R[j]\leftarrow R[j] \cup \{a\}$\tcp*{Assign to best partition}
+   }
+   \lIf(\tcp*[f]{If solution infeasible}){$H(R, B, Q)>\alpha$}{\textbf{break}}
+   \If(\tcp*[f]{If solution has lower cost}){$L(R, B, Q)<c^*$}{
+     $c^* \leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
+     $\mathcal{P}(B)\leftarrow R$\tcp*{Update the best partitioning}
+   } 
+}
+\Return $\mathcal{P}(B)$ \tcp*{Final set of sub-blocks}
+\end{algorithm}
+
+\subsection{Overlapping Partitions}\label{subsec:ov-heuristic}
+\noindent
+
+\paragraph*{Approach$\,$} We start the algorithm with a   partitioning based
+on what queries we have seen. Every query gets its own sub-block. This is the
+``ideal'' partitioning, because the I/O cost would be minimized for every
+query that we would have seen. The algorithm iteratively combines the two
+partitions that are closest together.  After each combination of partitions,
+the algorithm calculate the storage overhead for the partitioning. The
+algorithm stops when the  storage cost is below some specified threshold.  The
+result is the block partitioning.
+
+\begin{algorithm}[ht]
+\scriptsize
+\caption{Algorithm for partitioning blocks into sub-blocks with overlapping attributes.}
+\label{alg:overlappingP}
+\KwData{$B$: block, $Q$: set of queries}
+$\mathcal{P}(B) \leftarrow \{q.A: q\in Q\}$ \tcp*{Each query gets its own sub-block}
+\While(\tcp*[f]{Until storage overhead is below $\alpha$}){$H(\mathcal{P},B) > \alpha$}{
+  $c^{*}\gets \infty $ \tcp*{Lowest cost over all sub-block pairs}
+  $(b_x,b_y)\gets (\emptyset,\emptyset)$ \tcp*{Sub-block pair with the lowest cost}
+  \For(\tcp*[f]{For each pair of blocks}){$\{b_i,b_j\}\in\mathcal{P}(B)$}{
+    $\mathcal{P'}(B) \leftarrow \mathcal{P}(B) \setminus \{b_i, b_j\} \cup \{b_i \cup b_j\}$\\
+    $c\gets \frac{L(\mathcal{P}',B,Q)-L(\mathcal{P},B,Q)}{H(\mathcal{P},B)-H(\mathcal{P}',B)}$\tcp*{Cost of merge}
+    \If(\tcp*[f]{Cost is lower}){$c<c^{*}$}{
+        $c^{*}\gets c$\tcp*{Update the lowest cost}
+        $(b_x,b_y)\gets (b_i,b_j)$\tcp*{Update the best pair}
+    }
+  }
+  $\mathcal{P}(B) \leftarrow \mathcal{P}(B) \setminus \{b_x, b_y\} \cup \{b_x \cup b_y\}$ \\
+}
+\Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
+\end{algorithm}

### 2014-09-20 docs/edbt2015/text/abstract.tex
first attempt at an abstract

We are living in an ever more connected world, where data recording the
interactions between people, software systems, and the physical world is
becoming increasingly prevalent. This data often takes the form of a temporally
evolving graph, where entities are the vertices and the interactions between
them are the edges. We call such graphs interaction graphs. Various application
domains, including telecommunciations and social media, depend on analytics
performed on interaction graphs. The ability to efficiently support historical
analysis over interaction graphs require effective solutions for the problem of
data layout on disk.

This paper presents an adaptive disk layout called the railway layout for
optimizing disk block storage for interaction graphs. The key idea is to divide
blocks into one or more sub-blocks, where each sub- block contains a subset of
the attributes, but the entire graph structure is replicated within each
sub-block.  We introduce optimal ILP formulations for partitioning disk blocks
into sub-blocks with overlapping and non-overlapping attributes. Additionally,
wepresent greedy heuristic approaches that can scale better compared to the ILP
alternatives, yet achieve close to optimal query I/O. To demonstrate the
benefits of the railway layout, ee provide an extensive experimental study
comparing our approach to a few baseline alternatives.

### 2014-09-21 docs/edbt2015/text/related.tex
Added some related work text. did some bibtex cleanup

@@
 
+The railway layout and algorithms build on our prior work~\cite{gedik14}, which
+organized the disk layout for an interaction graph databases to improve access
+locality. The railway layout extends the previous design, by enabling the system
+to adapt the layout for changing workloads.
+
+Our adaptation scheme is similar to work on adaptive layouts for relational
+database. The $H_2O$~\cite{alagiannis14} system can adapt its data layout into
+three types, row-major, column-major, or groups of columns, depending on the
+workload. HYRISE~\cite{grund10} provides a similar adaptive layout scheme for an
+in-memory relational database. Both systems use heuristic, iterative solutions
+to determine partitioning. The railway layout scheme differs, in that it
+targets interaction graphs, and we present optimal solutions, in addition to 
+heuristic solutions.
+
  \begin{alltt}\scriptsize
-Gedik et al.~\cite{gedik14}
-H2O \cite{alagiannis14}
-HYRISE~\cite{grund10}
+% Gedik et al.~\cite{gedik14}
+% H2O \cite{alagiannis14}
+% HYRISE~\cite{grund10}
  Bornea et al.~\cite{bornea13}

### 2014-09-22 docs/edbt2015/text/paper.tex
renamed sections
preprint
more on system section

@@
 \input{config}
+
+\newcommand{\rjs}[1]{\textbf{\textcolor{red}{#1 -- rjs}}}
 
@@
 % A category with the (minimum) three required fields
-\category{H.4}{Information Systems Applications}{Miscellaneous}
+\category{H.2.2}{Database Management}{Physical Design - Access methods}
 %A category including the fourth, optional field follows...
-\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
+\category{H.2.4}{Database Management}{Systems - Query Processing}
 
-\terms{Theory}
+\terms{Algorithms, Performance}
 
-\keywords{ACM proceedings, \LaTeX, text tagging}
+\keywords{interaction graphs; disk layout; adaptive storage}
 
 \input{introduction}
-
-\begin{alltt}\scriptsize
-- Intro
-    - What is an interaction graphs
-    - Relationship graphs vs interaction graphs 
-        - interactions get big
-\end{alltt}
-
 \input{system.tex}
-\input{problem.tex}
 \input{optimal.tex}
+\input{ilp.tex}
 \input{heuristic.tex}

### 2014-09-22 docs/edbt2015/text/system.tex
more on system section

@@
-\section{System Overview}\label{sec:system}
-
-\begin{figure}[t]
+\begin{figure}[!Ht]
 \includegraphics[width=0.9\columnwidth]{figures/example_interaction.pdf} 
@@
 
+\section{System Overview}\label{sec:system}
+
 The design of the railway disk layout builds on our prior work~\cite{gedik14}, 
@@
 access locality. The railway layout extends the previous design, by enabling the
-system to adapt the layout for changing workloads.
+system to adapt the layout to reduce query I/O for changing workloads.
 
@@
 To explain the design of the railway layout, we first introduce a small,
-motivating example, in Figure~\ref{fig:example}. This figure illustrates an
-interaction graph for call data records, capturing the telephone call
-interactions among four people. Each edge in the graph is associated with
+motivating example. Figure~\ref{fig:example} shows a graph for the telephone
+call interactions among four people. This is a subset of the information that is
+captured for call data records. Each edge in the graph is associated with
 attributes for the interaction, including the time the call was placed, the
@@
 device used to place the call. In this example, there were four calls
-placed. One of them is a call from Alice to Bob, starting at 13:46. They spoke
+placed. One of them was a call from Alice to Bob, starting at 13:46. They spoke
 for 600 seconds. The call was received by cell phone tower 1, and the Alice's
@@
 A telecommunications company typically performs various analytics on this
-data. For example, for billing purposes, they might want to capture the
-duration of all calls. To plan for infrastructure provisioning, they might want
-to record a count of the number of calls that each cell phone tower handled.
+data. For example, in order to understand how they should price their service
+plans, they might want to capture the duration of all calls for each user. To
+plan for infrastructure provisioning, they might want to record a count of the
+number of calls that each cell phone tower handled.
 
@@
 %
-The lefthand side of Figure~\ref{fig:before_after} shows the typical storage for interaction graphs.
+There are several ways in which one might store an interaction graph on
+disk. The graph structure can be stored as a matrix representation or an
+adjacency list. Most graph databases choose an adjacency list representation
+because they reduce the storage overhead when graphs are sparse, and it is
+faster to iterate over the edges when traversing the graph.  Attributes
+associates with each edge could be stored in separately in a relational table,
+or along with the edges. Storing the attributes with the edges improves the
+locality, since the database can read the graph structure and associated
+attributes from the same disk block. \rjs{Talk about timestamps.}
+
+The lefthand side of Figure~\ref{fig:before_after} illustrates a disk layout
+design with an adjacency list representation in which attributes are stored with
+the edges. This is the design used in our prior work~\cite{gedik14}.
+
+ typical storage for interaction graphs.
 Each disk block contains an adjacency list representation of the interaction
